### 2021-11-18 manuscript/main.tex
added skeleton

\PassOptionsToPackage{utf8}{inputenc}
\documentclass{bioinfo}

\usepackage[draft]{hyperref}
\usepackage{makecell}
\usepackage{comment}

\usepackage{floatrow}

% singlelinecheck=false puts subcaptions on the left
\usepackage[singlelinecheck=false]{subcaption}

\usepackage{algorithm2e}
\usepackage[usenames,dvipsnames]{xcolor}

% we squeeze our figures even more together
\captionsetup{belowskip=-2pt}

\SetAlgoLined
\SetKwProg{MyStruct}{Struct}{ contains}{end}

\newcommand{\vocab}{\textbf}
\newcommand{\red}[1]{{\textcolor{Red}{#1}}}
\newcommand{\FIXME}[1]{\red{[FIXME: #1]}}

\def\labelitemi{--}

\copyrightyear{2021} \pubyear{XXXX}

\access{Advance Access Publication Date: Day Month Year}
\appnotes{Genome Analysis}

\begin{document}
    \firstpage{1}

    \subtitle{Genome Analysis}

    \title[ODGI: understanding pangenome graphs]{seqwish: very large variation graph inducer}
    \author[Garrison \textit{et~al}.]{
        Erik~Garrison\,$^{\text{\sfb 1} \dagger*}$
    }

    \address{
        $^{\text{\sf 1}}$University of Tennessee Health Science Center, Memphis, TN, USA
    }

    \corresp{$^\ast$To whom correspondence should be addressed. \\
        $^\dagger$Contributed equally.}

    \history{Received on XXXXX; revised on XXXXX; accepted on XXXXX}

    \editor{Associate Editor: XXXXXXX}

    \abstract{
        \textbf{Motivation:}
        ... \\
        \textbf{Results:}
        We wrote seqwish, a novel ... \\
        \textbf{Availability:}
        seqwish is published as free software under the MIT open source license.
        Source code and documentation are available at \url{https://github.com/ekg/seqwish}.
        seqwish can be installed via Bioconda \url{https://bioconda.github.io/recipes/seqwish/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/seqwish.scm}. \\
        \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
    }

    \maketitle

    \section{Introduction}
    \label{sec:introduction}
    \FIXME{to do}

    \section{Model}
    \label{sec:model}
    \FIXME{to do}

    \section{Implementation}
    \label{sec:implementation}
    \FIXME{to do}

    \section{Results}
    \label{sec:results}
    \FIXME{to do}

    \subsection{subsection 1}
    \label{subsec:subsec1}
    \FIXME{to do}

    \section{Discussion}
    \label{sec:discussion}
    \FIXME{to do}

    \section*{Acknowledgements}

    We thank members of the HPRC production teams for their development of resources used in our exposition.

    \section*{Funding}

    We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG).

    \section*{Data availability}

    Code and links to data resources used to build this manuscript and its figures, can be found in the paper's public repository: \url{https://github.com/pangenome/seqwish-paper}.

    \bibliographystyle{natbib}

    \bibliography{document}

\end{document}

### 2021-11-18 manuscript/main.tex
copy-pasted initial details

@@
 
-    \title[ODGI: understanding pangenome graphs]{seqwish: very large variation graph inducer}
+    \title[seqwish: very large variation graph inducer]{seqwish: very large variation graph inducer}
     \author[Garrison \textit{et~al}.]{
@@
     \label{sec:model}
-    \FIXME{to do}
+
+    A set of pairwise alignments imply a variation graph, however I know of no contained method that will generate the variation graph or lossless string graph from these alignments.
+    To explore this, I developed an algorithm to do so that operates in external memory, which I here present in detail.
+    It operates by conversion of the alignment set into an alignment graph and the subsequent use of this graph in the elaboration of the variation graph including paths representing the input sequences.
+    The resulting graph is a lossless representation of the input and the alignments between them.
+    To distinguish the approach from string graphs, which imply error correction, I call this variation graph induction model the \emph{squish graph}.
+
+    \texttt{seqwish} implements a lossless conversion from pairwise alignments between sequences to a variation graph encoding the sequences and their alignments.
+    As input, we typically take all-versus-all alignments, but the exact structure of the alignment set may be defined in an application specific way.
+    \texttt{seqwish} uses a series of disk-backed sorts and passes over the alignment and sequence inputs to allow the graph to be constructed in low memory relative to the size of the input sequence set.
+    Memory usage during construction and traversal is limited by the use of sorted disk-backed arrays and succinct rank/select dictionaries to record a queryable version of the graph.
 
@@
     \label{sec:implementation}
-    \FIXME{to do}
+    As input, we have $Q$, which is a concatenation of the sequences from which we will build the graph.
+    We build a compressed suffix array (CSA) mapping sequence names to offsets in $Q$, and also the inverse using a rank/select dictionary on a bitvector marking the starts of sequences in $Q$.
+    This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.
+    We encode the set of input pairwise alignments between sequences in $Q$ as object $A$.
+    Although these alignments tend to be represented using oriented interval pairs in $Q$, for simplicity and robustness to graph complexity, we describe $A$ as a set of pairs of bidirectional positions (sequence offsets and strands) $[1 \ldots |Q_1 \ldots Q_{|Q|}|]$ , such that $A = \{ (b_{q}, b_{r}), \ldots \}$.
+    We sort $A$ by the first member ($b_{q}$) of each pair, ensuring that the entries in $A$ are ordered according to their order in $Q$.
+
+    To query the induced graph we build a rank/select dictionary allowing efficient traversal of $A$, based on a bit vector $A_{bv}$ of the same length as $A$ such that we record a 1 at those positions which correspond to the first instance of a given $b_{q}$ and record a 0 in $A_{bv}$ otherwise.
+    We record which $b_{q}$ we have processed in the bitvector $Q_{seen}$ which is of the same length as $Q$.
+    This allows us to avoid a quadratic penalty in the order of the size of the transitive closures in $Q$ generated by pairs in $A$.
+
+    Now we inductively derive the graph implied by the alignments.
+    For each base $b_{q}$ in $Q$ not already marked in $Q_{seen}$, we find its transitive closure $c_{q} := \{b_{q}, b_{r_{1}}, \ldots \}$ by traversing aligned base pairs recorded in $A$.
+    We write the character of the base $b_{q}$ to an entry $s_i$ in a vector $S$, then for each $b_{c}$ in $c_{q}$ we record a pair $(s_{i}, b_{c})$ into $N$ and its reverse, $(b_{c}, s_{i})$ into $P$.
+    We mark $Q_{seen}$ for each base in each emitted cluster, so that we will not consider these bases in subsequent transitive closures.
+    By sorting $N$ and $P$ by their first entries, we can build rank/select dictionaries on them akin to that we built on $A$ that allow random access by graph base (as given in $S$) or input base (as given in $Q$).
+
+    To fully induce the variation graph we need to establish the links between bases in $S$ that would be required for us to find any sequence in the input as a walk through the graph.
+    We do so by rewriting $Q$ (in both the forward and reverse orientation) in terms of pairs of bases in $S$, then sorting the resulting pairs by their first element, which yields $L = [(b_{a}, b_{b}), \ldots ]$.
+    These pairs record the links and their frequencies, which we can emit or filter (such as by frequency) as needed in particular applications.
+    In typical use we take the graph to be given by the unique elements of $L$.
+
+    Our data model encodes the graph using single-base nodes, but often downstream use requires identifying nodes and thus we benefit from compressing the unitigs of the graph into single nodes, which reduces memory used by identifiers in analysis.
+    We can compress the node space of the graph by traversing $S$, and for each base querying the inbound links.
+    Maintaining a bitvector $S_{id}$ of length equal to $S$ we mark each base at which we see any link other than one from or to the previous base on the forward or reverse strand, or at bases where we have no incoming links.
+    By building a rank/select dictionary on $S_{id}$ we can assign a smaller set of node ids to the sequence space of the graph.
+
+    Given the id space encoded by $S_{id}$ we can materialize the graph in a variety of interchange formats, or provide id-based interfaces to the indexed squish graph.
+    To generate graphs in \texttt {vg} or GFA format, we want to decompose the graph into its nodes ($S$), edges ($L$) and paths ($P$).
+    The nodes are given by $S$ and $S_{id}$, and similarly we project $L$ and $P$ through $S_{id}$ to obtain a compressed variation graph.
 

### 2021-11-18 scripts/scalability_test.sh
added `scalability_test.sh` script

@@ -1,0 +1,24 @@
+# All-vs-all using the whole HPRC year 1 dataset...
+
+# ...from scratch
+sbatch -p lowmem -c 48 --wrap 'cd /scratch && /home/guarracino/wfmash/build/bin/wfmash -X -s 100000 -l 300000 -p 98 -n 90 -k 16 -t 48 /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz | pigz -c > HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz && mv HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz /lizardfs/guarracino/HPRC/'
+
+# ...from input approximate mappings
+sbatch -p lowmem -c 48 --wrap 'cd /scratch && /home/guarracino/wfmash/build/bin/wfmash -X -s 100000 -l 300000 -p 98 -n 90 -k 16 -t 48 /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz -i /lizardfs/guarracino/HPRC/HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.approx.paf | pigz -c > HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz && mv HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz /lizardfs/guarracino/HPRC/'
+
+# ...from input approximate mappings in chunks
+python3 /home/guarracino/wfmash/scripts/split_approx_mappings_in_chunks.py /lizardfs/guarracino/HPRC/HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.approx.paf 5
[…]

### 2021-11-19 scripts/scalability_test.sh
added log; made the script more explicit

@@ -12,13 +12,13 @@ seq 0 4 | while read i; do sbatch -p lowmem -c 48 --wrap 'cd /scratch && /home/g
 
 
 # Graph induction (seqwish - beb7a3805c5af599b8425db94b9524eecbfc73e0)
-mkdir -p /lizardfs/guarracino/HPRCy1/graphs
+mkdir -p /lizardfs/guarracino/HPRC/graphs
 
 # ...single run from a single input PAF format file
-sbatch -p lowmem -c 48 --wrap 'cd /scratch && \time -v seqwish -t 48 -s /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz -p /lizardfs/guarracino/HPRC/HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz -k 311 -g HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.seqwish.k311.gfa -B 50M -P && pigz HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.seqwish.k311.gfa && mv HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.seqwish.k311.gfa.gz /lizardfs/guarracino/HPRCy1/graphs/'
+sbatch -p lowmem -c 48 --wrap 'cd /scratch && \time -v seqwish -t 48 -s /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz -p /lizardfs/guarracino/HPRC/HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz -k 311 -g HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.seqwish.k311.B50M.gfa -B 50M -P && mv HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.seqwish.k311.B50M.gfa /lizardfs/guarracino/HPRC/graphs/'
 
 # ...single run from multiple input PAF format chunks
[…]

### 2021-11-19 scripts/scalability_test.sh
rework directories; added markdown

@@ -1,24 +1,0 @@
-# All-vs-all using the whole HPRC year 1 dataset...
-
-# ...from scratch
-sbatch -p lowmem -c 48 --wrap 'cd /scratch && /home/guarracino/wfmash/build/bin/wfmash -X -s 100000 -l 300000 -p 98 -n 90 -k 16 -t 48 /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz | pigz -c > HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz && mv HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz /lizardfs/guarracino/HPRC/'
-
-# ...from input approximate mappings
-sbatch -p lowmem -c 48 --wrap 'cd /scratch && /home/guarracino/wfmash/build/bin/wfmash -X -s 100000 -l 300000 -p 98 -n 90 -k 16 -t 48 /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz /lizardfs/erikg/HPRC/year1v2genbank/parts/HPRCy1.pan.fa.gz -i /lizardfs/guarracino/HPRC/HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.approx.paf | pigz -c > HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz && mv HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.paf.gz /lizardfs/guarracino/HPRC/'
-
-# ...from input approximate mappings in chunks
-python3 /home/guarracino/wfmash/scripts/split_approx_mappings_in_chunks.py /lizardfs/guarracino/HPRC/HPRCy1.pan.fa.s100k.l300k.p98.n90.k16.approx.paf 5
[…]

### 2021-11-19 manuscript/fig_from_alignment_to_variation_graph.tex
other details; image placeholder

\vspace{-1em}
\begin{figure}[ht!]
%  \centering
    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={left,top},capbesidewidth=0.45\linewidth}}]{figure}[\FBwidth]
    {\caption{\FIXME{put high-resolution image} Building variation graphs from alignment graphs. Top: an alignment graph between a set of sequences,
        showing alignment edges in grey and precedence (path) edges in colors associated with each path. Bottom: the variation graph
        resulting from the application of the transitive closure process in \textit{seqwish}, where aligned segments have been
        compressed into a single node.
    }\label{fig:from_alignment_to_variation_graph}}
    %{\includegraphics[width=\linewidth]{fig/OUP_First_SBk_Bot_8401}} %\includegraphics[width=5cm]{name}}
    %\caption{Performance evaluation of \textit{odgi build} when translating a 90-haplotype graph of human chromosome 6 into ODGI's native format.}

\end{figure}
\vspace{-1em}

### 2021-11-19 manuscript/main.tex
other details; image placeholder

@@
         \textbf{Results:}
-        We wrote seqwish, a novel ... \\
+        We wrote seqwish, a novel tool to build a pangenome graph from a set of sequences and alignments between them. \\
         \textbf{Availability:}
@@
     \label{sec:introduction}
-    \FIXME{to do}
+    A pangenome models the full genomic information of a species or clase~\FIXME{PMID:32034321}.
+    These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular genome, chosen as reference.
+    A class of methods to represent pangenomes involves the \texttt{sequence graphs} \FIXME{PMID:2488477}, where similar orthologous regions between genomes collapse into a single representative sequence.
+    In \texttt{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+    A \texttt{bidirected} sequence graph represents both strands of DNA.
+    On this model, variation graphs add the concept of \texttt{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph\citep{Garrison:2018}.
+    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
 
@@
     \label{sec:model}
+    A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.
+    In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
+    Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
+    Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
+    Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$.
 
-    A set of pairwise alignments imply a variation graph, however I know of no contained method that will generate the variation graph or lossless string graph from these alignments.
-    To explore this, I developed an algorithm to do so that operates in external memory, which I here present in detail.
-    It operates by conversion of the alignment set into an alignment graph and the subsequent use of this graph in the elaboration of the variation graph including paths representing the input sequences.
+    A set of pairwise alignments imply a variation graph, however, to our knowledge, there is no contained method that can
+    generate the variation graph or lossless string graph from large input alignments.
+    To explore this, we developed an algorithm to do so that operates in external memory by conversion of the alignment
+    set into an alignment graph and the subsequent use of this graph in the elaboration of the variation graph including paths representing the input sequences.
     The resulting graph is a lossless representation of the input and the alignments between them.
-    To distinguish the approach from string graphs, which imply error correction, I call this variation graph induction model the \emph{squish graph}.
+    To distinguish the approach from string graphs, which imply error correction, we call this variation graph induction model the \emph{squish graph}.
+
+    \FIXME{To integrate to the text} The alignment graph formalizes the collection of alignments between genomes.
+    In this graph, each base pair in the input genomes is represented by a node.
+    Two kinds of edges connect these nodes.
+    The first `precedence' edge type indicates successive pairs of base pairs in the input,
+    linking together the input genomes within the alignment graph.
+    The second “match” type of edge links nodes that alignment has asserted to be matches.
+    The alignment graph may be converted into a basic sequence graph by condensing all sets of nodes that are transitively linked by match edges.
+    By recording the relationship between sequence graph and alignment graph nodes, we are able to trace the paths of pangenome sequences through the
+    graph, producing our target: a pangenome variation graph that completely describes both the input assemblies and their alignments (Fig.~\ref{fig:from_alignment_to_variation_graph}).
 
@@
     Memory usage during construction and traversal is limited by the use of sorted disk-backed arrays and succinct rank/select dictionaries to record a queryable version of the graph.
+
+    \input{fig_from_alignment_to_variation_graph.tex}
 
@@
 
+    \FIXME{To integrate to the text}
+    To support this kind of data scale, we implement \texttt{seqwish}'s algorithm using disk-backed memory.
+    We write the alignment graph into a disk-backed implicit interval tree, wherein each match in the input alignments is a queryable range.
+    A lock-free, parallel union-find algorithm \FIXME{Wait-free parallel algorithms for the union-find problem} allows us to, for each base pair of input sequence, collect all other base pairs in the pangenome that are aligned to it.
+    By processing this transitive closure operation in chunks (\textit{e.g.} 10Mbp at a time), we limit the maximum working memory (RAM) requirements of the process without affecting the output.
+    After the transitive closure, we write our final graph on disk in a set of arrays, then augment it with the paths of our input sequences,
+    finally emitting the resulting graph in the standard Graphical Fragment Assembly (GFA) format~\citep{GFA}.
+
     \section{Results}

### 2021-11-20 manuscript/main.tex
notes

@@
     \label{sec:implementation}
+
+    \FIXME{The algorithm validates some aspects of itself, we should point that out}
+
     As input, we have $Q$, which is a concatenation of the sequences from which we will build the graph.
@@
 
+    %The alignment is taken as truth, so some of the cross validation things we do for pggb are irrelevant.
+    %The one experiment could be graph induction on many graphs, with a 2D/3D plot with input size (alignments) vs induction time and color/shape to show graph output size or something
+    %Or
+    %We go schematic with the figure to show a complex but very detailed example with the algorithm worked out
+
     \subsection{subsection 1}

### 2021-12-04 manuscript/main.tex
add `-k0` in the workflow; draft the abstract

@@
 
-    \title[seqwish: very large variation graph inducer]{seqwish: very large variation graph inducer}
+    \title[seqwish: inducing pangenome graphs]{seqwish: inducing pangenome graphs}
     \author[Garrison \textit{et~al}.]{
-        Erik~Garrison\,$^{\text{\sfb 1} \dagger*}$
+        Erik~Garrison\,$^{\text{\sfb 1}*}$
     }
@@
 
-    \corresp{$^\ast$To whom correspondence should be addressed. \\
-        $^\dagger$Contributed equally.}
+    \corresp{
+        $^\ast$To whom correspondence should be addressed. \\
+       % $^\dagger$Contributed equally.\
+    }
 
@@
         \textbf{Motivation:}
-        ...  \\
+        Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
+        One of its most common forms is the variation graph model.
+        A set of pairwise alignments imply a variation graph, but, to our knowledge, there are no methods to generate such a graph or lossless string graph from these alignments.\\
         \textbf{Results:}
-        We wrote seqwish, a novel tool to build a pangenome graph from a set of sequences and alignments between them. \\
+        We wrote seqwish, a novel tool to induce a variation graph from a set of sequences and alignments between them.
+        seqwish operates in external memory, by conversion of the alignment set into an alignment graph and the subsequent
+        use of this graph in the elaboration of the variation graph including paths representing the input sequences.
+        The resulting graph is a lossless representation of the input and its alignments.\\
         \textbf{Availability:}

### 2021-12-21 manuscript/main.tex
andrea++

@@
     \author[Garrison \textit{et~al}.]{
-        Erik~Garrison\,$^{\text{\sfb 1}*}$
+        Erik~Garrison\,$^{\text{\sfb 1}*}$,
+        Andrea~Guarracino\,$^{\text{\sfb 2}}$
     }
@@
     \address{
-        $^{\text{\sf 1}}$University of Tennessee Health Science Center, Memphis, TN, USA
+        $^{\text{\sf 1}}$University of Tennessee Health Science Center, Memphis, TN, USA \\
+        $^{\text{\sf 2}}$Genomics Research Centre, Human Technopole, Milan, Italy
     }

### 2021-12-21 manuscript/main.tex
motivate

@@
 
-    \title[seqwish: inducing pangenome graphs]{seqwish: inducing pangenome graphs}
+\title[Multi-reference pangenome graphs]{Multi-reference pangenome graphs}
 \author[Garrison \textit{et~al}.]{
@@
 Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
-        One of its most common forms is the variation graph model.
-        A set of pairwise alignments imply a variation graph, but, to our knowledge, there are no methods to generate such a graph or lossless string graph from these alignments.\\
+They are commonly modeled as variation graphs.
+A set of pairwise alignments imply a variation graph, but, to our knowledge, there are no methods to generate such a graph or lossless string graph from these alignments. \\
 \textbf{Results:}
-        We wrote seqwish, a novel tool to induce a variation graph from a set of sequences and alignments between them.
-        seqwish operates in external memory, by conversion of the alignment set into an alignment graph and the subsequent
-        use of this graph in the elaboration of the variation graph including paths representing the input sequences.
+We design the \textit{seqwish} algorithm, which builds variation graph from a set of sequences and alignments between them.
+Our seqwish implementation operates in external memory, using disk-backed implicit interval trees and a lock-free parallel union find algorithm to transform the alignment set into an alignment graph, then a sequence graph, and finally a variation graph with paths representing the input sequences.
+%        use of this graph in the elaboration of the variation graph including paths representing the input sequences.
 The resulting graph is a lossless representation of the input and its alignments.\\
@@
 \label{sec:introduction}
-    A pangenome models the full genomic information of a species or clase~\FIXME{PMID:32034321}.
-    These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular genome, chosen as reference.
-    A class of methods to represent pangenomes involves the \texttt{sequence graphs} \FIXME{PMID:2488477}, where similar orthologous regions between genomes collapse into a single representative sequence.
-    In \texttt{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
-    A \texttt{bidirected} sequence graph represents both strands of DNA.
-    On this model, variation graphs add the concept of \texttt{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph\citep{Garrison:2018}.
-    Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+A pangenome models the full genomic information of a species or clade~\FIXME{PMID:32034321}.
+These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular ``reference'' genome.
+%In principle, pangenome graphs can allow for a great reduction in reference bias, allowing any part of a pangenome to be considered in biological studies.
+%, and may provide a simpler representation of structurally-variable regions.
+In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to sequences that are very similar to a chosen reference genome.
+
+Surprisingly, existing methods to build pangenome graphs from large genome assemblies are reference-based.
+Minigraph progressively builds a pangenome graph using by layering an ordered series of input genomes onto a specific reference.
+Cactus depends on a single guide tree to structure pairwise alignments between genomes.
+Both kinds of methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results exhibit order and reference-dependency.
+
+Although a reference-guided approach is pragmatic when we only have a single reference genome, in humans and other species recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies.
+Motivated by the need to generate ``multi-reference'' pangenome graphs, we present \textit{seqwish}, an algorithm that allows us to build pangenome graphs directly from a collection of sequences and pairwise alignments between them.
+Here, we formally describe the algorithm, evaluate its basic combinatoric bounds and costs, and present applications of the method to pangenome building problems in a variety of species.
+Our approach is both highly generic---in that we can apply it to truly any collection of sequences and alignments---but it easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
+%Not only is our approach highly generic,but it also efficiently scales to both large and complex pangenomes.
+%with \textit{any} collection of alignments and sequences,
+
+%Given recent advances in genome sequencing and assembly, we expect that soon, many species will have multiple reference-quality genomes.
+%To build a pangenome model from such genomes that is truly unbiased, we need an approach to build a multi-reference 
+%Motivated by the current availability of two reference-quality human genome assemblies 
+%The inability to build pangenome graphs from
 
@@
     \label{sec:model}
+
+A class of methods to represent pangenomes involves the \textit{sequence graphs} \FIXME{PMID:2488477}, where similar orthologous regions between genomes collapse into a single representative component in the graph.
+In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+A \textit{bidirected} sequence graph represents both strands of DNA.
+On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph\citep{Garrison:2018}.
+Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+
     A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.

### 2021-12-21 manuscript/main.tex
reorder

@@
 
-    A pangenome graph is a sequence model that encodes the mutual alignment of many genomes~\citep{Garrison_2019_thesis,Eizenga_2020}.
+A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Garrison_2019_thesis,Eizenga_2020}.
 In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
@@
 
-    A set of pairwise alignments imply a variation graph, however, to our knowledge, there is no contained method that can
-    generate the variation graph or lossless string graph from large input alignments.
-    To explore this, we developed an algorithm to do so that operates in external memory by conversion of the alignment
-    set into an alignment graph and the subsequent use of this graph in the elaboration of the variation graph including paths representing the input sequences.
-    The resulting graph is a lossless representation of the input and the alignments between them.
-    To distinguish the approach from string graphs, which imply error correction, we call this variation graph induction model the \emph{squish graph}.
+A variation graph represents all pairwise alignments between its paths.
+It follows that we can construct a variation graph from a set of pairwise alignments.
+We do so in a process based on three distinct graph models.
+(\textit{1}) We transform the alignments and input sequences into an \textit{alignment graph} \cite{Kehr_2014}.
+(\textit{2}) We then contract match components in this graph into single nodes of a \textit{sequence graph}.
+(\textit{3}) By recording the translation from alignment graph nodes to sequence graph nodes, we are able to walk the input sequences through the graph.
+In doing so, we are able to derive edges and label the paths of a \textit{variation graph} which captures the pairwise alignments and all implied transitive relationships between them.
+
+%Finally, we derive edges and label the paths of the output variation graph by
+
+%A set of pairwise alignments imply a variation graph.
+
+%We can see that pairs of aligned sequences would be represented as single
+
+%, however, to our knowledge, there is no existing method that can generate the variation graph or lossless string graph from large input alignments with arbitrary properties.
+
+%We first formalize the graph induction algorithm that we will implement
+
+%To explore this,  algorithm to do so that operates in external memory by conversion of the alignment set into an alignment graph and the subsequent use of this graph in the elaboration of the variation graph including paths representing the input sequences.
+%The resulting graph is a lossless representation of the input and the alignments between them.
+%To distinguish the approach from string graphs, which imply error correction, we call the result of this variation graph induction model the \emph{squish graph}.
+
+\subsection{Sequence index}
+
+As input, we have $Q$, which is a concatenation of the sequences from which we will build the graph.
+We build a compressed suffix array (CSA) mapping sequence names to offsets in $Q$, and also the inverse using a rank/select dictionary on a bitvector marking the starts of sequences in $Q$.
+This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.
+
+\subsection{Alignment graph}
 
@@
 The second “match” type of edge links nodes that alignment has asserted to be matches.
-    The alignment graph may be converted into a basic sequence graph by condensing all sets of nodes that are transitively linked by match edges.
-    By recording the relationship between sequence graph and alignment graph nodes, we are able to trace the paths of pangenome sequences through the
-    graph, producing our target: a pangenome variation graph that completely describes both the input assemblies and their alignments (Fig.~\ref{fig:from_alignment_to_variation_graph}).
 
-    \texttt{seqwish} implements a lossless conversion from pairwise alignments between sequences to a variation graph encoding the sequences and their alignments.
-    As input, we typically take all-versus-all alignments, but the exact structure of the alignment set may be defined in an application specific way.
-    \texttt{seqwish} uses a series of disk-backed sorts and passes over the alignment and sequence inputs to allow the graph to be constructed in low memory relative to the size of the input sequence set.
-    Memory usage during construction and traversal is limited by the use of sorted disk-backed arrays and succinct rank/select dictionaries to record a queryable version of the graph.
-
-    \input{fig_from_alignment_to_variation_graph.tex}
-
-    \section{Implementation}
-    \label{sec:implementation}
-
-    \FIXME{The algorithm validates some aspects of itself, we should point that out}
-
-    As input, we have $Q$, which is a concatenation of the sequences from which we will build the graph.
-    We build a compressed suffix array (CSA) mapping sequence names to offsets in $Q$, and also the inverse using a rank/select dictionary on a bitvector marking the starts of sequences in $Q$.
-    This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.
 We encode the set of input pairwise alignments between sequences in $Q$ as object $A$.
@@
 By sorting $N$ and $P$ by their first entries, we can build rank/select dictionaries on them akin to that we built on $A$ that allow random access by graph base (as given in $S$) or input base (as given in $Q$).
+
+\subsection{Sequence graph}
+
+The alignment graph may be converted into a basic sequence graph by condensing all sets of nodes that are transitively linked by match edges.
+
+\subsection{Variation graph}
+
+By recording the relationship between sequence graph and alignment graph nodes, we are able to trace the paths of pangenome sequences through the graph, producing our target: a pangenome variation graph that completely describes both the input assemblies and their alignments (Fig.~\ref{fig:from_alignment_to_variation_graph}).
+
+Seqwish implements a lossless conversion from pairwise alignments between sequences to a variation graph encoding the sequences and their alignments.
+As input, we typically take all-versus-all alignments, but the exact structure of the alignment set may be defined in an application specific way.
+Seqwish uses a series of disk-backed sorts and passes over the alignment and sequence inputs to allow the graph to be constructed in low memory relative to the size of the input sequence set.
+Memory usage during construction and traversal is limited by the use of sorted disk-backed arrays and succinct rank/select dictionaries to record a queryable version of the graph.
+
+    \input{fig_from_alignment_to_variation_graph.tex}
+
+    \section{Implementation}
+    \label{sec:implementation}
+
+    \FIXME{The algorithm validates some aspects of itself, we should point that out}
 

### 2022-01-04 manuscript/main.tex
abstractify

@@
 \textbf{Motivation:}
-Pangenome graphs provide a complete representation of the mutual alignment of collections of genomes.
-They are commonly modeled as variation graphs.
-A set of pairwise alignments imply a variation graph, but, to our knowledge, there are no methods to generate such a graph or lossless string graph from these alignments. \\
+Pangenome graphs, often formally encoded as variation graphs, moodel the mutual alignment of collections of genomes.
+A set of pairwise alignments imply a variation graph, but there are no scalable methods to generate such a graph or lossless string graph from these alignments.
+Existing approaches depend on a single reference or a specific ordering of genomes, and do not provide base-level information about the mapping of genomes into the pangenome.
+A scalable method to build pangenome graphs without such biases would be a key functional step in pangenome construction and manipulation pipelines. \\
 \textbf{Results:}
 We design the \textit{seqwish} algorithm, which builds variation graph from a set of sequences and alignments between them.
-Our seqwish implementation operates in external memory, using disk-backed implicit interval trees and a lock-free parallel union find algorithm to transform the alignment set into an alignment graph, then a sequence graph, and finally a variation graph with paths representing the input sequences.
-%        use of this graph in the elaboration of the variation graph including paths representing the input sequences.
-The resulting graph is a lossless representation of the input and its alignments.\\
+We first transform the alignment set into an interval tree.
+To build up the variation graph, we query this tree-based representation of the alignments to reduce transitive matches into single DNA segments in a sequence graph.
+By recording the mapping from input sequence to output graph, we can trace the original paths through this graph, yielding a pangenome graph.
+We present an implementation that operates in external memory, using disk-backed implicit interval trees and a lock-free parallel algorithm to drive the core graph induction step.
+We demonstrate that our method scales to very large graph induction problems, and apply our work to build graphs from dozens of genomes from the same or different species in clades throughout the tree of life.
+\\
 \textbf{Availability:}
@@
 Cactus depends on a single guide tree to structure pairwise alignments between genomes.
-Both kinds of methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results exhibit order and reference-dependency.
+Both kinds of methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, and guide-tree topology.
 
@@
 
-    We thank members of the HPRC production teams for their development of resources used in our exposition.
+    We thank Richard Durbin, Benedict Paten, and Simon Puglisi for support and helpful discussions in the development of this algorithm.
+    %We are greatful to members of the HPRC production teams for their development of resources used in our exposition.
 

### 2022-01-04 manuscript/main.tex
working on model lead-in

@@
 \label{sec:introduction}
-A pangenome models the full genomic information of a species or clade~\FIXME{PMID:32034321}.
-These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular ``reference'' genome.
+A pangenome models the full genomic information of a species or clade \citep{Medini_2005,Sherman_2020}.
+In contrast to reference-based approaches that relate sequences to a particular reference genome, methods that use pangenome reference systems attempt to model the mutual relationship between all represented genomes \citep{cpang2018}.
+%These models encode the mutual relationships between all the genomes represented, in contrast to reference-based approaches which relate sequences only to a particular ``reference'' genome.
 %In principle, pangenome graphs can allow for a great reduction in reference bias, allowing any part of a pangenome to be considered in biological studies.
 %, and may provide a simpler representation of structurally-variable regions.
-In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to sequences that are very similar to a chosen reference genome.
+Popular approaches model the pangenome alignment as a \textit{pangenome graph}.
+Conceptually, the pangenome graph relates genomes as walks through and underlying DNA sequence language encoded as a sequence graph. 
+In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to sequences that are very similar to a chosen reference genome \cite{Eizenga_2020}.
 
 Surprisingly, existing methods to build pangenome graphs from large genome assemblies are reference-based.
-Minigraph progressively builds a pangenome graph using by layering an ordered series of input genomes onto a specific reference.
-Cactus depends on a single guide tree to structure pairwise alignments between genomes.
+Minigraph progressively builds a pangenome graph using by layering an ordered series of input genomes onto a specific reference \citep{Li:2020}.
+Cactus depends on a single guide tree to structure pairwise alignments between genomes \citep{Armstrong:2020}.
 Both kinds of methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, and guide-tree topology.
 
-Although a reference-guided approach is pragmatic when we only have a single reference genome, in humans and other species recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies.
-Motivated by the need to generate ``multi-reference'' pangenome graphs, we present \textit{seqwish}, an algorithm that allows us to build pangenome graphs directly from a collection of sequences and pairwise alignments between them.
+Although a reference-guided approach is pragmatic when we only have a single reference genome, in humans and other species recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies \cite{Nurk_2021}.
+Motivated by the need to generate ``multi-reference'' pangenome graphs, we develop \textit{seqwish}, an algorithm that allows us to build pangenome graphs directly from a collection of sequences and pairwise alignments between them.
 Here, we formally describe the algorithm, evaluate its basic combinatoric bounds and costs, and present applications of the method to pangenome building problems in a variety of species.
-Our approach is both highly generic---in that we can apply it to truly any collection of sequences and alignments---but it easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
+%Our approach is both highly generic and scalable.
+We can apply our method to any collection of sequences and alignments.
+It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
 %Not only is our approach highly generic,but it also efficiently scales to both large and complex pangenomes.
@@
 
-A class of methods to represent pangenomes involves the \textit{sequence graphs} \FIXME{PMID:2488477}, where similar orthologous regions between genomes collapse into a single representative component in the graph.
+A class of methods to represent pangenomes involves the \textit{sequence graph} \citep{Hein_1989}, where identical segments of different genomes collapse into a single representative segment (often node) in the graph.
 In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
 A \textit{bidirected} sequence graph represents both strands of DNA.
-On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph\citep{Garrison:2018}.
+On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{Garrison_2018}.
 Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
@@
 In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
-Each node $n_i$ has an identifier $i$ and an implicit reverse complement $\bar{n_i}$, and a node strand $s$ corresponds to one such orientations.
-Edges $E = e_1\ldots e_{|E|}$ represent ordered pairs of node strands: $e_i = ( s_a, s_b )$.
-Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$.
+Each node $n_i$ has a unique numerical identifier $i$ and an implicit reverse complement $\bar{n_i}$.
+A node strand $s$ corresponds to one node orientation.
+Edges $E = e_1\ldots e_{|E|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
+Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written terms of the sequence graph.
 
-A variation graph represents all pairwise alignments between its paths.
-It follows that we can construct a variation graph from a set of pairwise alignments.
-We do so in a process based on three distinct graph models.
-(\textit{1}) We transform the alignments and input sequences into an \textit{alignment graph} \cite{Kehr_2014}.
-(\textit{2}) We then contract match components in this graph into single nodes of a \textit{sequence graph}.
+A variation graph models a set of pairwise alignments between its embedded paths.
+The variation graph shows that two sequences match when they both ``step'' on the same node or series of nodes.
+This is trivial given our definition.
+If two sub-paths contain the same set of steps, they must have identical subsequences.
+%Thus, to represent alignments in the graph, we built it by rendering matches into the graph
+We can construct a variation graph from a set of pairwise alignments by guaranteeing that each matching pair of subsequences are represented in a single walk in the output graph.
+
+We do so in a process based on three complementary graph models.
+(\textit{1}) We transform the alignments and input sequences into an \textit{alignment graph} \citep{Kehr_2014}.
+(\textit{2}) We then contract match components in this graph into single nodes of a \textit{sequence graph} \citep{Hein_1989}.
 (\textit{3}) By recording the translation from alignment graph nodes to sequence graph nodes, we are able to walk the input sequences through the graph.
-In doing so, we are able to derive edges and label the paths of a \textit{variation graph} which captures the pairwise alignments and all implied transitive relationships between them.
-
-%Finally, we derive edges and label the paths of the output variation graph by
-
-%A set of pairwise alignments imply a variation graph.
-
-%We can see that pairs of aligned sequences would be represented as single
-
-%, however, to our knowledge, there is no existing method that can generate the variation graph or lossless string graph from large input alignments with arbitrary properties.
-
-%We first formalize the graph induction algorithm that we will implement
-
-%To explore this,  algorithm to do so that operates in external memory by conversion of the alignment set into an alignment graph and the subsequent use of this graph in the elaboration of the variation graph including paths representing the input sequences.
-%The resulting graph is a lossless representation of the input and the alignments between them.
-%To distinguish the approach from string graphs, which imply error correction, we call the result of this variation graph induction model the \emph{squish graph}.
+In doing so, we are able to derive edges and label the paths of a \textit{variation graph} \citep{Garrison_2018} which captures the pairwise alignments and all implied transitive relationships between them.
 

### 2022-01-05 manuscript/main.tex
onward

@@
 
-\section{Model}
+\section{Variation graph induction}
 \label{sec:model}
@@
 Edges $E = e_1\ldots e_{|E|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
-Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written terms of the sequence graph.
+Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written in terms of the sequence graph.
 
-A variation graph models a set of pairwise alignments between its embedded paths.
-The variation graph shows that two sequences match when they both ``step'' on the same node or series of nodes.
-This is trivial given our definition.
+A variation graph represents pairwise alignments between its embedded paths.
+We infer that two sequences match when they both ``step'' on the same node or series of nodes.
+This follows from our definition of the graph itself.
 If two sub-paths contain the same set of steps, they must have identical subsequences.
 %Thus, to represent alignments in the graph, we built it by rendering matches into the graph
+%The core of our graph constr TODO HERE
 We can construct a variation graph from a set of pairwise alignments by guaranteeing that each matching pair of subsequences are represented in a single walk in the output graph.
@@
 
-\subsection{Sequence index}
+\subsection{Sequence input}
 
-As input, we have $Q$, which is a concatenation of the sequences from which we will build the graph.
-We build a compressed suffix array (CSA) mapping sequence names to offsets in $Q$, and also the inverse using a rank/select dictionary on a bitvector marking the starts of sequences in $Q$.
+As input, we have $Q = g_1 \ldots g_N$, which is a concatenation of the $N$ genome sequences from which we will build the graph.
+Using a compressed suffix array (CSA) over sequence names, we build an index that maps between offsets in all genomes and offsets in $Q$ via lookup of the sequence name in the CSA.
 This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.
@@
 \subsection{Alignment graph}
+
+Alignments between genomes imply a set of exact matches between subsequences of $Q$, which we represent as interval pairs $m := ((q_b, q_e), (t_b, t_e))$.
+We encode the $X$ matches in a set of alignments made by an external method \cite{wfmash,Li_2018} in $A = m_1 \ldots m_X$.
+
+\citep{Li_bedtk_2020}.
 

### 2022-01-05 manuscript/main.tex
formally

@@
 \usepackage[usenames,dvipsnames]{xcolor}
+
+\usepackage{amsthm}
+\theoremstyle{definition}
+\newtheorem{definition}{Definition}[section]
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{corollary}{Corollary}[theorem]
+\newtheorem{lemma}[theorem]{Lemma}
 
@@
 \section{Variation graph induction}
-\label{sec:model}
+\label{sec:algorithm}
 
@@
 Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Eizenga_2020}.
 
-A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Garrison_2019_thesis,Eizenga_2020}.
+\begin{definition}
+\label{def:vg}
+Variation graphs are a common formalism to encode pangenome graphs \citep{Garrison_2019_thesis}.
 In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
@@
 Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written in terms of the sequence graph.
+\end{definition}
 
+\begin{theorem}
 A variation graph represents pairwise alignments between its embedded paths.
-We infer that two sequences match when they both ``step'' on the same node or series of nodes.
-This follows from our definition of the graph itself.
-If two sub-paths contain the same set of steps, they must have identical subsequences.
-%Thus, to represent alignments in the graph, we built it by rendering matches into the graph
-%The core of our graph constr TODO HERE
-We can construct a variation graph from a set of pairwise alignments by guaranteeing that each matching pair of subsequences are represented in a single walk in the output graph.
+\end{theorem}
 
-We do so in a process based on three complementary graph models.
+\begin{proof}
+By definition \ref{def:vg}, two paths have identical subsequences where they walk (or ``step'') through the same series of oriented nodes (e.g. $s_1 s_2 s_3$).
+An identical set of path steps is thus equivalent to a sequence match.
+Pairwise alignments are by definition collections of matches between sequences.
+The variation graph thus models a set of pairwise alignments between paths in $P$.
+\end{proof}
+
+\begin{theorem}
+We can construct a variation graph from pairwise alignments.
+\end{theorem}
+
+This follows from \ref{def:vg}.
+Our input $Q = S \vee \bar{S}$ is a set of $N$ DNA sequences $S = g_1 \ldots g_N$ and their reverse complements $\bar{S} = \bar{g_1} \ldots \bar{g_N}$.
+%Assume that $i$ and $j$ are coordinates in $Q$.
+A match $m = (i, j)$ asserts the evolutionary equivalence of two charaters in sequences in $Q$.
+%$\forall_{(i, j) : j = |g| - i} g[i] = \bar{g}[j]$.
+Pairwise alignments between sequences in $Q$ are a set of matches $A = \{ m_1 \ldots m_{|A|} \}$.
+By standard definition, each sequence matches its own reverse complement: $g[i] = \bar{g}[j]$ for all $j = |g| - i$, and we assume these matches are included in $A$.
+%A match $m$ can be understood as a component in a binary relation.
+The transitive closure of a match, $m^+ = \{ i \ldots j \}$, is a set of characters in $Q$ that are transitively linked together by other matches.
+By definition of $m$, each $m^+$ implies a single, identical character $c(m^+)$.
+
+We build a graph $V$ inductively.
+We take the first match in $A$, $m_1$, and execute a union-find operation to obtain $m_1^+$.
+We add the character of the match $c(m_1^+)$ as a node $n_1$ in $V$, record the mapping from $m_1^+ \to n_1$.
+To induce the graph, we take the next un-used match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $V$ while recording the match transitive closure to graph mapping.
+We continue until all matches have been used.
+Finally, we establish paths ($P$) by projecting $Q$ through the match-to-node mapping, and record edges ($E$) where nodes occur successively in paths.
+
+\begin{proof}
+After the first step of induction, the graph represents all pairwise matches in $m_1^+$.
+Each subsequent step includes progressively more of $A$, until at completion, all pairwise relationships are accounted for in $V$.
+The variation graph is constructed from pairwise alignments, and because its paths are equivalent to input sequences, it represents them losslessly.
+\end{proof}
+
+\begin{comment}
+
+Our algorithm flows through three complementary graph models.
 (\textit{1}) We transform the alignments and input sequences into an \textit{alignment graph} \citep{Kehr_2014}.
-(\textit{2}) We then contract match components in this graph into single nodes of a \textit{sequence graph} \citep{Hein_1989}.
+(\textit{2}) We then transitively close match relations in this graph to produce single nodes of a \textit{sequence graph} \citep{Hein_1989}.
 (\textit{3}) By recording the translation from alignment graph nodes to sequence graph nodes, we are able to walk the input sequences through the graph.
@@
 
-As input, we have $Q = g_1 \ldots g_N$, which is a concatenation of the $N$ genome sequences from which we will build the graph.
 Using a compressed suffix array (CSA) over sequence names, we build an index that maps between offsets in all genomes and offsets in $Q$ via lookup of the sequence name in the CSA.
@@
 Memory usage during construction and traversal is limited by the use of sorted disk-backed arrays and succinct rank/select dictionaries to record a queryable version of the graph.
+
+\end{comment}
 

### 2022-01-05 manuscript/fig_induction.tex
stuff

\documentclass{article}
\usepackage{tikz}
\usetikzlibrary{positioning}
\begin{document}
\begin{tikzpicture}[
roundnode/.style={circle, draw=green!60, fill=green!5, very thick, minimum size=7mm},
squarednode/.style={rectangle, draw=red!60, fill=red!5, very thick, minimum size=5mm},
]
%Nodes
\node[squarednode] (seqA1) {G};
\node[squarednode] (seqA2) [right=of seqA1] {A};
\node[squarednode] (seqA3) [right=of seqA2] {T};
\node[squarednode] (seqA4) [right=of seqA3] {T};
\node[squarednode] (seqA5) [right=of seqA4] {A};
\node[squarednode] (seqA6) [right=of seqA5] {C};
\node[squarednode] (seqA7) [right=of seqA6] {A};

\node[squarednode] (seqB1) [below=of seqA1] {Q};

%\node[roundnode]        (uppercircle)       [above=of seqA1] {1};
%\node[squarednode]      (rightsquare)       [below=of seqA1] {3};
%\node[roundnode]        (lowercircle)       [below=of seqA1] {4};

%Lines
\draw (seqA1) -- (seqA2);
\draw (seqA2) -- (seqA3);
\draw (seqA3) -- (seqA4);
\draw (seqA4) -- (seqA5);
\draw (seqA5) -- (seqA6);
\draw (seqA6) -- (seqA7);

%\draw[->] (uppercircle.south) -- (seq1.north);
%\draw[->] (seq1.east) -- (rightsquare.west);
%\draw[->] (rightsquare.south) .. controls +(down:7mm) and +(right:7mm) .. (lowercircle.east);
%\draw[->] (rightsquare.south) -- (lowercircle.east);
\end{tikzpicture}
\end{document}

### 2022-01-05 manuscript/fig_induction.tex
an attempt to illustrate seqwish

@@
 \usetikzlibrary{positioning}
+
+\pgfdeclarelayer{bg}    % declare background layer
+\pgfsetlayers{bg,main}  % set the order of the layers (main is the standard layer)
+
 \begin{document}
-\begin{tikzpicture}[
-roundnode/.style={circle, draw=green!60, fill=green!5, very thick, minimum size=7mm},
-squarednode/.style={rectangle, draw=red!60, fill=red!5, very thick, minimum size=5mm},
+\begin{tikzpicture}[auto,
+    node distance= 10mm and 20mm,
+    roundnode/.style={circle, draw=black!60, fill=gray!10, very thick, minimum size=7mm},
+    squarednode/.style={rectangle, draw=black!60, fill=gray!10, very thick, minimum size=5mm},
 ]
@@
 \node[squarednode] (seqA1) {G};
-\node[squarednode] (seqA2) [right=of seqA1] {A};
-\node[squarednode] (seqA3) [right=of seqA2] {T};
-\node[squarednode] (seqA4) [right=of seqA3] {T};
-\node[squarednode] (seqA5) [right=of seqA4] {A};
-\node[squarednode] (seqA6) [right=of seqA5] {C};
-\node[squarednode] (seqA7) [right=of seqA6] {A};
-
-\node[squarednode] (seqB1) [below=of seqA1] {Q};
+\node[squarednode] (seqA2) [right of=seqA1] {A};
+\node[squarednode] (seqA3) [right of=seqA2] {T};
+\node[squarednode] (seqA4) [right of=seqA3] {T};
+\node[squarednode] (seqA5) [right of=seqA4] {A};
+\node[squarednode] (seqA6) [right of=seqA5] {C};
+\node[squarednode] (seqA7) [right of=seqA6] {A};
 
@@
 %\node[roundnode]        (lowercircle)       [below=of seqA1] {4};
+
+\node[squarednode] (seqB1) [below of=seqA1] {G};
+\node[squarednode] (seqB2) [right of=seqB1] {C};
+\node[squarednode] (seqB3) [right of=seqB2] {T};
+\node[squarednode] (seqB4) [right of=seqB3] {T};
+\node[squarednode] (seqB5) [right of=seqB4] {A};
+\node[squarednode] (seqB6) [right of=seqB5] {C};
+\node[squarednode] (seqB7) [right of=seqB6] {G};
 
@@
 
+\draw (seqB1) -- (seqB2);
+\draw (seqB2) -- (seqB3);
+\draw (seqB3) -- (seqB4);
+\draw (seqB4) -- (seqB5);
+\draw (seqB5) -- (seqB6);
+\draw (seqB6) -- (seqB7);
+
+\node[squarednode] (seqC1) [below of=seqB1] {G};
+\node[squarednode] (seqC2) [right of=seqC1] {A};
+\node[squarednode] (seqC3) [right of=seqC2] {T};
+\node[squarednode] (seqC4) [right of=seqC3] {C};
+\node[squarednode] (seqC5) [right of=seqC4] {A};
+
+\draw (seqC1) -- (seqC2);
+\draw (seqC2) -- (seqC3);
+\draw (seqC3) -- (seqC4);
+\draw (seqC4) -- (seqC5);
+
+%\begin{scope}[node distance=0.5cm]
+  \node[roundnode] (g2) [below = 3cm of seqA1] {A};
+\node[roundnode] (g1) [below left of=g2] {G};
+\node[roundnode] (g3) [below right of=g2] {C};
+\node[roundnode] (g4) [above right of=g3] {T};
+\node[roundnode] (g5) [right of=g4] {T};
+\node[roundnode] (g6) [right of=g5] {A};
+\node[roundnode] (g7) [right of=g6] {C};
+\node[roundnode] (g8) [above right of=g7] {A};
+\node[roundnode] (g9) [below right of=g7] {G};
+%\end{scope}
+
+\draw (g1) to (g2);
+\draw (g1) to (g3);
+\draw (g2) to (g4);
+\draw (g3) to (g4);
+\draw (g4) to (g5);
+\draw (g4) to [looseness=1] (g7);
+\draw (g5) to (g6);
+\draw (g6) to (g7);
+\draw (g7) to (g8);
+\draw (g7) to (g9);
+
+\begin{pgfonlayer}{bg}    % select the background layer
+\draw[line width=1,color=red!100,looseness=1] (seqA1.south) to (g1);
+\draw[line width=1,color=red!100,looseness=1] (seqB1.south) to (g1);
+\draw[line width=1,color=red!100,looseness=1] (seqC1.south) to (g1);
+
+\draw[line width=1,color={rgb:red,1;orange,1},looseness=0.5] (seqA2.south) to (g2);
+\draw[line width=1,color={rgb:red,1;orange,1},looseness=0.5] (seqC2.south) to (g2);
+
+\draw[line width=1,color=orange!100,looseness=0.5] (seqA3.south) to (g4);
+\draw[line width=1,color=orange!100,looseness=0.5] (seqB3.south) to (g4);
+\draw[line width=1,color=orange!100,looseness=0.5] (seqC3.south) to (g4);
+
+\draw[line width=1,color=green!100,looseness=0.5] (seqA4.south) to (g5);
+\draw[line width=1,color=green!100,looseness=0.5] (seqB4.south) to (g5);
+%\draw[line width=1,color=blue!100,looseness=0.5] (seqC3.south) to (g4);
+
+\draw[line width=1,color={rgb:blue,1;green,1},looseness=0.5] (seqA5.south) to (g6);
+\draw[line width=1,color={rgb:blue,1;green,1},looseness=0.5] (seqB5.south) to (g6);
+
+\draw[line width=1,color=blue!100,looseness=0.5] (seqA6.south) to (g7);
+\draw[line width=1,color=blue!100,looseness=0.5] (seqB6.south) to (g7);
+
+\draw[line width=1,color={rgb:blue,1;purple,1},looseness=0.5] (seqA7.south) to (g8);
+\draw[line width=1,color={rgb:blue,1;purple,1},looseness=0.5] (seqC5.south) to (g8);
+
+\draw[line width=1,color=purple!100,looseness=0.5] (seqB7.south) to (g9);
+
+\draw[line width=1,color=gray,looseness=0.5] (seqB2.south) to (g3);
+
+\begin{scope}[node distance=0.5cm]
+  \node[squarednode,fill=red!100] (c1) [right = 1cm of seqA7] {1};
+  \node[squarednode,fill={rgb:red,1;orange,1}] (c2) [below of=c1] {2};
+  \node[squarednode,fill=orange!100] (c3) [below of=c2] {3};
+  \node[squarednode,fill=green!100] (c4) [below of=c3] {4};
+  \node[squarednode,fill={rgb:blue,1;green,1},text=white] (c5) [below of=c4] {5};
+  \node[squarednode,fill=blue!100,text=white] (c6) [below of=c5] {6};
+  \node[squarednode,fill={rgb:blue,1;purple,1},text=white] (c7) [below of=c6] {7};
+  \node[squarednode,fill=purple!100] (c8) [below of=c7] {8};
+  \node[squarednode,fill=gray] (c9) [below of=c8] {9};
+\end{scope}
+
+\end{pgfonlayer}
+
 %\draw[->] (uppercircle.south) -- (seq1.north);

### 2022-01-05 manuscript/fig_induction.tex
lighter lines

@@
 \begin{pgfonlayer}{bg}    % select the background layer
-\draw[line width=1,color=red!100,looseness=1] (seqA1.south) to (g1);
-\draw[line width=1,color=red!100,looseness=1] (seqB1.south) to (g1);
-\draw[line width=1,color=red!100,looseness=1] (seqC1.south) to (g1);
+\draw[line width=0.5,color=red!100,looseness=1] (seqA1.south) to (g1);
+\draw[line width=0.5,color=red!100,looseness=1] (seqB1.south) to (g1);
+\draw[line width=0.5,color=red!100,looseness=1] (seqC1.south) to (g1);
 
-\draw[line width=1,color={rgb:red,1;orange,1},looseness=0.5] (seqA2.south) to (g2);
-\draw[line width=1,color={rgb:red,1;orange,1},looseness=0.5] (seqC2.south) to (g2);
+\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=0.5] (seqA2.south) to (g2);
+\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=0.5] (seqC2.south) to (g2);
 
-\draw[line width=1,color=orange!100,looseness=0.5] (seqA3.south) to (g4);
-\draw[line width=1,color=orange!100,looseness=0.5] (seqB3.south) to (g4);
-\draw[line width=1,color=orange!100,looseness=0.5] (seqC3.south) to (g4);
+\draw[line width=0.5,color=orange!100,looseness=0.5] (seqA3.south) to (g4);
+\draw[line width=0.5,color=orange!100,looseness=0.5] (seqB3.south) to (g4);
+\draw[line width=0.5,color=orange!100,looseness=0.5] (seqC3.south) to (g4);
 
-\draw[line width=1,color=green!100,looseness=0.5] (seqA4.south) to (g5);
-\draw[line width=1,color=green!100,looseness=0.5] (seqB4.south) to (g5);
-%\draw[line width=1,color=blue!100,looseness=0.5] (seqC3.south) to (g4);
+\draw[line width=0.5,color=green!100,looseness=0.5] (seqA4.south) to (g5);
+\draw[line width=0.5,color=green!100,looseness=0.5] (seqB4.south) to (g5);
+%\draw[line width=0.5,color=blue!100,looseness=0.5] (seqC3.south) to (g4);
 
-\draw[line width=1,color={rgb:blue,1;green,1},looseness=0.5] (seqA5.south) to (g6);
-\draw[line width=1,color={rgb:blue,1;green,1},looseness=0.5] (seqB5.south) to (g6);
+\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=0.5] (seqA5.south) to (g6);
+\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=0.5] (seqB5.south) to (g6);
 
-\draw[line width=1,color=blue!100,looseness=0.5] (seqA6.south) to (g7);
-\draw[line width=1,color=blue!100,looseness=0.5] (seqB6.south) to (g7);
+\draw[line width=0.5,color=blue!100,looseness=0.5] (seqA6.south) to (g7);
+\draw[line width=0.5,color=blue!100,looseness=0.5] (seqB6.south) to (g7);
 
-\draw[line width=1,color={rgb:blue,1;purple,1},looseness=0.5] (seqA7.south) to (g8);
-\draw[line width=1,color={rgb:blue,1;purple,1},looseness=0.5] (seqC5.south) to (g8);
+\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=0.5] (seqA7.south) to (g8);
+\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=0.5] (seqC5.south) to (g8);
 
-\draw[line width=1,color=purple!100,looseness=0.5] (seqB7.south) to (g9);
+\draw[line width=0.5,color=purple!100,looseness=0.5] (seqB7.south) to (g9);
 
-\draw[line width=1,color=gray,looseness=0.5] (seqB2.south) to (g3);
+\draw[line width=0.5,color=gray,looseness=0.5] (seqB2.south) to (g3);
 

### 2022-01-05 manuscript/fig_induction.tex
smaller key

@@
     squarednode/.style={rectangle, draw=black!60, fill=gray!10, very thick, minimum size=5mm},
+    keynode/.style={rectangle, draw=black!60, fill=gray!10, thick, minimum size=1mm}
 ]
@@
 \begin{scope}[node distance=0.5cm]
-  \node[squarednode,fill=red!100] (c1) [right = 1cm of seqA7] {1};
-  \node[squarednode,fill={rgb:red,1;orange,1}] (c2) [below of=c1] {2};
-  \node[squarednode,fill=orange!100] (c3) [below of=c2] {3};
-  \node[squarednode,fill=green!100] (c4) [below of=c3] {4};
-  \node[squarednode,fill={rgb:blue,1;green,1},text=white] (c5) [below of=c4] {5};
-  \node[squarednode,fill=blue!100,text=white] (c6) [below of=c5] {6};
-  \node[squarednode,fill={rgb:blue,1;purple,1},text=white] (c7) [below of=c6] {7};
-  \node[squarednode,fill=purple!100] (c8) [below of=c7] {8};
-  \node[squarednode,fill=gray] (c9) [below of=c8] {9};
+  \node[keynode,fill=red!100] (c1) [right = 1cm of seqA7] {1};
+  \node[keynode,fill={rgb:red,1;orange,1}] (c2) [below of=c1] {2};
+  \node[keynode,fill=orange!100] (c3) [below of=c2] {3};
+  \node[keynode,fill=green!100] (c4) [below of=c3] {4};
+  \node[keynode,fill={rgb:blue,1;green,1},text=white] (c5) [below of=c4] {5};
+  \node[keynode,fill=blue!100,text=white] (c6) [below of=c5] {6};
+  \node[keynode,fill={rgb:blue,1;purple,1},text=white] (c7) [below of=c6] {7};
+  \node[keynode,fill=purple!100] (c8) [below of=c7] {8};
+  \node[keynode,fill=gray] (c9) [below of=c8] {9};
 \end{scope}

### 2022-01-05 manuscript/fig_induction.tex
maybe better

@@
 \usepackage{tikz}
+\usepackage{comment}
 \usetikzlibrary{positioning}
@@
 
-%Lines
-\draw (seqA1) -- (seqA2);
-\draw (seqA2) -- (seqA3);
-\draw (seqA3) -- (seqA4);
-\draw (seqA4) -- (seqA5);
-\draw (seqA5) -- (seqA6);
-\draw (seqA6) -- (seqA7);
-
-\draw (seqB1) -- (seqB2);
-\draw (seqB2) -- (seqB3);
-\draw (seqB3) -- (seqB4);
-\draw (seqB4) -- (seqB5);
-\draw (seqB5) -- (seqB6);
-\draw (seqB6) -- (seqB7);
-
 \node[squarednode] (seqC1) [below of=seqB1] {G};
@@
 
-\draw (seqC1) -- (seqC2);
-\draw (seqC2) -- (seqC3);
-\draw (seqC3) -- (seqC4);
-\draw (seqC4) -- (seqC5);
+%Lines
+\draw[line width=1.0] (seqA1) -- (seqA2);
+\draw[line width=1.0] (seqA2) -- (seqA3);
+\draw[line width=1.0] (seqA3) -- (seqA4);
+\draw[line width=1.0] (seqA4) -- (seqA5);
+\draw[line width=1.0] (seqA5) -- (seqA6);
+\draw[line width=1.0] (seqA6) -- (seqA7);
 
-%\begin{scope}[node distance=0.5cm]
-  \node[roundnode] (g2) [below = 3cm of seqA1] {A};
+\draw[line width=1.0] (seqB1) -- (seqB2);
+\draw[line width=1.0] (seqB2) -- (seqB3);
+\draw[line width=1.0] (seqB3) -- (seqB4);
+\draw[line width=1.0] (seqB4) -- (seqB5);
+\draw[line width=1.0] (seqB5) -- (seqB6);
+\draw[line width=1.0] (seqB6) -- (seqB7);
+
+\draw[line width=1.0] (seqC1) -- (seqC2);
+\draw[line width=1.0] (seqC2) -- (seqC3);
+\draw[line width=1.0] (seqC3) -- (seqC4);
+\draw[line width=1.0] (seqC4) -- (seqC5);
+
+\begin{scope}[node distance=1.3cm]
+\node[roundnode] (g2) [below right = 3cm and 1.5cm of seqA1] {A};
 \node[roundnode] (g1) [below left of=g2] {G};
-\node[roundnode] (g3) [below right of=g2] {C};
+\node[roundnode] (g3) [below right of=g1] {C};
 \node[roundnode] (g4) [above right of=g3] {T};
@@
 \node[roundnode] (g9) [below right of=g7] {G};
-%\end{scope}
+\end{scope}
 
-\draw (g1) to (g2);
-\draw (g1) to (g3);
-\draw (g2) to (g4);
-\draw (g3) to (g4);
-\draw (g4) to (g5);
-\draw (g4) to [looseness=1] (g7);
-\draw (g5) to (g6);
-\draw (g6) to (g7);
-\draw (g7) to (g8);
-\draw (g7) to (g9);
+\draw[line width=1.0] (g1) to (g2);
+\draw[line width=1.0] (g1) to (g3);
+\draw[line width=1.0] (g2) to (g4);
+\draw[line width=1.0] (g3) to (g4);
+\draw[line width=1.0] (g4) to (g5);
+\draw[line width=1.0] (g4) to [looseness=1] (g7);
+\draw[line width=1.0] (g5) to (g6);
+\draw[line width=1.0] (g6) to (g7);
+\draw[line width=1.0] (g7) to (g8);
+\draw[line width=1.0] (g7) to (g9);
 
@@
 
-\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=0.5] (seqA2.south) to (g2);
-\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=0.5] (seqC2.south) to (g2);
+\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=1.0] (seqA2.south) to (g2);
+\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=1.0] (seqC2.south) to (g2);
 
-\draw[line width=0.5,color=orange!100,looseness=0.5] (seqA3.south) to (g4);
-\draw[line width=0.5,color=orange!100,looseness=0.5] (seqB3.south) to (g4);
-\draw[line width=0.5,color=orange!100,looseness=0.5] (seqC3.south) to (g4);
+\draw[line width=0.5,color=orange!100,looseness=1.0] (seqA3.south) to (g4);
+\draw[line width=0.5,color=orange!100,looseness=1.0] (seqB3.south) to (g4);
+\draw[line width=0.5,color=orange!100,looseness=1.0] (seqC3.south) to (g4);
 
-\draw[line width=0.5,color=green!100,looseness=0.5] (seqA4.south) to (g5);
-\draw[line width=0.5,color=green!100,looseness=0.5] (seqB4.south) to (g5);
-%\draw[line width=0.5,color=blue!100,looseness=0.5] (seqC3.south) to (g4);
+\draw[line width=0.5,color=green!100,looseness=1.0] (seqA4.south) to (g5);
+\draw[line width=0.5,color=green!100,looseness=1.0] (seqB4.south) to (g5);
+%\draw[line width=0.5,color=blue!100,looseness=1.0] (seqC3.south) to (g4);
 
-\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=0.5] (seqA5.south) to (g6);
-\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=0.5] (seqB5.south) to (g6);
+\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=1.0] (seqA5.south) to (g6);
+\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=1.0] (seqB5.south) to (g6);
 
-\draw[line width=0.5,color=blue!100,looseness=0.5] (seqA6.south) to (g7);
-\draw[line width=0.5,color=blue!100,looseness=0.5] (seqB6.south) to (g7);
+\draw[line width=0.5,color=blue!100,looseness=1.0] (seqA6.south) to (g7);
+\draw[line width=0.5,color=blue!100,looseness=1.0] (seqB6.south) to (g7);
 
-\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=0.5] (seqA7.south) to (g8);
-\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=0.5] (seqC5.south) to (g8);
+\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=1.0] (seqA7.south) to (g8);
+\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=1.0] (seqC5.south) to (g8);
 
-\draw[line width=0.5,color=purple!100,looseness=0.5] (seqB7.south) to (g9);
+\draw[line width=0.5,color=purple!100,looseness=1.0] (seqB7.south) to (g9);
 
-\draw[line width=0.5,color=gray,looseness=0.5] (seqB2.south) to (g3);
+\draw[line width=0.5,color=gray,looseness=1.0] (seqB2.south) to (g3);
 
+\begin{comment}
 \begin{scope}[node distance=0.5cm]
@@
 \end{scope}
+\end{comment}
 

### 2022-01-08 manuscript/fig_induction.tex
correct a bug
yes

@@
 
-%\node[roundnode]        (uppercircle)       [above=of seqA1] {1};
-%\node[squarednode]      (rightsquare)       [below=of seqA1] {3};
-%\node[roundnode]        (lowercircle)       [below=of seqA1] {4};
-
 \node[squarednode] (seqB1) [below of=seqA1] {G};
@@
 \node[squarednode] (seqC3) [right of=seqC2] {T};
-\node[squarednode] (seqC4) [right of=seqC3] {C};
+\node[squarednode] (seqC4) [below of=seqB6] {C};
 \node[squarednode] (seqC5) [right of=seqC4] {A};
@@
 
-\begin{scope}[node distance=1.3cm]
-\node[roundnode] (g2) [below right = 3cm and 1.5cm of seqA1] {A};
-\node[roundnode] (g1) [below left of=g2] {G};
-\node[roundnode] (g3) [below right of=g1] {C};
-\node[roundnode] (g4) [above right of=g3] {T};
-\node[roundnode] (g5) [right of=g4] {T};
-\node[roundnode] (g6) [right of=g5] {A};
-\node[roundnode] (g7) [right of=g6] {C};
-\node[roundnode] (g8) [above right of=g7] {A};
-\node[roundnode] (g9) [below right of=g7] {G};
-\end{scope}
+% alignments
+\draw[line width=1.5,color=gray!50] (seqA1) -- (seqB1);
+\draw[line width=1.5,color=gray!50] (seqA1) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC1);
+%\draw[line width=1.0,color=gray!100] (seqA2) -- (seqB2);
+\draw[line width=1.5,color=gray!50] (seqA2) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC2);
+
+\draw[line width=1.5,color=gray!50] (seqA3) -- (seqB3);
+\draw[line width=1.5,color=gray!50] (seqA3) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC3);
+\draw[line width=1.5,color=gray!50] (seqB3) -- (seqC3);
+
+\draw[line width=1.5,color=gray!50] (seqA4) -- (seqB4);
+\draw[line width=1.5,color=gray!50] (seqA5) -- (seqB5);
+
+%\draw[line width=1.5,color=gray!50] (seqA6) -- (seqB6);
+\draw[line width=1.5,color=gray!50] (seqA6) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC4);
+\draw[line width=1.5,color=gray!50] (seqB6) -- (seqC4);
+
+%\draw[line width=1.5,color=gray!50] (seqA7) -- (seqB7);
+\draw[line width=1.5,color=gray!50] (seqA7) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC5);
+%\draw[line width=1.5,color=gray!50] (seqB7) -- (seqC5);
+
+%    \draw (i) to[out=150, in=210, looseness=4, edge node={node [sloped,below] {$\leftarrow$}}] (l);
+%    \draw (k) to[out=30, in=-30, looseness=4, edge node={node [sloped,above] {$\leftarrow$}}] (j);
+
+%\begin{scope}[node distance=1cm]
+\node[squarednode] (g2) [below = 0.75cm of seqC2] {A};
+\node[squarednode] (g1) [below = 1.25cm of seqC1] {G};
+\node[squarednode] (g3) [below = 0.5cm of g2] {C};
+\node[squarednode] (g4) [below = 1.25cm of seqC3] {T};
+\node[squarednode] (g5) [right of=g4] {T};
+\node[squarednode] (g6) [right of=g5] {A};
+\node[squarednode] (g7) [below = 1.25cm of seqC4] {C};
+\node[squarednode] (g8) [below = 0.75cm of seqC5] {A};
+\node[squarednode] (g9) [below = 0.5cm of g8] {G};
+%\end{scope}
 
@@
 \begin{pgfonlayer}{bg}    % select the background layer
-\draw[line width=0.5,color=red!100,looseness=1] (seqA1.south) to (g1);
-\draw[line width=0.5,color=red!100,looseness=1] (seqB1.south) to (g1);
-\draw[line width=0.5,color=red!100,looseness=1] (seqC1.south) to (g1);
+\draw[line width=1.0,color=red!100] (seqA1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[line width=1.0,color=red!100] (seqB1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[line width=1.0,color=red!100] (seqC1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+%\draw[line width=1.0,color=red!100,looseness=0.5] (seqB1.south) to (g1.north);
+%\draw[line width=1.0,color=red!100,looseness=0.5] (seqC1.south) to (g1.north);
 
-\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=1.0] (seqA2.south) to (g2);
-\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=1.0] (seqC2.south) to (g2);
+\draw[line width=1.0,color={rgb:red,1;orange,1}] (seqA2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
+\draw[line width=1.0,color={rgb:red,1;orange,1}] (seqC2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
 
-\draw[line width=0.5,color=orange!100,looseness=1.0] (seqA3.south) to (g4);
-\draw[line width=0.5,color=orange!100,looseness=1.0] (seqB3.south) to (g4);
-\draw[line width=0.5,color=orange!100,looseness=1.0] (seqC3.south) to (g4);
+\draw[line width=1.0,color=orange!100] (seqA3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[line width=1.0,color=orange!100] (seqB3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[line width=1.0,color=orange!100] (seqC3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
 
-\draw[line width=0.5,color=green!100,looseness=1.0] (seqA4.south) to (g5);
-\draw[line width=0.5,color=green!100,looseness=1.0] (seqB4.south) to (g5);
-%\draw[line width=0.5,color=blue!100,looseness=1.0] (seqC3.south) to (g4);
+\draw[line width=1.0,color=green!100] (seqA4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
+\draw[line width=1.0,color=green!100] (seqB4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
+%\draw[line width=1.0,color=blue!100] (seqC3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
 
-\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=1.0] (seqA5.south) to (g6);
-\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=1.0] (seqB5.south) to (g6);
+\draw[line width=1.0,color={rgb:blue,1;green,1}] (seqA5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
+\draw[line width=1.0,color={rgb:blue,1;green,1}] (seqB5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
 
-\draw[line width=0.5,color=blue!100,looseness=1.0] (seqA6.south) to (g7);
-\draw[line width=0.5,color=blue!100,looseness=1.0] (seqB6.south) to (g7);
+\draw[line width=1.0,color=blue!100] (seqA6) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[line width=1.0,color=blue!100] (seqB6) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[line width=1.0,color=blue!100] (seqC4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
 
-\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=1.0] (seqA7.south) to (g8);
-\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=1.0] (seqC5.south) to (g8);
+\draw[line width=1.0,color={rgb:blue,1;purple,1}] (seqA7) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
+\draw[line width=1.0,color={rgb:blue,1;purple,1}] (seqC5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
 
-\draw[line width=0.5,color=purple!100,looseness=1.0] (seqB7.south) to (g9);
+\draw[line width=1.0,color=purple!100] (seqB7) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g9);
 
-\draw[line width=0.5,color=gray,looseness=1.0] (seqB2.south) to (g3);
+\draw[line width=1.0,color=brown] (seqB2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g3);
 

### 2022-01-08 manuscript/fig_induction.tex
sanserif

--- 2022-01-05 manuscript/fig_induction.tex
@@
     node distance= 10mm and 20mm,
+    font=\sffamily,
     roundnode/.style={circle, draw=black!60, fill=gray!10, very thick, minimum size=7mm},
     squarednode/.style={rectangle, draw=black!60, fill=gray!10, very thick, minimum size=5mm},
+    graphnode/.style={rectangle, draw=gray!60, fill=black!100, text=white, very thick, minimum size=5mm},
     keynode/.style={rectangle, draw=black!60, fill=gray!10, thick, minimum size=1mm}
@@
 \node[squarednode] (seqA7) [right of=seqA6] {A};
-
-%\node[roundnode]        (uppercircle)       [above=of seqA1] {1};
-%\node[squarednode]      (rightsquare)       [below=of seqA1] {3};
-%\node[roundnode]        (lowercircle)       [below=of seqA1] {4};
 
@@
 %Lines
-\draw[line width=1.0] (seqA1) -- (seqA2);
-\draw[line width=1.0] (seqA2) -- (seqA3);
-\draw[line width=1.0] (seqA3) -- (seqA4);
-\draw[line width=1.0] (seqA4) -- (seqA5);
-\draw[line width=1.0] (seqA5) -- (seqA6);
-\draw[line width=1.0] (seqA6) -- (seqA7);
+\draw[very thick] (seqA1) -- (seqA2);
+\draw[very thick] (seqA2) -- (seqA3);
+\draw[very thick] (seqA3) -- (seqA4);
+\draw[very thick] (seqA4) -- (seqA5);
+\draw[very thick] (seqA5) -- (seqA6);
+\draw[very thick] (seqA6) -- (seqA7);
 
-\draw[line width=1.0] (seqB1) -- (seqB2);
-\draw[line width=1.0] (seqB2) -- (seqB3);
-\draw[line width=1.0] (seqB3) -- (seqB4);
-\draw[line width=1.0] (seqB4) -- (seqB5);
-\draw[line width=1.0] (seqB5) -- (seqB6);
-\draw[line width=1.0] (seqB6) -- (seqB7);
+\draw[very thick] (seqB1) -- (seqB2);
+\draw[very thick] (seqB2) -- (seqB3);
+\draw[very thick] (seqB3) -- (seqB4);
+\draw[very thick] (seqB4) -- (seqB5);
+\draw[very thick] (seqB5) -- (seqB6);
+\draw[very thick] (seqB6) -- (seqB7);
 
-\draw[line width=1.0] (seqC1) -- (seqC2);
-\draw[line width=1.0] (seqC2) -- (seqC3);
-\draw[line width=1.0] (seqC3) -- (seqC4);
-\draw[line width=1.0] (seqC4) -- (seqC5);
+\draw[very thick] (seqC1) -- (seqC2);
+\draw[very thick] (seqC2) -- (seqC3);
+\draw[very thick] (seqC3) -- (seqC4);
+\draw[very thick] (seqC4) -- (seqC5);
 
-\begin{scope}[node distance=1.3cm]
-\node[roundnode] (g2) [below right = 3cm and 1.5cm of seqA1] {A};
-\node[roundnode] (g1) [below left of=g2] {G};
-\node[roundnode] (g3) [below right of=g1] {C};
-\node[roundnode] (g4) [above right of=g3] {T};
-\node[roundnode] (g5) [right of=g4] {T};
-\node[roundnode] (g6) [right of=g5] {A};
-\node[roundnode] (g7) [right of=g6] {C};
-\node[roundnode] (g8) [above right of=g7] {A};
-\node[roundnode] (g9) [below right of=g7] {G};
-\end{scope}
+% alignments
+\draw[very thick,color=gray!50] (seqA1) -- (seqB1);
+\draw[very thick,color=gray!50] (seqA1) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC1);
+%\draw[line width=1.0,color=gray!100] (seqA2) -- (seqB2);
+\draw[very thick,color=gray!50] (seqA2) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC2);
+
+\draw[very thick,color=gray!50] (seqA3) -- (seqB3);
+\draw[very thick,color=gray!50] (seqA3) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC3);
+\draw[very thick,color=gray!50] (seqB3) -- (seqC3);
+
+\draw[very thick,color=gray!50] (seqA4) -- (seqB4);
+\draw[very thick,color=gray!50] (seqA5) -- (seqB5);
+
+%\draw[very thick,color=gray!50] (seqA6) -- (seqB6);
+\draw[very thick,color=gray!50] (seqA6) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC4);
+\draw[very thick,color=gray!50] (seqB6) -- (seqC4);
+
+%\draw[very thick,color=gray!50] (seqA7) -- (seqB7);
+\draw[very thick,color=gray!50] (seqA7) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC5);
+%\draw[very thick,dashed,color=gray!50] (seqB7) -- (seqC5);
+
+%    \draw (i) to[out=150, in=210, looseness=4, edge node={node [sloped,below] {$\leftarrow$}}] (l);
+%    \draw (k) to[out=30, in=-30, looseness=4, edge node={node [sloped,above] {$\leftarrow$}}] (j);
+
+%\begin{scope}[node distance=1cm]
+\node[graphnode] (g2) [below = 0.75cm of seqC2] {2};
+\node[graphnode] (g1) [below = 1.25cm of seqC1] {1};
+\node[graphnode] (g3) [below = 0.5cm of g2] {8};
+\node[graphnode] (g4) [below = 1.25cm of seqC3] {3};
+\node[graphnode] (g5) [right of=g4] {4};
+\node[graphnode] (g6) [right of=g5] {5};
+\node[graphnode] (g7) [below = 1.25cm of seqC4] {6};
+\node[graphnode] (g8) [below = 0.75cm of seqC5] {7};
+\node[graphnode] (g9) [below = 0.5cm of g8] {9};
+%\end{scope}
 
@@
 
-\begin{pgfonlayer}{bg}    % select the background layer
-\draw[line width=0.5,color=red!100,looseness=1] (seqA1.south) to (g1);
-\draw[line width=0.5,color=red!100,looseness=1] (seqB1.south) to (g1);
-\draw[line width=0.5,color=red!100,looseness=1] (seqC1.south) to (g1);
+%\begin{pgfonlayer}{bg}    % select the background layer
+\draw[very thick,color=red!100] (seqA1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[very thick,color=red!100] (seqB1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[very thick,color=red!100] (seqC1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+%\draw[very thick,color=red!100,looseness=0.5] (seqB1.south) to (g1.north);
+%\draw[very thick,color=red!100,looseness=0.5] (seqC1.south) to (g1.north);
 
-\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=1.0] (seqA2.south) to (g2);
-\draw[line width=0.5,color={rgb:red,1;orange,1},looseness=1.0] (seqC2.south) to (g2);
+\draw[very thick,color={rgb:red,1;orange,1}] (seqA2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
+\draw[very thick,color={rgb:red,1;orange,1}] (seqC2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
 
-\draw[line width=0.5,color=orange!100,looseness=1.0] (seqA3.south) to (g4);
-\draw[line width=0.5,color=orange!100,looseness=1.0] (seqB3.south) to (g4);
-\draw[line width=0.5,color=orange!100,looseness=1.0] (seqC3.south) to (g4);
+\draw[very thick,color=orange!100] (seqA3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=orange!100] (seqB3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=orange!100] (seqC3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
 
-\draw[line width=0.5,color=green!100,looseness=1.0] (seqA4.south) to (g5);
-\draw[line width=0.5,color=green!100,looseness=1.0] (seqB4.south) to (g5);
-%\draw[line width=0.5,color=blue!100,looseness=1.0] (seqC3.south) to (g4);
+\draw[very thick,color=green!100] (seqA4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
+\draw[very thick,color=green!100] (seqB4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
+%\draw[very thick,color=blue!100] (seqC3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
 
-\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=1.0] (seqA5.south) to (g6);
-\draw[line width=0.5,color={rgb:blue,1;green,1},looseness=1.0] (seqB5.south) to (g6);
+\draw[very thick,color={rgb:blue,1;green,1}] (seqA5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
+\draw[very thick,color={rgb:blue,1;green,1}] (seqB5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
 
-\draw[line width=0.5,color=blue!100,looseness=1.0] (seqA6.south) to (g7);
-\draw[line width=0.5,color=blue!100,looseness=1.0] (seqB6.south) to (g7);
+\draw[very thick,color=blue!100] (seqA6) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[very thick,color=blue!100] (seqB6) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[very thick,color=blue!100] (seqC4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
 
-\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=1.0] (seqA7.south) to (g8);
-\draw[line width=0.5,color={rgb:blue,1;purple,1},looseness=1.0] (seqC5.south) to (g8);
+\draw[very thick,color={rgb:blue,1;purple,1}] (seqA7) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
+\draw[very thick,color={rgb:blue,1;purple,1}] (seqC5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
 
-\draw[line width=0.5,color=purple!100,looseness=1.0] (seqB7.south) to (g9);
+\draw[very thick,color=purple!100] (seqB7) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g9);
 
-\draw[line width=0.5,color=gray,looseness=1.0] (seqB2.south) to (g3);
+\draw[very thick,color=brown] (seqB2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g3);
 
@@
 
-\end{pgfonlayer}
+%\end{pgfonlayer}
 

### 2022-01-08 manuscript/fig_induction.tex
vizual pruuf

@@
 %Nodes
-\node[squarednode] (seqA1) {G};
+\node[squarednode,label=left:text1] (seqA1) {G};
 \node[squarednode] (seqA2) [right of=seqA1] {A};
@@
 
-\node[squarednode] (seqB1) [below of=seqA1] {G};
+\node[squarednode,label=left:text2] (seqB1) [below of=seqA1] {G};
 \node[squarednode] (seqB2) [right of=seqB1] {C};
@@
 
-\node[squarednode] (seqC1) [below of=seqB1] {G};
+\node[squarednode,label=left:text3] (seqC1) [below of=seqB1] {G};
 \node[squarednode] (seqC2) [right of=seqC1] {A};
@@
 \draw[very thick,color=gray!50] (seqA1) -- (seqB1);
-\draw[very thick,color=gray!50] (seqA1) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC1);
-%\draw[line width=1.0,color=gray!100] (seqA2) -- (seqB2);
+%\draw[very thick,color=gray!50] (seqA1) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC1);
+\draw[very thick,color=gray!50] (seqB1) -- (seqC1);
+
 \draw[very thick,color=gray!50] (seqA2) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC2);
@@
 %\begin{scope}[node distance=1cm]
-\node[graphnode] (g2) [below = 0.75cm of seqC2] {2};
-\node[graphnode] (g1) [below = 1.25cm of seqC1] {1};
-\node[graphnode] (g3) [below = 0.5cm of g2] {8};
-\node[graphnode] (g4) [below = 1.25cm of seqC3] {3};
-\node[graphnode] (g5) [right of=g4] {4};
-\node[graphnode] (g6) [right of=g5] {5};
-\node[graphnode] (g7) [below = 1.25cm of seqC4] {6};
-\node[graphnode] (g8) [below = 0.75cm of seqC5] {7};
-\node[graphnode] (g9) [below = 0.5cm of g8] {9};
+\node[graphnode,draw={rgb:red,1;orange,1}] (g2) [below = 0.75cm of seqC2] {2};
+\node[graphnode,draw=red!100,label=left:graph] (g1) [below = 1.25cm of seqC1] {1};
+\node[graphnode,draw=brown] (g3) [below = 0.5cm of g2] {8};
+\node[graphnode,draw=orange!100] (g4) [below = 1.25cm of seqC3] {3};
+\node[graphnode,draw=green!100] (g5) [right of=g4] {4};
+\node[graphnode,draw={rgb:blue,1;green,1}] (g6) [right of=g5] {5};
+\node[graphnode,draw=blue!100] (g7) [below = 1.25cm of seqC4] {6};
+\node[graphnode,draw={rgb:blue,1;purple,1}] (g8) [below = 0.75cm of seqC5] {7};
+\node[graphnode,draw=purple!100] (g9) [below = 0.5cm of g8] {9};
 %\end{scope}
 
-\draw[line width=1.0] (g1) to (g2);
-\draw[line width=1.0] (g1) to (g3);
-\draw[line width=1.0] (g2) to (g4);
-\draw[line width=1.0] (g3) to (g4);
-\draw[line width=1.0] (g4) to (g5);
-\draw[line width=1.0] (g4) to [looseness=1] (g7);
-\draw[line width=1.0] (g5) to (g6);
-\draw[line width=1.0] (g6) to (g7);
-\draw[line width=1.0] (g7) to (g8);
-\draw[line width=1.0] (g7) to (g9);
+\draw[very thick] (g1) to (g2);
+\draw[very thick] (g1) to (g3);
+\draw[very thick] (g2) to (g4);
+\draw[very thick] (g3) to (g4);
+\draw[very thick] (g4) to (g5);
+\draw[very thick] (g4) to [looseness=1] (g7);
+\draw[very thick] (g5) to (g6);
+\draw[very thick] (g6) to (g7);
+\draw[very thick] (g7) to (g8);
+\draw[very thick] (g7) to (g9);
 
 %\begin{pgfonlayer}{bg}    % select the background layer
-\draw[very thick,color=red!100] (seqA1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
-\draw[very thick,color=red!100] (seqB1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
-\draw[very thick,color=red!100] (seqC1) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[very thick,color=red!100] (seqA1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[very thick,color=red!100] (seqB1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[very thick,color=red!100] (seqC1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
 %\draw[very thick,color=red!100,looseness=0.5] (seqB1.south) to (g1.north);
@@
 
-\draw[very thick,color={rgb:red,1;orange,1}] (seqA2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
-\draw[very thick,color={rgb:red,1;orange,1}] (seqC2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
+\draw[very thick,color={rgb:red,1;orange,1}] (seqA2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
+\draw[very thick,color={rgb:red,1;orange,1}] (seqC2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
 
-\draw[very thick,color=orange!100] (seqA3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
-\draw[very thick,color=orange!100] (seqB3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
-\draw[very thick,color=orange!100] (seqC3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=orange!100] (seqA3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=orange!100] (seqB3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=orange!100] (seqC3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
 
-\draw[very thick,color=green!100] (seqA4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
-\draw[very thick,color=green!100] (seqB4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
-%\draw[very thick,color=blue!100] (seqC3) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=green!100] (seqA4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
+\draw[very thick,color=green!100] (seqB4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
+%\draw[very thick,color=blue!100] (seqC3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
 
-\draw[very thick,color={rgb:blue,1;green,1}] (seqA5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
-\draw[very thick,color={rgb:blue,1;green,1}] (seqB5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
+\draw[very thick,color={rgb:blue,1;green,1}] (seqA5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
+\draw[very thick,color={rgb:blue,1;green,1}] (seqB5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
 
-\draw[very thick,color=blue!100] (seqA6) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
-\draw[very thick,color=blue!100] (seqB6) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
-\draw[very thick,color=blue!100] (seqC4) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[very thick,color=blue!100] (seqA6) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[very thick,color=blue!100] (seqB6) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[very thick,color=blue!100] (seqC4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
 
-\draw[very thick,color={rgb:blue,1;purple,1}] (seqA7) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
-\draw[very thick,color={rgb:blue,1;purple,1}] (seqC5) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
+\draw[very thick,color={rgb:blue,1;purple,1}] (seqA7) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
+\draw[very thick,color={rgb:blue,1;purple,1}] (seqC5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
 
-\draw[very thick,color=purple!100] (seqB7) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g9);
+\draw[very thick,color=purple!100] (seqB7) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g9);
 
-\draw[very thick,color=brown] (seqB2) to [looseness=0.3,in=45,out=-45, edge node={node [sloped,above] {}}] (g3);
+\draw[very thick,color=brown] (seqB2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g3);
 

### 2022-01-08 manuscript/fig_induction.tex
gradient colors

@@
     squarednode/.style={rectangle, draw=black!60, fill=gray!10, very thick, minimum size=5mm},
-    graphnode/.style={rectangle, draw=gray!60, fill=black!100, text=white, very thick, minimum size=5mm},
+    graphnode/.style={rectangle, draw=black!60, fill=black!100, text=white, very thick, minimum size=5mm},
     keynode/.style={rectangle, draw=black!60, fill=gray!10, thick, minimum size=1mm}
+    %draw matchclosure/.style={very thick,color=gray!100}
 ]
@@
 % alignments
-\draw[very thick,color=gray!50] (seqA1) -- (seqB1);
-%\draw[very thick,color=gray!50] (seqA1) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC1);
-\draw[very thick,color=gray!50] (seqB1) -- (seqC1);
+\draw[very thick,dashed] (seqA1) -- (seqB1);
+%\draw[very thick,dashed] (seqA1) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC1);
+\draw[very thick,dashed] (seqB1) -- (seqC1);
 
-\draw[very thick,color=gray!50] (seqA2) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC2);
+\draw[very thick,dashed] (seqA2) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC2);
 
-\draw[very thick,color=gray!50] (seqA3) -- (seqB3);
-\draw[very thick,color=gray!50] (seqA3) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC3);
-\draw[very thick,color=gray!50] (seqB3) -- (seqC3);
+\draw[very thick,dashed] (seqA3) -- (seqB3);
+\draw[very thick,dashed] (seqA3) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC3);
+\draw[very thick,dashed] (seqB3) -- (seqC3);
 
-\draw[very thick,color=gray!50] (seqA4) -- (seqB4);
-\draw[very thick,color=gray!50] (seqA5) -- (seqB5);
+\draw[very thick,dashed] (seqA4) -- (seqB4);
+\draw[very thick,dashed] (seqA5) -- (seqB5);
 
-%\draw[very thick,color=gray!50] (seqA6) -- (seqB6);
-\draw[very thick,color=gray!50] (seqA6) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC4);
-\draw[very thick,color=gray!50] (seqB6) -- (seqC4);
+%\draw[very thick,dashed] (seqA6) -- (seqB6);
+\draw[very thick,dashed] (seqA6) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC4);
+\draw[very thick,dashed] (seqB6) -- (seqC4);
 
-%\draw[very thick,color=gray!50] (seqA7) -- (seqB7);
-\draw[very thick,color=gray!50] (seqA7) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC5);
+%\draw[very thick,dashed] (seqA7) -- (seqB7);
+\draw[very thick,dashed] (seqA7) to [in=135,out=225,looseness=0.5, edge node={node [sloped,above] {}}] (seqC5);
 %\draw[very thick,dashed,color=gray!50] (seqB7) -- (seqC5);
@@
 %\begin{scope}[node distance=1cm]
-\node[graphnode,draw={rgb:red,1;orange,1}] (g2) [below = 0.75cm of seqC2] {2};
-\node[graphnode,draw=red!100,label=left:graph] (g1) [below = 1.25cm of seqC1] {1};
-\node[graphnode,draw=brown] (g3) [below = 0.5cm of g2] {8};
-\node[graphnode,draw=orange!100] (g4) [below = 1.25cm of seqC3] {3};
-\node[graphnode,draw=green!100] (g5) [right of=g4] {4};
-\node[graphnode,draw={rgb:blue,1;green,1}] (g6) [right of=g5] {5};
-\node[graphnode,draw=blue!100] (g7) [below = 1.25cm of seqC4] {6};
-\node[graphnode,draw={rgb:blue,1;purple,1}] (g8) [below = 0.75cm of seqC5] {7};
-\node[graphnode,draw=purple!100] (g9) [below = 0.5cm of g8] {9};
+\node[graphnode,draw=gray!25] (g2) [below = 0.75cm of seqC2] {2};
+\node[graphnode,label=left:graph,draw=gray!20] (g1) [below = 1.25cm of seqC1] {1};
+\node[graphnode,draw=gray!80] (g3) [below = 0.5cm of g2] {8};
+\node[graphnode,draw=gray!30] (g4) [below = 1.25cm of seqC3] {3};
+\node[graphnode,draw=gray!35] (g5) [right of=g4] {4};
+\node[graphnode,draw=gray!40] (g6) [right of=g5] {5};
+\node[graphnode,draw=gray!50] (g7) [below = 1.25cm of seqC4] {6};
+\node[graphnode,draw=gray!60] (g8) [below = 0.75cm of seqC5] {7};
+\node[graphnode,draw=gray!95] (g9) [below = 0.5cm of g8] {9};
 %\end{scope}
@@
 %\begin{pgfonlayer}{bg}    % select the background layer
-\draw[very thick,color=red!100] (seqA1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
-\draw[very thick,color=red!100] (seqB1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
-\draw[very thick,color=red!100] (seqC1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
-%\draw[very thick,color=red!100,looseness=0.5] (seqB1.south) to (g1.north);
-%\draw[very thick,color=red!100,looseness=0.5] (seqC1.south) to (g1.north);
+\draw[very thick,color=gray!20] (seqA1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[very thick,color=gray!20] (seqB1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+\draw[very thick,color=gray!20] (seqC1) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g1);
+%\draw[very thick,color=gray!50,color=red!50,looseness=0.5] (seqB1.south) to (g1.north);
+%\draw[very thick,color=gray!50,color=red!50,looseness=0.5] (seqC1.south) to (g1.north);
 
-\draw[very thick,color={rgb:red,1;orange,1}] (seqA2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
-\draw[very thick,color={rgb:red,1;orange,1}] (seqC2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
+\draw[very thick,color=gray!25] (seqA2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
+\draw[very thick,color=gray!25] (seqC2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g2);
 
-\draw[very thick,color=orange!100] (seqA3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
-\draw[very thick,color=orange!100] (seqB3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
-\draw[very thick,color=orange!100] (seqC3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=gray!30] (seqA3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=gray!30] (seqB3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=gray!30] (seqC3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
 
-\draw[very thick,color=green!100] (seqA4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
-\draw[very thick,color=green!100] (seqB4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
-%\draw[very thick,color=blue!100] (seqC3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
+\draw[very thick,color=gray!35] (seqA4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
+\draw[very thick,color=gray!35] (seqB4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g5);
+%\draw[very thick,color=gray!50,color=blue!50] (seqC3) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g4);
 
-\draw[very thick,color={rgb:blue,1;green,1}] (seqA5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
-\draw[very thick,color={rgb:blue,1;green,1}] (seqB5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
+\draw[very thick,color=gray!40] (seqA5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
+\draw[very thick,color=gray!40] (seqB5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g6);
 
-\draw[very thick,color=blue!100] (seqA6) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
-\draw[very thick,color=blue!100] (seqB6) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
-\draw[very thick,color=blue!100] (seqC4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[very thick,color=gray!50] (seqA6) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[very thick,color=gray!50] (seqB6) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
+\draw[very thick,color=gray!50] (seqC4) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g7);
 
-\draw[very thick,color={rgb:blue,1;purple,1}] (seqA7) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
-\draw[very thick,color={rgb:blue,1;purple,1}] (seqC5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
+\draw[very thick,color=gray!60] (seqA7) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
+\draw[very thick,color=gray!60] (seqC5) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g8);
 
-\draw[very thick,color=purple!100] (seqB7) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g9);
+\draw[very thick,color=gray!95] (seqB7) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g9);
 
-\draw[very thick,color=brown] (seqB2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g3);
+\draw[very thick,color=gray!80] (seqB2) to [looseness=0.4,in=45,out=-45, edge node={node [sloped,above] {}}] (g3);
 

### 2022-01-08 manuscript/fig_from_alignment_to_variation_graph.tex
drop xcolor
get the first figure in

@@
 %  \centering
-    \floatbox[{\capbeside\thisfloatsetup{capbesideposition={left,top},capbesidewidth=0.45\linewidth}}]{figure}[\FBwidth]
-    {\caption{\FIXME{put high-resolution image} Building variation graphs from alignment graphs. Top: an alignment graph between a set of sequences,
-        showing alignment edges in grey and precedence (path) edges in colors associated with each path. Bottom: the variation graph
-        resulting from the application of the transitive closure process in \textit{seqwish}, where aligned segments have been
-        compressed into a single node.
-    }\label{fig:from_alignment_to_variation_graph}}
+   \includegraphics[width=\linewidth]{fig_induction.pdf}
+   \caption{
+   A visual description of variation graph induction.
+   At top, the alignment graph model.
+   %   At bottom, a sequence graph obtained by contracting
+
+   %   At top, three sequences (text1, text2, text3) 
+    }
     %{\includegraphics[width=\linewidth]{fig/OUP_First_SBk_Bot_8401}} %\includegraphics[width=5cm]{name}}
     %\caption{Performance evaluation of \textit{odgi build} when translating a 90-haplotype graph of human chromosome 6 into ODGI's native format.}
-
 \end{figure}

### 2022-01-08 manuscript/fig_induction.tex
get the first figure in
not like this

@@
-\documentclass{article}
-\usepackage{tikz}
+\documentclass[tikz,margin=0pt]{standalone}
+
+%\setlength{\paperwidth}{108mm}
+%\setlength{\paperheight}{210mm}
+%\ifx\pdfoutput\undefined
+%\else
+%\setlength{\pdfpagewidth}{\paperwidth}
+%\setlength{\pdfpageheight}{\paperheight}
+%\fi
+
+%\usepackage{tikz}
 \usepackage{comment}
@@
     graphnode/.style={rectangle, draw=black!60, fill=black!100, text=white, very thick, minimum size=5mm},
-    keynode/.style={rectangle, draw=black!60, fill=gray!10, thick, minimum size=1mm}
+    keynode/.style={rectangle, draw=black!60, fill=gray!10, thick, minimum size=1mm},
+    pathnode/.style={rectangle, minimum size=5mm},
     %draw matchclosure/.style={very thick,color=gray!100}
@@
 %Lines
-\draw[very thick] (seqA1) -- (seqA2);
-\draw[very thick] (seqA2) -- (seqA3);
-\draw[very thick] (seqA3) -- (seqA4);
-\draw[very thick] (seqA4) -- (seqA5);
-\draw[very thick] (seqA5) -- (seqA6);
-\draw[very thick] (seqA6) -- (seqA7);
+\draw[very thick,color=red] (seqA1) -- (seqA2);
+\draw[very thick,color=red] (seqA2) -- (seqA3);
+\draw[very thick,color=red] (seqA3) -- (seqA4);
+\draw[very thick,color=red] (seqA4) -- (seqA5);
+\draw[very thick,color=red] (seqA5) -- (seqA6);
+\draw[very thick,color=red] (seqA6) -- (seqA7);
 
-\draw[very thick] (seqB1) -- (seqB2);
-\draw[very thick] (seqB2) -- (seqB3);
-\draw[very thick] (seqB3) -- (seqB4);
-\draw[very thick] (seqB4) -- (seqB5);
-\draw[very thick] (seqB5) -- (seqB6);
-\draw[very thick] (seqB6) -- (seqB7);
+\draw[very thick,color=green] (seqB1) -- (seqB2);
+\draw[very thick,color=green] (seqB2) -- (seqB3);
+\draw[very thick,color=green] (seqB3) -- (seqB4);
+\draw[very thick,color=green] (seqB4) -- (seqB5);
+\draw[very thick,color=green] (seqB5) -- (seqB6);
+\draw[very thick,color=green] (seqB6) -- (seqB7);
 
-\draw[very thick] (seqC1) -- (seqC2);
-\draw[very thick] (seqC2) -- (seqC3);
-\draw[very thick] (seqC3) -- (seqC4);
-\draw[very thick] (seqC4) -- (seqC5);
+\draw[very thick,color=blue] (seqC1) -- (seqC2);
+\draw[very thick,color=blue] (seqC2) -- (seqC3);
+\draw[very thick,color=blue] (seqC3) -- (seqC4);
+\draw[very thick,color=blue] (seqC4) -- (seqC5);
 
@@
 
+\node[pathnode,label=left:path1] (pathA1) [below = 1cm of g1] {1};
+\node[pathnode] (pathA2) [right of=pathA1] {2};
+\node[pathnode] (pathA3) [right of=pathA2] {3};
+\node[pathnode] (pathA4) [right of=pathA3] {4};
+\node[pathnode] (pathA5) [right of=pathA4] {5};
+\node[pathnode] (pathA6) [right of=pathA5] {6};
+\node[pathnode] (pathA7) [right of=pathA6] {7};
+
+\node[pathnode,label=left:path2] (pathB1) [below of=pathA1] {1};
+\node[pathnode] (pathB2) [right of=pathB1] {8};
+\node[pathnode] (pathB3) [right of=pathB2] {3};
+\node[pathnode] (pathB4) [right of=pathB3] {4};
+\node[pathnode] (pathB5) [right of=pathB4] {5};
+\node[pathnode] (pathB6) [right of=pathB5] {6};
+\node[pathnode] (pathB7) [right of=pathB6] {9};
+
+\node[pathnode,label=left:path3] (pathC1) [below of=pathB1] {1};
+\node[pathnode] (pathC2) [right of=pathC1] {2};
+\node[pathnode] (pathC3) [right of=pathC2] {3};
+\node[pathnode] (pathC4) [below of=pathB6] {6};
+\node[pathnode] (pathC5) [right of=pathC4] {7};
+
+\draw[very thick] (pathA1) -- (pathA2);
+\draw[very thick] (pathA2) -- (pathA3);
+\draw[very thick] (pathA3) -- (pathA4);
+\draw[very thick] (pathA4) -- (pathA5);
+\draw[very thick] (pathA5) -- (pathA6);
+\draw[very thick] (pathA6) -- (pathA7);
+
+\draw[very thick] (pathB1) -- (pathB2);
+\draw[very thick] (pathB2) -- (pathB3);
+\draw[very thick] (pathB3) -- (pathB4);
+\draw[very thick] (pathB4) -- (pathB5);
+\draw[very thick] (pathB5) -- (pathB6);
+\draw[very thick] (pathB6) -- (pathB7);
+
+\draw[very thick] (pathC1) -- (pathC2);
+\draw[very thick] (pathC2) -- (pathC3);
+\draw[very thick] (pathC3) -- (pathC4);
+\draw[very thick] (pathC4) -- (pathC5);
+
 %\end{pgfonlayer}
@@
 \end{tikzpicture}
+
 \end{document}

### 2022-01-08 manuscript/fig_induction.tex
figure 1 now with colored paths

@@
 %Lines
-\draw[very thick,color=red] (seqA1) -- (seqA2);
-\draw[very thick,color=red] (seqA2) -- (seqA3);
-\draw[very thick,color=red] (seqA3) -- (seqA4);
-\draw[very thick,color=red] (seqA4) -- (seqA5);
-\draw[very thick,color=red] (seqA5) -- (seqA6);
-\draw[very thick,color=red] (seqA6) -- (seqA7);
+\draw[line width=2.0,color=red] (seqA1) -- (seqA2);
+\draw[line width=2.0,color=red] (seqA2) -- (seqA3);
+\draw[line width=2.0,color=red] (seqA3) -- (seqA4);
+\draw[line width=2.0,color=red] (seqA4) -- (seqA5);
+\draw[line width=2.0,color=red] (seqA5) -- (seqA6);
+\draw[line width=2.0,color=red] (seqA6) -- (seqA7);
 
-\draw[very thick,color=green] (seqB1) -- (seqB2);
-\draw[very thick,color=green] (seqB2) -- (seqB3);
-\draw[very thick,color=green] (seqB3) -- (seqB4);
-\draw[very thick,color=green] (seqB4) -- (seqB5);
-\draw[very thick,color=green] (seqB5) -- (seqB6);
-\draw[very thick,color=green] (seqB6) -- (seqB7);
+\draw[line width=2.0,color=green] (seqB1) -- (seqB2);
+\draw[line width=2.0,color=green] (seqB2) -- (seqB3);
+\draw[line width=2.0,color=green] (seqB3) -- (seqB4);
+\draw[line width=2.0,color=green] (seqB4) -- (seqB5);
+\draw[line width=2.0,color=green] (seqB5) -- (seqB6);
+\draw[line width=2.0,color=green] (seqB6) -- (seqB7);
 
-\draw[very thick,color=blue] (seqC1) -- (seqC2);
-\draw[very thick,color=blue] (seqC2) -- (seqC3);
-\draw[very thick,color=blue] (seqC3) -- (seqC4);
-\draw[very thick,color=blue] (seqC4) -- (seqC5);
+\draw[line width=2.0,color=blue] (seqC1) -- (seqC2);
+\draw[line width=2.0,color=blue] (seqC2) -- (seqC3);
+\draw[line width=2.0,color=blue] (seqC3) -- (seqC4);
+\draw[line width=2.0,color=blue] (seqC4) -- (seqC5);
 
@@
 
+\begin{pgfonlayer}{bg}    % select the background layer
+\draw[line width=2,transform canvas={yshift=0.05cm},color=red] (g1) to (g2);
+\draw[line width=2,transform canvas={yshift=-0.05cm},color=blue] (g1) to (g2);
+\draw[line width=2,transform canvas={yshift=0cm},color=green] (g1) to (g3);
+\draw[line width=2,transform canvas={yshift=0cm},color=green] (g3) to (g4);
+\draw[line width=2,transform canvas={yshift=0.05cm},color=red] (g2) to (g4);
+\draw[line width=2,transform canvas={yshift=-0.05cm},color=blue] (g2) to (g4);
+\draw[line width=2,transform canvas={yshift=0.05cm},color=red] (g4) to (g5);
+\draw[line width=2,transform canvas={yshift=-0.1cm},color=blue,looseness=1] (g4) to (g7);
+\draw[line width=2,transform canvas={yshift=-0.05cm},color=green] (g4) to (g5);
+\draw[line width=2,transform canvas={yshift=0.05cm},color=red] (g5) to (g6);
+\draw[line width=2,transform canvas={yshift=-0.05cm},color=green] (g5) to (g6);
+\draw[line width=2,transform canvas={yshift=0.05cm},color=red] (g6) to (g7);
+\draw[line width=2,transform canvas={yshift=-0.05cm},color=green] (g6) to (g7);
+\draw[line width=2,transform canvas={yshift=0.05cm},color=red] (g7) to (g8);
+\draw[line width=2,transform canvas={yshift=-0.05cm},color=blue] (g7) to (g8);
+\draw[line width=2,transform canvas={yshift=0cm},color=green] (g7) to (g9);
+
+\end{pgfonlayer}
+
+\begin{comment}
 \draw[very thick] (g1) to (g2);
@@
 \draw[very thick] (g7) to (g9);
+\end{comment}
+
+%\coordinate[yshift=-0.6cm, right=0.1cm of A1.east] (aux2);
 
@@
 
+\begin{comment}
 \node[pathnode,label=left:path1] (pathA1) [below = 1cm of g1] {1};
@@
 \draw[very thick] (pathC4) -- (pathC5);
+\end{comment}
 

### 2022-01-09 manuscript/fig_from_alignment_to_variation_graph.tex
bugfix
figure1

@@
    A visual description of variation graph induction.
-   At top, the alignment graph model.
+   \textit{Top}: an alignment graph model for three sequences and their alignments.
+   Nodes are single characters (DNA base pairs, of which the forward strand is shown) in individual sequences.
+   Solid ``precedence'' edges link successive characters in each sequence, and are colored (red, green, blue) to identify each.
+   Dashed edges indicate aligned pairs of characters.
+    \textit{Bottom}: a variation graph model induced from the alignment graph.
+   %   We transform the alignment graph into a variation graph  first condensing sets of nodes in the alignment graph that are transitively linked by alignment edges.
+   Each transitive match closure (gray shaded edges, of increasing darkness) in the alignment graph results in a single node in the output graph, which is labeled by the rank of the transitive closure operation that produced it.
+   By recording the full set of match closures, we can project the sequences in the input through to paths in the variation graph (colored edges).
+   The unique set of node pairings in the paths provide the edges of the output graph.
+   %   Black nodes are numbered by the rank of the transitive closure operation
    %   At bottom, a sequence graph obtained by contracting 
-
    %   At top, three sequences (text1, text2, text3) 

### 2022-01-09 manuscript/main.tex
stuff
drop xcolor
get the first figure in
tweaks

@@
 \usepackage{floatrow}
+
+\usepackage{comment}
 
@@
 We add the character of the match $c(m_1^+)$ as a node $n_1$ in $V$, record the mapping from $m_1^+ \to n_1$.
-To induce the graph, we take the next un-used match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $V$ while recording the match transitive closure to graph mapping.
+To induce the graph, we take the next un-used match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $V$.
+To allow the annotation of paths, we record the set of characters in $Q$ that match to a given node in $V$ in mapping $Z = Q \to V$
 We continue until all matches have been used.
-Finally, we establish paths ($P$) by projecting $Q$ through the match-to-node mapping, and record edges ($E$) where nodes occur successively in paths.
+Finally, we establish paths ($P$) by walking them in $V$ using $Z$, and record edges ($E$) where nodes occur successively in paths.
 
@@
 Each subsequent step includes progressively more of $A$, until at completion, all pairwise relationships are accounted for in $V$.
-The variation graph is constructed from pairwise alignments, and because its paths are equivalent to input sequences, it represents them losslessly.
+The paths of the variation graph are equivalent to input sequences, and the path relationship within the graph is defined by their alignments.
 \end{proof}
 
+The set of alignments represented by a variation graph is strictly larger than the set of alignments used to induce it.
+The graph must contain at least the set of alignments given on input.
+It may also contain new implied pairwise relationships that arise due to transitive match relationships, as shown in closures 1 and 6 in Figure \ref{fig:induction}.
+However, it cannot contain less match information than represented in $A$, by definition of $V$.
+
+\input{fig_from_alignment_to_variation_graph.tex}
+
+\section{Algorithm}
+
+\section{Implementation}
+
 \begin{comment}
-
-Our algorithm flows through three complementary graph models.
-(\textit{1}) We transform the alignments and input sequences into an \textit{alignment graph} \citep{Kehr_2014}.
-(\textit{2}) We then transitively close match relations in this graph to produce single nodes of a \textit{sequence graph} \citep{Hein_1989}.
-(\textit{3}) By recording the translation from alignment graph nodes to sequence graph nodes, we are able to walk the input sequences through the graph.
-In doing so, we are able to derive edges and label the paths of a \textit{variation graph} \citep{Garrison_2018} which captures the pairwise alignments and all implied transitive relationships between them.
 
@@
 We encode the $X$ matches in a set of alignments made by an external method \cite{wfmash,Li_2018} in $A = m_1 \ldots m_X$.
-
 \citep{Li_bedtk_2020}.
 
-\FIXME{To integrate to the text} The alignment graph formalizes the collection of alignments between genomes.
+%\FIXME{To integrate to the text
+The alignment graph formalizes the collection of alignments between genomes.
 In this graph, each base pair in the input genomes is represented by a node.
 Two kinds of edges connect these nodes.
-The first `precedence' edge type indicates successive pairs of base pairs in the input,
-linking together the input genomes within the alignment graph.
+The first ``precedence'' edge type indicates successive pairs of base pairs in the input, linking together the input genomes within the alignment graph.
 The second “match” type of edge links nodes that alignment has asserted to be matches.
@@
 
-\end{comment}
-
-    \input{fig_from_alignment_to_variation_graph.tex}
-
-    \section{Implementation}
+\section{Seqwish implementation}
 \label{sec:implementation}
 
-    \FIXME{The algorithm validates some aspects of itself, we should point that out}
-
-    To fully induce the variation graph we need to establish the links between bases in $S$ that would be required for us to find any sequence in the input as a walk through the graph.
+We need to establish the links between bases in $S$ that would be required for us to find any sequence in the input as a walk through the graph.
 We do so by rewriting $Q$ (in both the forward and reverse orientation) in terms of pairs of bases in $S$, then sorting the resulting pairs by their first element, which yields $L = [(b_{a}, b_{b}), \ldots ]$.
-    These pairs record the links and their frequencies, which we can emit or filter (such as by frequency) as needed in particular applications.
-    In typical use we take the graph to be given by the unique elements of $L$.
+In typical use we take the edges of the graph to be given by the unique elements of $L$.
 
@@
 
-    \FIXME{To integrate to the text}
-    To support this kind of data scale, we implement \texttt{seqwish}'s algorithm using disk-backed memory.
-    We write the alignment graph into a disk-backed implicit interval tree, wherein each match in the input alignments is a queryable range.
-    A lock-free, parallel union-find algorithm \FIXME{Wait-free parallel algorithms for the union-find problem} allows us to, for each base pair of input sequence, collect all other base pairs in the pangenome that are aligned to it.
-    By processing this transitive closure operation in chunks (\textit{e.g.} 10Mbp at a time), we limit the maximum working memory (RAM) requirements of the process without affecting the output.
-    After the transitive closure, we write our final graph on disk in a set of arrays, then augment it with the paths of our input sequences,
-    finally emitting the resulting graph in the standard Graphical Fragment Assembly (GFA) format~\citep{GFA}.
+\end{comment}
+
+%    \FIXME{To integrate to the text}
+%    To support this kind of data scale, we implement \texttt{seqwish}'s algorithm using disk-backed memory.
+%    We write the alignment graph into a disk-backed implicit interval tree, wherein each match in the input alignments is a queryable range.
+%    A lock-free, parallel union-find algorithm \FIXME{Wait-free parallel algorithms for the union-find problem} allows us to, for each base pair of input sequence, collect all other base pairs in the pangenome that are aligned to it.
+%    By processing this transitive closure operation in chunks (\textit{e.g.} 10Mbp at a time), we limit the maximum working memory (RAM) requirements of the process without affecting the output.
+%    After the transitive closure, we write our final graph on disk in a set of arrays, then augment it with the paths of our input sequences,
+%    finally emitting the resulting graph in the standard Graphical Fragment Assembly (GFA) format~\citep{GFA}.
 
@@
 
-    \subsection{subsection 1}
-    \label{subsec:subsec1}
-    \FIXME{to do}
+    %    \subsection{subsection 1}
+    %\label{subsec:subsec1}
+    %\FIXME{to do}
 

### 2022-01-12 scripts/seqwish_with_logging.sh
add script for logged seqwish

@@ -1,0 +1,20 @@
+#!/bin/bash
+
+ASSEMBLIES=$1
+PAF=$2
+GFA=$3
+k=$4
+B=$5
+LOG=$6
+SECONDS=$7
+
+hostname
+
+cd /scratch
+
+(while true; do (date +%s; du -s .; ls -s; echo) >>"$LOG"; sleep "$SECONDS"; done) &
+SIZE_DEMON_PID=$!
+
+\time -v ~/tools/seqwish/bin/seqwish-ccfefb016fcfc9937817ce61dc06bbcf382be75e -t 48 -s "$ASSEMBLIES" -p "$PAF" -g "$GFA" -k "$k" -B "$B" -P
+
+kill $SIZE_DEMON_PID

### 2022-01-12 manuscript/main.tex

@@
 An identical set of path steps is thus equivalent to a sequence match.
-Pairwise alignments are by definition collections of matches between sequences.
+Pairwise alignments are by definition collections of character-level matches between sequences.
 The variation graph thus models a set of pairwise alignments between paths in $P$.
@@
 Each subsequent step includes progressively more of $A$, until at completion, all pairwise relationships are accounted for in $V$.
-The paths of the variation graph are equivalent to input sequences, and the path relationship within the graph is defined by their alignments.
+%All characters in $S$ have been accounted for in $V$, and we can make this mapping explicit by tracing $S$ through the
+%The paths of the variation graph are equivalent to input sequences, and the path relationship within the graph is defined by their alignments.
 \end{proof}

### 2022-01-16 manuscript/main.tex
get bounded

@@
 \usepackage{comment}
+
+\usepackage{siunitx}
 
@@
 
-\section{Variation graph induction}
+\section{Algorithm}
 \label{sec:algorithm}
@@
 
+In this section, we provide a formal definition of variation graph induction.
+We then examine the bounds of a na\"{i}ve implementation of this algorithm.
+Finally, we propose compression and partitioning techniques to reduce the space and working memory complexity of the induction process by a large constant factor modulated by the degree of sequence divengence in the input pangenome.
+This yields a practical algorithm for variation graph induction that can scale to the largest available pangenomes.
+
+\subsection{Variation graph induction}
+
 \begin{definition}
@@
 Variation graphs are a common formalism to encode pangenome graphs \citep{Garrison_2019_thesis}.
-In the variation graph, $V = (N, E, P)$, nodes $N = n_1\ldots n_{|N|}$ contain sequences of DNA.
+In the variation graph, $\mathcal{V} = (\mathcal{N}, \mathcal{E}, \mathcal{P})$, nodes $\mathcal{N} = n_1\ldots n_{|\mathcal{N}|}$ contain sequences of DNA.
 Each node $n_i$ has a unique numerical identifier $i$ and an implicit reverse complement $\bar{n_i}$.
 A node strand $s$ corresponds to one node orientation.
-Edges $E = e_1\ldots e_{|E|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
-Paths $P = p_1\ldots p_{|P|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written in terms of the sequence graph.
+Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
+Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written in terms of the sequence graph.
 \end{definition}
@@
 Pairwise alignments are by definition collections of character-level matches between sequences.
-The variation graph thus models a set of pairwise alignments between paths in $P$.
+The variation graph thus models a set of pairwise alignments between paths in $\mathcal{P}$.
 \end{proof}
@@
 \begin{theorem}
-We can construct a variation graph from pairwise alignments.
+\label{thm:seqwish}
+%\textit{seqwish}:
+We can build a variation graph from sequences and pairwise alignments.
+The variation graph will fully embed both the sequences and all pairwise relationships in the input.
 \end{theorem}
@@
 
-We build a graph $V$ inductively.
+We build a graph $\mathcal{V}$ inductively.
 We take the first match in $A$, $m_1$, and execute a union-find operation to obtain $m_1^+$.
-We add the character of the match $c(m_1^+)$ as a node $n_1$ in $V$, record the mapping from $m_1^+ \to n_1$.
-To induce the graph, we take the next un-used match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $V$.
-To allow the annotation of paths, we record the set of characters in $Q$ that match to a given node in $V$ in mapping $Z = Q \to V$
+We add the character of the match $c(m_1^+)$ as a node $n_1$ in $\mathcal{V}$, record the mapping from $m_1^+ \to n_1$.
+To induce the graph, we take the next un-used match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $\mathcal{V}$.
+To allow the annotation of paths, we record the set of characters in $Q$ that match to a given node in $\mathcal{V}$ in mapping $Z = Q \to \mathcal{N} = m_1 \ldots m_{|\mathcal{N}|}$.
 We continue until all matches have been used.
-Finally, we establish paths ($P$) by walking them in $V$ using $Z$, and record edges ($E$) where nodes occur successively in paths.
+Finally, we establish paths ($P$) by walking them in $\mathcal{V}$ using $Z$, and record edges ($E$) where nodes occur successively in paths.
 
@@
 It may also contain new implied pairwise relationships that arise due to transitive match relationships, as shown in closures 1 and 6 in Figure \ref{fig:induction}.
-However, it cannot contain less match information than represented in $A$, by definition of $V$.
+However, by definition of $V$, it cannot contain \textit{less} match information than represented in $A$.
 
@@
 
-\section{Algorithm}
+\subsection{Induction algorithm sketch}
+
+%Although the space bounds alone indicate that this approach is infeasible, we can also consider its time complexity.
+%First, we must enable random access to match elements in $A$.
+%Assuming that $A$ contains pairs of characters in $Q$, then we can sort it
+For the sake of time and space complexity analysis, we consider a simple algorithm to implement the induction process.
+The induction depends on our ability to compute transitive closures of matches $m^+$.
+If $A$ is sorted, we can find the matches of a given character in $Q$ using binary search, which allows us to compute $m^+$ for each character.
+We do so non-redundantly by marking each used charcater in $Q$ in an auxiliary data structure (e.g. a bitvector of $|Q|$).
+As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{N}$ and the the sequence-to-graph mapping $Z$, which like $A$ consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{N}$.
+As with $A$, we can sort $Z$ to obtain random access via binary search.
+Finally, we derive elements in $\mathcal{P}$ by iterating through the characters of $Q$ and looking up their mapping in $Z$ using binary search.
+The edge set $\mathcal{E}$ are the unique pairs of steps found in $\mathcal{P}$, and can be computed by sorting pairs of steps in $\mathcal{P}$.
+
+%We construct an index $A_{idx}$ which allows us to query the matches of each character in $Q$.
+
+\subsection{Na\"{i}ve algorithm bounds}
+\label{sec:bounds}
+
+The inductive proof of theorem \ref{thm:seqwish} demonstrates how to build a variation graph from sequences and their pairwise alignments.
+However, a na\"{i}ve algorithm based on this model will require a very large amount of space.
+Take $\mathcal{Q} = |S|$.
+Although our identifier space $\mathcal{Q}$ must include all of $Q$, in practice, we only store $S$, as $\bar{S}$ can be trivially computed.
+Assume an all-to-all alignment of $N$ sequences in $A$ as input, and that all sequences are approximately identical, so that the induced variation graph has $\mathcal{Q}/N$ nodes.
+The induction must maintain reference to all characters in all input sequences $O(\mathcal{Q})$, all character-to-character matches $O(|A|) \approx O(\mathcal{Q}N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(\mathcal{Q})$, the nodes of the graph $O(|\mathcal{N}|) \approx O(\mathcal{Q}/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(\mathcal{Q})$, and the set of paths $O(|\mathcal{P}|) \approx O(\mathcal{Q})$.
+In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(\mathcal{Q}N^2 + 4\mathcal{Q} + \mathcal{Q}/N)$ space.
+
+Assuming that we want to build a graph for e.g.
+100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$ we might expect to use $\approx O(\num{e15})$ identifiers to store the full model.
+Such a design is almost certainly infeasible for inputs larger than a handful of genomes.
+For instance, we would need $O(\num{3e12})$ identifiers for just 5 human genomes.
+Although it is feasible to compute such a graph using external memory, the approximately 200-fold increase in space relative to the input renders this clearly impractical.
+
+Considering the time complexity of induction, we anticipate $O(|A|\log|A|)$ time to sort the match set and $O(\mathcal{N} \log |A|)$ to query it and compute our $\mathcal{N}$ transitive closures.
+Computing the variation graph paths $\mathcal{P}$ involves converting sequences in $S$ to walks through $\mathcal{N}$.
+We first sort the sequence-to-graph mapping array $Z$ in $O(\mathcal{Q} \log \mathcal{Q})$ operations, and then compute $\mathcal{P}$ in $O(\mathcal{Q})$ queries which each cost $O(\log \mathcal{Q})$.
+To obtain unique edges and generate $\mathcal{E}$, we must build and sort an array of $O(2\mathcal{P})$, and then iterate through it for $O(2\mathcal{P} \log 2\mathcal{P} + 2\mathcal{P})$ operations.
+In sum, we would expect to require $O(|A|\log|A| + \mathcal{N} \log |A| + 2(\mathcal{Q} \log \mathcal{Q}) + 2(\mathcal{P} \log 2\mathcal{P} + \mathcal{P}))$.
+Using our approximate relationships to $\mathcal{Q}$ given previously, and simplifying we arrive at $O[\mathcal{Q}(2N^2 \log N / \log \mathcal{Q} + (1/N + 4)\log \mathcal{Q} + 2 + log(4))]$.
+
+Due to our dependence on sorting, and the logarithmic-time cost of queries, crowth in $\mathcal{Q}$ drives $\Omega(\mathcal{Q} \log \mathcal{Q})$ growth in overall complexity.
+As $N$ grows, the time complexity is dominated by the number of alignments, which in the case of our example is $O(N^2)$.
+Clearly, for large numbers of highly-similar genomes, we may not require all pairs.
+Various approaches could be used to reduce the size of $A$, but these lie outside the scope of this work.
+
+%In practice, it is likely that this can be considerably reduced with minimal effect , particularly for
+%In practice, for large numbers of highly similar genomes, it a true all-pairs alignment may be unnecessary, but the construction of a graph that represents all-pairwise alignments 
+%In practical applications, it is unlikely that we will need a true all-pairs alignments for highly similar genomes, and this could be considerably reduced with heuristics that
+
+%we likely  a true all pairs alignment, but limiting this requires heuristic schemes to generate a set of alignments that does not cause any match transitive closures to be partitioned.
+This topic lies outside the scope of this work.
+
+%$O[(\mathcal{Q}/N + 5\mathcal{Q}) \log \mathcal{Q} + \mathcal{Q}(2 + log(4))]$.
+%$O(\mathcal{Q} \log \mathcal{Q} + \mathcal{Q}/N \log \mathcal{Q} + 2(\mathcal{Q} \log \mathcal{Q}) + 2(\mathcal{Q} \log 2\mathcal{Q} + \mathcal{Q}))$.
+
+%(note that this would 
+%To enable random access to match elements between $A$
+%Continuing our assumed graph context, with quasi-identical genomes and an all-to-all alignment between them, we can compute each match closure operation 
+%The cost of each transitive closure operation should be
+
+\subsection{Match compression}
+
+As the analysis in section \ref{sec:bounds} shows, space requirements make it impractical to apply a trivial version of theorem \ref{thm:seqwish} to generate a large pangenome graph.
+We seek a compression approach that will exploit redundancy in the input genomes to reduce the costs of the algorithm.
+With large numbers of genomes, alignments dominate our costs.
+A simple technique is to generalize matches $m = (i, j)$ to pairs of ranges in $Q$.
+For highly-similar sequences, we expect exact matches to occur in long runs.
+If the average pairwise diversity of sequences in our input is $1/k$, we expect exact matches to be around $k$ characters long.
+By encoding matches as pairs of ranges of characters, $r = (a, b) : a, b = (i, j) : i, j \in Q$, we can obtain a $\approx$ $k$-fold compression of $A$.
+The same encoding can be used to replace the sequence-to-graph mapping $Z$.
+Rather than pairs of characters in $Q$ and $\mathcal{N}$, we can record runs of matchhes between them as range matches.
+
+%We can save $\approx k$ match pairs per recorded range 
+%a pair of ranges $r = [(a_0, a_1), (b_0, b_1)]$ in $Q$ requires twice the space, but we expect it to
+
+%The dominance of alignments in our space costs make
+%In practice, the cost of
+
+%We render this
+
+\subsection{Induction partitioning}
+
+\subsection{Node compaction}
+
+%Taking $\mathcal{Q} = |Q|$, and assuming an all-to-all alignment of $N$ sequences in $A$ would give an approximate space bound of $O(2\mathcal{Q} + \mathcal{Q}N^2)$
+%in the order of the length of sequences, the number of character matches in the alignments, and
 
@@
 
-Using a compressed suffix array (CSA) over sequence names, we build an index that maps between offsets in all genomes and offsets in $Q$ via lookup of the sequence name in the CSA.
+Using a compressed suffix array (CSA) \citep{Sadakane_2000} over sequence names, we build an index that maps between offsets in all genomes and offsets in $Q$ via lookup of the sequence name in the CSA.
 This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.
@@
 
-    We thank Richard Durbin, Benedict Paten, and Simon Puglisi for support and helpful discussions in the development of this algorithm.
-    %We are greatful to members of the HPRC production teams for their development of resources used in our exposition.
+    %We thank Richard Durbin, Benedict Paten, and Simon Puglisi for support and helpful discussions in the development of this algorithm.
+    We are greatful to members of the HPRC production teams for their development of resources used in our exposition.
 

### 2022-01-18 manuscript/main.tex
implementation

@@
 We design the \textit{seqwish} algorithm, which builds variation graph from a set of sequences and alignments between them.
-We first transform the alignment set into an interval tree.
+We first transform the alignment set into an implicit interval tree.
 To build up the variation graph, we query this tree-based representation of the alignments to reduce transitive matches into single DNA segments in a sequence graph.
 By recording the mapping from input sequence to output graph, we can trace the original paths through this graph, yielding a pangenome graph.
-We present an implementation that operates in external memory, using disk-backed implicit interval trees and a lock-free parallel algorithm to drive the core graph induction step.
+We present an implementation that operates in external memory, using disk-backed data structures and a lock-free parallel algorithm to drive the core graph induction step.
 We demonstrate that our method scales to very large graph induction problems, and apply our work to build graphs from dozens of genomes from the same or different species in clades throughout the tree of life.
@@
 Conceptually, the pangenome graph relates genomes as walks through and underlying DNA sequence language encoded as a sequence graph. 
-In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to sequences that are very similar to a chosen reference genome \cite{Eizenga_2020}.
+In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to sequences that are very similar to a chosen reference genome \citep{Eizenga_2020}.
 
@@
 
-Although a reference-guided approach is pragmatic when we only have a single reference genome, in humans and other species recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies \cite{Nurk_2021}.
+Although a reference-guided approach is pragmatic when we only have a single reference genome, in humans and other species recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies \citep{Nurk_2021}.
 Motivated by the need to generate ``multi-reference'' pangenome graphs, we develop \textit{seqwish}, an algorithm that allows us to build pangenome graphs directly from a collection of sequences and pairwise alignments between them.
@@
 It may also contain new implied pairwise relationships that arise due to transitive match relationships, as shown in closures 1 and 6 in Figure \ref{fig:induction}.
-However, by definition of $V$, it cannot contain \textit{less} match information than represented in $A$.
+However, by definition of $\mathcal{V}$, it cannot contain \textit{less} match information than represented in $A$.
 
@@
 
-Due to our dependence on sorting, and the logarithmic-time cost of queries, crowth in $\mathcal{Q}$ drives $\Omega(\mathcal{Q} \log \mathcal{Q})$ growth in overall complexity.
-As $N$ grows, the time complexity is dominated by the number of alignments, which in the case of our example is $O(N^2)$.
-Clearly, for large numbers of highly-similar genomes, we may not require all pairs.
-Various approaches could be used to reduce the size of $A$, but these lie outside the scope of this work.
+Due to our dependence on sorting, and the logarithmic-time cost of queries, growth in $\mathcal{Q}$ drives $\Omega(\mathcal{Q} \log \mathcal{Q})$ growth in overall complexity.
+As $N$ grows, both time and space complexity are dominated by the number of alignments, which in the case of our example is $O(N^2)$.
+For large numbers of highly-similar genomes, we may not require all pairs of alignments to build a graph that contains all pairwise alignments.
+Various approaches could be used to reduce the size of $A$ without disrupting the induced graph.
+We leave these to later work.
 
@@
 \subsection{Match compression}
+\label{sec:matchcompression}
 
-As the analysis in section \ref{sec:bounds} shows, space requirements make it impractical to apply a trivial version of theorem \ref{thm:seqwish} to generate a large pangenome graph.
-We seek a compression approach that will exploit redundancy in the input genomes to reduce the costs of the algorithm.
+As the bounds analysis shows, space requirements make it impractical to apply a trivial version of theorem \ref{thm:seqwish} to generate a large pangenome graph.
+We need a compression approach that will exploit redundancy in the input genomes to reduce the costs of the algorithm.
 With large numbers of genomes, alignments dominate our costs.
@@
 If the average pairwise diversity of sequences in our input is $1/k$, we expect exact matches to be around $k$ characters long.
-By encoding matches as pairs of ranges of characters, $r = (a, b) : a, b = (i, j) : i, j \in Q$, we can obtain a $\approx$ $k$-fold compression of $A$.
-The same encoding can be used to replace the sequence-to-graph mapping $Z$.
-Rather than pairs of characters in $Q$ and $\mathcal{N}$, we can record runs of matchhes between them as range matches.
+By encoding matches as pairs of ranges of characters, $r = (a, b) : a, b = (i, j) : i, j \in Q$, we can obtain a $\approx$ $k$-fold compression of $A$, which for clarity we label as $\mathcal{A}$.
+If sorted, $\mathcal{A}$ can be treated as an \textit{implicit interval tree} \citep{Li_bedtk_2020}, which allows queries of containment and overlap in $O(|\mathcal{A}| \log |\mathcal{A}|)$ time.
+This compression requires trivial changes to our graph induction model.
+To obtain our match transitive closures ($m^+$), we query $\mathcal{A}$ for the range of a single character in $Q$, computing the character-level transitive relationships from the relative offsets of the ranges in $\mathcal{A}$.
+Match compression thus reduces our alignment storage memory bounds by a factor of $k$ without affecting our time complexity bounds.
+
+The same encoding can be used to replace the sequence-to-graph mapping $Z$, yielding $\mathcal{Z}$.
+Rather than pairs of characters in $Q$ and $\mathcal{N}$, we record runs of matches between them as range matches.
+Although in expectation the length of these matches should be strictly less than $k$, due to the interruption of the graph by variation between genomes, this still allows us to reduce the size of $Z$ using runs of matches between $Q$ and $\mathcal{N}$.
+Additionally, we store the inverse of $\mathcal{Z}$, which maps ranges from $\mathcal{N} \to Q$, as $\bar{\mathcal{Z}}$.
+We use $\mathcal{Z}$ to compact unbranching regions of $\mathcal{V}$ into single nodes, and $\bar{\mathcal{Z}}$ to accelerate our calculation of links in the graph.
 
@@
 
+\subsection{Node compaction}
+
+For simplicity, we have thus-far presented a character-level model of variation graph induction.
+However, range, or ``run'' compression can also reduce the representation size of the graph.
+Rather than recording an identifier for each character in a sequence graph, it is useful to ``compact'' characters that form trivial linear components in the graph into single nodes.
+Broadly, the size of nodes will be bounded by the average distance between variants, which for pangenomes built from ~100 individuals of the same species, often provides a great reduction in the total number of nodes (and thus identifiers) required for $\mathcal{V}$ and its components.
+
+To compact $\mathcal{V}$, we traverse $Q$, finding each entry in $\mathcal{Z}$ in turn, recording its start and end in $\mathcal{N}$, which can be understood as a character vector or string containing all the sequence in the nodes of $\mathcal{V}$.
+%in a ``graph vector'' $\mathcal{G} = n_1,n_2 \ldots n_{|\mathcal{N}|}$ containing the ordered characters of the graph.
+%Any position in $\mathcal{G}$ so marked must be the start or end of a node.
+We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ such that $|n_i| \ge 1 \forall_{i \in 1 \ldots |\mathcal{N}'|}$.
+
 \subsection{Induction partitioning}
 
-\subsection{Node compaction}
-
-%Taking $\mathcal{Q} = |Q|$, and assuming an all-to-all alignment of $N$ sequences in $A$ would give an approximate space bound of $O(2\mathcal{Q} + \mathcal{Q}N^2)$
-%in the order of the length of sequences, the number of character matches in the alignments, and
+Although match compression provides an approximate factor $k$ improvement in memory bounds for key data structures used in the induction, the approach we present in section \ref{sec:matchcompression} requires working memory in the order of the set of transitive match closures in our graph.
+A simple approach to reduce this bound is to divide the induction problem into smaller pieces.
+We do so by computing the graph induction for a collection of initial characters in $Q$.
+We induce the related graph for each partition, appending the induction results to the appropriate data structures.
+This partitioning can introduce boundary effects which change the contents of $\mathcal{Z}$ and $\bar{\mathcal{Z}}$) by splitting ranges at the boundaries of our partitions.
+However, while this will affect the compressed node definition $\mathcal{N}'$, it does not affect $\mathcal{N}$, and it can be corrected via a post-processing step to sort and compact the id space.
+Partitioning reduces the working memory limits to be in the order of the size of the largest transitive match closure found in the input.
 
 \section{Implementation}
+
+We have presented a complete model for variation graph induction from sequences and their pairwise alignments.
+Here, we describe our specific implementation of this algorithm: \textit{seqwish}.
+In general, our approach uses external memory to elaborate the graph, and we assume the availability of low-latency storage media (like SSDs) to maximize the performance of this approach.
+
+\subsection{Key disk-backed data structures}
+
+In our implementation, we rely on several basic external memory kernels.
+To reduce working memory requirements to an absolute minimum, we use a disk-backed version of the implicit interval tree that memory-maps the sorted array of intervals (mmiitree).
+Indexing the implicit interval tree requires a sorting step which dominates the runtime of our algorithm.
+We adapt the current best-performing in-place parallel sorting algorithm, In-place Parallel Super Scalar Samplesort (IPS$^4$o), to work on a disk-backed, memory-mapped array \citep{axtmann2017}.
+This allows us work with $\mathcal{A}$, $\mathcal{Z}$, and $\bar{\mathcal{Z}}$ in external memory.
+By storing numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$.
+The graph sequence vector $\mathcal{N}$ is simply written by appending characters to a file.
+We ``mark'' nodes to generate $\mathcal{N}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \cite{Gog_2014} for support of the final emission of the graph $\mathcal{V}$.
+
+\subsection{Input and output processing}
+
+A practical implementation introduces a new set of problems related to input and output. % that we have not yet introduced.
+Our implementation reads standard data formats, FASTA or FASTQ for the input sequences, and PAF \citep{Li_2018} for pairwise alignments.
+It writes the graph in standard Graphical Fragment Assembly (GFA) format \citep{GFA}.
+
+In PAF, the input set of alignments is not directly expressed in terms of matches between specific characters in $Q$.
+Rather, each record lists the name of the aligned pair of query and target sequences and offsets in each.
+To efficiently process the input PAF, we thus need to build a sequence index that allows us to generate $\mathcal{A}$.
+We often wish to build graphs from very large collections of sequences, such as raw sequencing reads or contigs from many thousands of samples.
+We build a compressed suffix array (CSA) \citep{Sadakane_2000} over sequence names, and provide auxiliary supporting data structures that allow us to map between our input and the abstract concatenation of all input sequences and their reverse complements: $Q$.
+This \textit{seqidx} avoids the overheads associated with a hash table on string names of input sequences.
+To enable highly-efficient random access, we cache the input sequences in a disk-backed version of $Q$, into which our queries of sequence name and offset point.
+This trades time that might be spent accessing a compressed representation of the input for space in external memory.
+
+For output in GFA, the process is relatively simple.
+We iterate over nodes in $\mathcal{N}'$, writing each as a node record.
+Edges can be similarly produced from the disk-backed multiset representing $\mathcal{E}$.
+The most involved part of graph emission is the rendering of our input sequences $S$ as paths $\mathcal{P}$ through the graph.
+For each input sequence in our \textit{seqidx}, we walk through the offsets in $S$ contained in the sequence and look up their mapping into $\mathcal{N}'$ using $\mathcal{Z}$.
+Range compression allows us to complete one lookup per range.
+By definition, each character is $S$ is covered by only one range in $\mathcal{Z}$.
+We can thus iterate through the ranges in $\mathcal{Z}$ without considering each character.
+Following the GFA format, we are able to independently generate $\mathcal{P}$, as each path is represented on a separate record in the GFA.
 
@@
 Alignments between genomes imply a set of exact matches between subsequences of $Q$, which we represent as interval pairs $m := ((q_b, q_e), (t_b, t_e))$.
-We encode the $X$ matches in a set of alignments made by an external method \cite{wfmash,Li_2018} in $A = m_1 \ldots m_X$.
+We encode the $X$ matches in a set of alignments made by an external method \citep{wfmash,Li_2018} in $A = m_1 \ldots m_X$.
 \citep{Li_bedtk_2020}.

### 2022-01-18 manuscript/main.tex
remove commented-out old text

@@
 
-\begin{comment}
-
-\subsection{Sequence input}
-
-Using a compressed suffix array (CSA) \citep{Sadakane_2000} over sequence names, we build an index that maps between offsets in all genomes and offsets in $Q$ via lookup of the sequence name in the CSA.
-This allows us to map between positions in the sequences of $Q$, which is the format in which alignment algorithms typically express alignments, and positions in $Q$ itself, which is the coordinate space we will use as a basis for the generation of our graph.
-
-\subsection{Alignment graph}
-
-Alignments between genomes imply a set of exact matches between subsequences of $Q$, which we represent as interval pairs $m := ((q_b, q_e), (t_b, t_e))$.
-We encode the $X$ matches in a set of alignments made by an external method \citep{wfmash,Li_2018} in $A = m_1 \ldots m_X$.
-\citep{Li_bedtk_2020}.
-
-%\FIXME{To integrate to the text
-The alignment graph formalizes the collection of alignments between genomes.
-In this graph, each base pair in the input genomes is represented by a node.
-Two kinds of edges connect these nodes.
-The first ``precedence'' edge type indicates successive pairs of base pairs in the input, linking together the input genomes within the alignment graph.
-The second “match” type of edge links nodes that alignment has asserted to be matches.
-
-We encode the set of input pairwise alignments between sequences in $Q$ as object $A$.
-Although these alignments tend to be represented using oriented interval pairs in $Q$, for simplicity and robustness to graph complexity, we describe $A$ as a set of pairs of bidirectional positions (sequence offsets and strands) $[1 \ldots |Q_1 \ldots Q_{|Q|}|]$ , such that $A = \{ (b_{q}, b_{r}), \ldots \}$.
-We sort $A$ by the first member ($b_{q}$) of each pair, ensuring that the entries in $A$ are ordered according to their order in $Q$.
-
-To query the induced graph we build a rank/select dictionary allowing efficient traversal of $A$, based on a bit vector $A_{bv}$ of the same length as $A$ such that we record a 1 at those positions which correspond to the first instance of a given $b_{q}$ and record a 0 in $A_{bv}$ otherwise.
-We record which $b_{q}$ we have processed in the bitvector $Q_{seen}$ which is of the same length as $Q$.
-This allows us to avoid a quadratic penalty in the order of the size of the transitive closures in $Q$ generated by pairs in $A$.
-
-Now we inductively derive the graph implied by the alignments.
-For each base $b_{q}$ in $Q$ not already marked in $Q_{seen}$, we find its transitive closure $c_{q} := \{b_{q}, b_{r_{1}}, \ldots \}$ by traversing aligned base pairs recorded in $A$.
-We write the character of the base $b_{q}$ to an entry $s_i$ in a vector $S$, then for each $b_{c}$ in $c_{q}$ we record a pair $(s_{i}, b_{c})$ into $N$ and its reverse, $(b_{c}, s_{i})$ into $P$.
-We mark $Q_{seen}$ for each base in each emitted cluster, so that we will not consider these bases in subsequent transitive closures.
-By sorting $N$ and $P$ by their first entries, we can build rank/select dictionaries on them akin to that we built on $A$ that allow random access by graph base (as given in $S$) or input base (as given in $Q$).
-
-\subsection{Sequence graph}
-
-The alignment graph may be converted into a basic sequence graph by condensing all sets of nodes that are transitively linked by match edges.
-
-\subsection{Variation graph}
-
-By recording the relationship between sequence graph and alignment graph nodes, we are able to trace the paths of pangenome sequences through the graph, producing our target: a pangenome variation graph that completely describes both the input assemblies and their alignments (Fig.~\ref{fig:from_alignment_to_variation_graph}).
-
-Seqwish implements a lossless conversion from pairwise alignments between sequences to a variation graph encoding the sequences and their alignments.
-As input, we typically take all-versus-all alignments, but the exact structure of the alignment set may be defined in an application specific way.
-Seqwish uses a series of disk-backed sorts and passes over the alignment and sequence inputs to allow the graph to be constructed in low memory relative to the size of the input sequence set.
-Memory usage during construction and traversal is limited by the use of sorted disk-backed arrays and succinct rank/select dictionaries to record a queryable version of the graph.
-
-\section{Seqwish implementation}
-\label{sec:implementation}
-
-We need to establish the links between bases in $S$ that would be required for us to find any sequence in the input as a walk through the graph.
-We do so by rewriting $Q$ (in both the forward and reverse orientation) in terms of pairs of bases in $S$, then sorting the resulting pairs by their first element, which yields $L = [(b_{a}, b_{b}), \ldots ]$.
-In typical use we take the edges of the graph to be given by the unique elements of $L$.
-
-Our data model encodes the graph using single-base nodes, but often downstream use requires identifying nodes and thus we benefit from compressing the unitigs of the graph into single nodes, which reduces memory used by identifiers in analysis.
-We can compress the node space of the graph by traversing $S$, and for each base querying the inbound links.
-Maintaining a bitvector $S_{id}$ of length equal to $S$ we mark each base at which we see any link other than one from or to the previous base on the forward or reverse strand, or at bases where we have no incoming links.
-By building a rank/select dictionary on $S_{id}$ we can assign a smaller set of node ids to the sequence space of the graph.
-
-Given the id space encoded by $S_{id}$ we can materialize the graph in a variety of interchange formats, or provide id-based interfaces to the indexed squish graph.
-To generate graphs in \texttt {vg} or GFA format, we want to decompose the graph into its nodes ($S$), edges ($L$) and paths ($P$).
-The nodes are given by $S$ and $S_{id}$, and similarly we project $L$ and $P$ through $S_{id}$ to obtain a compressed variation graph.
-
-\end{comment}
-
 %    \FIXME{To integrate to the text}

### 2022-01-18 manuscript/main.tex
minor

@@
 This allows us work with $\mathcal{A}$, $\mathcal{Z}$, and $\bar{\mathcal{Z}}$ in external memory.
-By storing numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$.
+By storing pairs of numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$ in terms of offsets in $\mathcal{N}$.
 The graph sequence vector $\mathcal{N}$ is simply written by appending characters to a file.

### 2022-01-18 manuscript/main.tex
Introduction tips

@@
 
-\copyrightyear{2021} \pubyear{XXXX}
+\copyrightyear{2022} \pubyear{XXXX}
 
@@
 \textbf{Motivation:}
-Pangenome graphs, often formally encoded as variation graphs, moodel the mutual alignment of collections of genomes.
-A set of pairwise alignments imply a variation graph, but there are no scalable methods to generate such a graph or lossless string graph from these alignments.
+Pangenome graphs, often formally encoded as variation graphs, model the mutual alignment of collections of genomes.
+A set of pairwise alignments implies a variation graph, but there are no scalable methods to generate such a graph or lossless string graph from these alignments.
 Existing approaches depend on a single reference or a specific ordering of genomes, and do not provide base-level information about the mapping of genomes into the pangenome.
-A scalable method to build pangenome graphs without such biases would be a key functional step in pangenome construction and manipulation pipelines. \\
+A scalable method to build pangenome graphs without such limitations would be a key functional step in pangenome construction and manipulation pipelines. \\
 \textbf{Results:}
-We design the \textit{seqwish} algorithm, which builds variation graph from a set of sequences and alignments between them.
+We design the \textit{seqwish} algorithm, which builds a variation graph from a set of sequences and alignments between them.
 We first transform the alignment set into an implicit interval tree.
@@
 Popular approaches model the pangenome alignment as a \textit{pangenome graph}.
-Conceptually, the pangenome graph relates genomes as walks through and underlying DNA sequence language encoded as a sequence graph. 
-In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to sequences that are very similar to a chosen reference genome \citep{Eizenga_2020}.
+\FIXME{Conceptually, the pangenome graph relates genomes as walks through and underlying DNA sequence language encoded as a sequence graph.}
+In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to only sequences that are very similar to a chosen reference genome \citep{Eizenga_2020}.
 
 Surprisingly, existing methods to build pangenome graphs from large genome assemblies are reference-based.
-Minigraph progressively builds a pangenome graph using by layering an ordered series of input genomes onto a specific reference \citep{Li:2020}.
-Cactus depends on a single guide tree to structure pairwise alignments between genomes \citep{Armstrong:2020}.
-Both kinds of methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, and guide-tree topology.
+Minigraph progressively builds a pangenome graph by layering an ordered series of input genomes onto a specific reference \citep{Li:2020}.
+Cactus depends on a single guide-tree to structure pairwise alignments between genomes \citep{Armstrong:2020}.
+Both methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, and guide-tree topology.
+Furthermore, Minigraph does not provide base-level information, limiting the analysis to only large structural variation.
+Although a reference-guided approach is pragmatic when we only have a single reference genome, recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies in humans and other species \citep{Nurk_2021}.
 
-Although a reference-guided approach is pragmatic when we only have a single reference genome, in humans and other species recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies \citep{Nurk_2021}.
-Motivated by the need to generate ``multi-reference'' pangenome graphs, we develop \textit{seqwish}, an algorithm that allows us to build pangenome graphs directly from a collection of sequences and pairwise alignments between them.
-Here, we formally describe the algorithm, evaluate its basic combinatoric bounds and costs, and present applications of the method to pangenome building problems in a variety of species.
+A class of methods to represent pangenomes involves the \textit{sequence graph} \citep{Hein_1989}, where identical segments of different genomes collapse into a single representative segment (often node) in the graph.
+In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+A \textit{bidirected} sequence graph represents both strands of DNA.
+On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{Garrison_2018}.
+Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+
+A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Eizenga_2020}.
+Motivated by the need to generate ``multi-reference'' pangenome graphs, we develop \textit{seqwish}, an algorithm that allows us to build pangenome graphs in the form of variation graphs directly from a collection of sequences and pairwise alignments between them.
+Here, we formally describe the algorithm, evaluate its basic combinatorial bounds and costs, and present applications of the method to pangenome building problems in a variety of species.
 %Our approach is both highly generic and scalable.
-We can apply our method to any collection of sequences and alignments.
-It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
 %Not only is our approach highly generic,but it also efficiently scales to both large and complex pangenomes.
@@
 \label{sec:algorithm}
-
-A class of methods to represent pangenomes involves the \textit{sequence graph} \citep{Hein_1989}, where identical segments of different genomes collapse into a single representative segment (often node) in the graph.
-In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
-A \textit{bidirected} sequence graph represents both strands of DNA.
-On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{Garrison_2018}.
-Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
-A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Eizenga_2020}.
 
@@
 \FIXME{to do}
+We can apply our method to any collection of sequences and alignments.
+It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
 

### 2022-01-19 manuscript/fig_from_alignment_to_variation_graph.tex
tweaks
other tips

@@
-\vspace{-1em}
-\begin{figure}[ht!]
+%\vspace{-1em}
+\begin{figure}
 %  \centering
@@
    Nodes are single characters (DNA base pairs, of which the forward strand is shown) in individual sequences.
-   Solid ``precedence'' edges link successive characters in each sequence, and are colored (red, green, blue) to identify each.
+   Solid edges link successive characters in each sequence, and are colored (red, green, blue) to identify each.
    Dashed edges indicate aligned pairs of characters.
@@
    The unique set of node pairings in the paths provide the edges of the output graph.
+   Closures 1 and 6 imply pairwise relationships between sequences text1 and text3, and text1 and text2, respectively, that are both absent in the input alignments.
    %   Black nodes are numbered by the rank of the transitive closure operation
@@
     }
+    \label{fig:induction}
     %{\includegraphics[width=\linewidth]{fig/OUP_First_SBk_Bot_8401}} %\includegraphics[width=5cm]{name}}
@@
 \end{figure}
-\vspace{-1em}
+%\vspace{-1em}

### 2022-01-19 manuscript/main.tex
other tips

@@
 A \textit{bidirected} sequence graph represents both strands of DNA.
-On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{Garrison_2018}.
+On this model, variation graphs add the concept of \textit{path} to embed linear sequences (\textit{e.g.}, genomes or haplotypes) into the graph \citep{Garrison_2018}.
 Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
@@
 Variation graphs are a common formalism to encode pangenome graphs \citep{Garrison_2019_thesis}.
-In the variation graph, $\mathcal{V} = (\mathcal{N}, \mathcal{E}, \mathcal{P})$, nodes $\mathcal{N} = n_1\ldots n_{|\mathcal{N}|}$ contain sequences of DNA.
-Each node $n_i$ has a unique numerical identifier $i$ and an implicit reverse complement $\bar{n_i}$.
+In the variation graph $\mathcal{V} = (\mathcal{N}, \mathcal{E}, \mathcal{P})$, nodes $\mathcal{N} = n_1\ldots n_{|\mathcal{N}|}$ contain sequences of DNA.
+Each node $n_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{n_i}$.
 A node strand $s$ corresponds to one node orientation.
-Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands: $e_i = ( s_a, s_b )$, encoding the base topology of the sequence graph.
-Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ describe walks over node strands: $p_i = s_1 \ldots s_{|p_i|}$, storing a collection of genomes written in terms of the sequence graph.
+Edges $\mathcal{E} = e_1\ldots e_{|\mathcal{E}|}$ connect ordered pairs of node strands ($e_i = ( s_a, s_b )$), encoding the base topology of the graph.
+Paths $\mathcal{P} = p_1\ldots p_{|\mathcal{P}|}$ describe walks over node strands ($p_i = s_1 \ldots s_{|p_i|}$), representing the collection of genomes embedded in the graph.
 \end{definition}
@@
 \begin{proof}
-By definition \ref{def:vg}, two paths have identical subsequences where they walk (or ``step'') through the same series of oriented nodes (e.g. $s_1 s_2 s_3$).
+By definition \ref{def:vg}, two paths have identical subsequences where they walk (or step) through the same series of oriented nodes (\textit{e.g.} $s_1 s_2 s_3$).
 An identical set of path steps is thus equivalent to a sequence match.
@@
 We can build a variation graph from sequences and pairwise alignments.
-The variation graph will fully embed both the sequences and all pairwise relationships in the input.
+The resulting variation graph fully embeds both the sequences and all pairwise relationships in the input.
 \end{theorem}
@@
 %Assume that $i$ and $j$ are coordinates in $Q$.
-A match $m = (i, j)$ asserts the evolutionary equivalence of two charaters in sequences in $Q$.
+A match $m = (i, j)$ asserts the evolutionary equivalence of two characters in sequences in $Q$.
 %$\forall_{(i, j) : j = |g| - i} g[i] = \bar{g}[j]$.
 Pairwise alignments between sequences in $Q$ are a set of matches $A = \{ m_1 \ldots m_{|A|} \}$.
-By standard definition, each sequence matches its own reverse complement: $g[i] = \bar{g}[j]$ for all $j = |g| - i$, and we assume these matches are included in $A$.
+By standard definition, each sequence matches its own reverse complement, that is $g[i] = \bar{g}[j]$ for all $j = |g| - i$, and we assume these matches are included in $A$.
 %A match $m$ can be understood as a component in a binary relation.
@@
 We take the first match in $A$, $m_1$, and execute a union-find operation to obtain $m_1^+$.
-We add the character of the match $c(m_1^+)$ as a node $n_1$ in $\mathcal{V}$, record the mapping from $m_1^+ \to n_1$.
-To induce the graph, we take the next un-used match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $\mathcal{V}$.
+We add the character of the match $c(m_1^+)$ as a node $n_1$ in $\mathcal{V}$, and record the mapping from $m_1^+ \to n_1$.
+To induce the graph, we take the next unused match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $\mathcal{V}$.
 To allow the annotation of paths, we record the set of characters in $Q$ that match to a given node in $\mathcal{V}$ in mapping $Z = Q \to \mathcal{N} = m_1 \ldots m_{|\mathcal{N}|}$.
@@
 The set of alignments represented by a variation graph is strictly larger than the set of alignments used to induce it.
-The graph must contain at least the set of alignments given on input.
-It may also contain new implied pairwise relationships that arise due to transitive match relationships, as shown in closures 1 and 6 in Figure \ref{fig:induction}.
-However, by definition of $\mathcal{V}$, it cannot contain \textit{less} match information than represented in $A$.
+The graph must contain at least the set of alignments given in input.
+It may also contain new implied pairwise relationships that arise due to transitive match relationships, as shown in Figure \ref{fig:induction} for closures 1 and 6.
+However, by definition of $\mathcal{V}$, it cannot contain \textit{less} match information than represented in the set of matches ($A$).
 
@@
 If $A$ is sorted, we can find the matches of a given character in $Q$ using binary search, which allows us to compute $m^+$ for each character.
-We do so non-redundantly by marking each used charcater in $Q$ in an auxiliary data structure (e.g. a bitvector of $|Q|$).
-As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{N}$ and the the sequence-to-graph mapping $Z$, which like $A$ consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{N}$.
+We do so non-redundantly by marking each used character in $Q$ in an auxiliary data structure (\textit{e.g.} a bitvector of $|Q|$).
+As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{N}$ and the sequence-to-graph mapping $Z$, which, like $A$, consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{N}$.
 As with $A$, we can sort $Z$ to obtain random access via binary search.
@@
 The inductive proof of theorem \ref{thm:seqwish} demonstrates how to build a variation graph from sequences and their pairwise alignments.
-However, a na\"{i}ve algorithm based on this model will require a very large amount of space.
+However, a na\"{i}ve algorithm based on this model would require a very large amount of space.
 Take $\mathcal{Q} = |S|$.
@@
 
-Assuming that we want to build a graph for e.g.
-100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$ we might expect to use $\approx O(\num{e15})$ identifiers to store the full model.
-Such a design is almost certainly infeasible for inputs larger than a handful of genomes.
+Assuming that we want to build a graph, \textit{e.g.}, 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx O(\num{e15})$ identifiers to store the full model.
+Such a design is almost infeasible for inputs larger than a handful of genomes.
 For instance, we would need $O(\num{3e12})$ identifiers for just 5 human genomes.
@@
 In sum, we would expect to require $O(|A|\log|A| + \mathcal{N} \log |A| + 2(\mathcal{Q} \log \mathcal{Q}) + 2(\mathcal{P} \log 2\mathcal{P} + \mathcal{P}))$.
-Using our approximate relationships to $\mathcal{Q}$ given previously, and simplifying we arrive at $O[\mathcal{Q}(2N^2 \log N / \log \mathcal{Q} + (1/N + 4)\log \mathcal{Q} + 2 + log(4))]$.
+Using our approximate relationships to $\mathcal{Q}$ given previously and simplifying, we arrive at $O[\mathcal{Q}(2N^2 \log N / \log \mathcal{Q} + (1/N + 4)\log \mathcal{Q} + 2 + log(4))]$.
 
@@
 As the bounds analysis shows, space requirements make it impractical to apply a trivial version of theorem \ref{thm:seqwish} to generate a large pangenome graph.
-We need a compression approach that will exploit redundancy in the input genomes to reduce the costs of the algorithm.
-With large numbers of genomes, alignments dominate our costs.
+Therefore, we need a compression approach that exploits redundancy in the input genomes to reduce the costs of the algorithm.
+In pratical, with large numbers of genomes, alignments dominate the computational costs.
 A simple technique is to generalize matches $m = (i, j)$ to pairs of ranges in $Q$.
@@
 
-For simplicity, we have thus-far presented a character-level model of variation graph induction.
-However, range, or ``run'' compression can also reduce the representation size of the graph.
-Rather than recording an identifier for each character in a sequence graph, it is useful to ``compact'' characters that form trivial linear components in the graph into single nodes.
-Broadly, the size of nodes will be bounded by the average distance between variants, which for pangenomes built from ~100 individuals of the same species, often provides a great reduction in the total number of nodes (and thus identifiers) required for $\mathcal{V}$ and its components.
+For simplicity, we have thus far presented a character-level model of variation graph induction.
+However, range (or run) compression can also reduce the representation size of the graph.
+Rather than recording an identifier for each character in a sequence graph, it is useful to compact characters that form trivial linear components in the graph into single nodes.
+Broadly, the size of nodes will be bounded by the average distance between variants, which, for pangenomes built from \textasciitilde100 individuals of the same species, often provides a great reduction in the total number of nodes (and thus identifiers) required for $\mathcal{V}$ and its components.
 
@@
 
+\FIXME{For reasons of computational efficiency, node identifiers are numeric.}
+
 \subsection{Induction partitioning}
 
-Although match compression provides an approximate factor $k$ improvement in memory bounds for key data structures used in the induction, the approach we present in section \ref{sec:matchcompression} requires working memory in the order of the set of transitive match closures in our graph.
+Although match compression provides an approximate factor $k$ improvement in memory bounds for key data structures used in the induction, the approach we present in section \ref{sec:matchcompression} requires working memory in the order of the set of transitive match closures in the graph.
 A simple approach to reduce this bound is to divide the induction problem into smaller pieces.
@@
 We induce the related graph for each partition, appending the induction results to the appropriate data structures.
-This partitioning can introduce boundary effects which change the contents of $\mathcal{Z}$ and $\bar{\mathcal{Z}}$) by splitting ranges at the boundaries of our partitions.
+This partitioning can introduce boundary effects which change the contents of $\mathcal{Z}$ and $\bar{\mathcal{Z}}$ by splitting ranges at the boundaries of our partitions.
 However, while this will affect the compressed node definition $\mathcal{N}'$, it does not affect $\mathcal{N}$, and it can be corrected via a post-processing step to sort and compact the id space.
@@
 Here, we describe our specific implementation of this algorithm: \textit{seqwish}.
-In general, our approach uses external memory to elaborate the graph, and we assume the availability of low-latency storage media (like SSDs) to maximize the performance of this approach.
-
-\subsection{Key disk-backed data structures}
-
-In our implementation, we rely on several basic external memory kernels.
-To reduce working memory requirements to an absolute minimum, we use a disk-backed version of the implicit interval tree that memory-maps the sorted array of intervals (mmiitree).
-Indexing the implicit interval tree requires a sorting step which dominates the runtime of our algorithm.
-We adapt the current best-performing in-place parallel sorting algorithm, In-place Parallel Super Scalar Samplesort (IPS$^4$o), to work on a disk-backed, memory-mapped array \citep{axtmann2017}.
-This allows us work with $\mathcal{A}$, $\mathcal{Z}$, and $\bar{\mathcal{Z}}$ in external memory.
-By storing pairs of numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$ in terms of offsets in $\mathcal{N}$.
-The graph sequence vector $\mathcal{N}$ is simply written by appending characters to a file.
-We ``mark'' nodes to generate $\mathcal{N}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \cite{Gog_2014} for support of the final emission of the graph $\mathcal{V}$.
+In general, our approach uses external memory to elaborate the graph, taking advantage of the availability of low-latency storage media, like solid-state drives (SSDs), to maximize the performance of this approach.
 
@@
 
-A practical implementation introduces a new set of problems related to input and output. % that we have not yet introduced.
+%A practical implementation introduces a new set of problems related to input and output. % that we have not yet introduced.
 Our implementation reads standard data formats, FASTA or FASTQ for the input sequences, and PAF \citep{Li_2018} for pairwise alignments.
@@
 To efficiently process the input PAF, we thus need to build a sequence index that allows us to generate $\mathcal{A}$.
-We often wish to build graphs from very large collections of sequences, such as raw sequencing reads or contigs from many thousands of samples.
-We build a compressed suffix array (CSA) \citep{Sadakane_2000} over sequence names, and provide auxiliary supporting data structures that allow us to map between our input and the abstract concatenation of all input sequences and their reverse complements: $Q$.
+In particular, we build a compressed suffix array (CSA) \citep{Sadakane_2000} over sequence names, that we call \textit{seqidx}, and provide auxiliary supporting data structures that allow us to map between our input and the abstract concatenation of all input sequences and their reverse complements ($Q$).
+We often build graphs from very large collections of sequences, such as raw sequencing reads or contigs from many thousands of samples.
 This \textit{seqidx} avoids the overheads associated with a hash table on string names of input sequences.
@@
 
-For output in GFA, the process is relatively simple.
-We iterate over nodes in $\mathcal{N}'$, writing each as a node record.
-Edges can be similarly produced from the disk-backed multiset representing $\mathcal{E}$.
-The most involved part of graph emission is the rendering of our input sequences $S$ as paths $\mathcal{P}$ through the graph.
-For each input sequence in our \textit{seqidx}, we walk through the offsets in $S$ contained in the sequence and look up their mapping into $\mathcal{N}'$ using $\mathcal{Z}$.
+For output in GFA, we iterate over nodes in $\mathcal{N}'$, writing each as a node record.
+Edges are similarly produced from the disk-backed multiset representing $\mathcal{E}$.
+The most computationally expensive part of graph emission is the rendering of the input sequences $S$ as paths $\mathcal{P}$ through the graph.
+For each input sequence in the \textit{seqidx}, we walk through the offsets in $S$ contained in the sequence and look up their mapping into $\mathcal{N}'$ using $\mathcal{Z}$.
 Range compression allows us to complete one lookup per range.
-By definition, each character is $S$ is covered by only one range in $\mathcal{Z}$.
+By definition, each character in $S$ is covered by only one range in $\mathcal{Z}$.
 We can thus iterate through the ranges in $\mathcal{Z}$ without considering each character.
 Following the GFA format, we are able to independently generate $\mathcal{P}$, as each path is represented on a separate record in the GFA.
+
+\subsection{Key disk-backed data structures}
+
+In our implementation, we rely on several basic external memory kernels.
+To reduce working memory requirements to an absolute minimum, we use a disk-backed version of the implicit interval tree that memory-maps the sorted array of intervals \citep{mmmulti}.
+Indexing the implicit interval tree requires a sorting step which dominates the runtime of our algorithm.
+We adapt the current best-performing in-place parallel sorting algorithm, In-place Parallel Super Scalar Samplesort (IPS$^4$o), to work on a disk-backed, memory-mapped array \citep{axtmann2017}.
+This allows us work with $\mathcal{A}$, $\mathcal{Z}$, and $\bar{\mathcal{Z}}$ in external memory.
+By storing pairs of numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$ in terms of offsets in $\mathcal{N}$.
+The graph sequence vector $\mathcal{N}$ is simply written by appending characters to a file.
+We mark nodes to generate $\mathcal{N}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \cite{Gog_2014} for support of the final emission of the graph $\mathcal{V}$.
 
@@
 %We thank Richard Durbin, Benedict Paten, and Simon Puglisi for support and helpful discussions in the development of this algorithm.
-We are greatful to members of the HPRC production teams for their development of resources used in our exposition.
+We are grateful to members of the HPRC production teams for their development of resources used in our exposition.
 
@@
 We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG).
+\linebreak
+\linebreak
+\textit{Conflict of Interest}: none declared.
 

### 2022-01-19 scripts/log2info.py
update fish and Zmays workflows; added `log2info.py` script to get info from the log files

@@ -1,0 +1,47 @@
+# Usage:
+#    grep 'Command\|wall\|Maximum' logs/*/*.log | python3 scripts/log2info.py | sort -k 1,3 -k4,4n -k 5,6n -k 7,7 | column -t
+
+import sys
+
+input = ''
+s = ''
+l = ''
+p = ''
+n = ''
+k = ''
+B = ''
+elapsed_wall_clock_time = ''
+max_resident_set_size = ''
+
+for line in sys.stdin:
+    if 'Command being timed' in line:
+        gfa_name = line.split('.gfa')[0].split('/')[-1]
+        if len(gfa_name.split('.')) == 7:
+            # Input dataset
+            input = gfa_name.split('.')[0]
+
+            # Remove the parameter name (single character) from each string
+            s, l, p, n, k, B = [x[1:] for x in gfa_name.split('.')[1:]]
+
+        elapsed_wall_clock_time = ''
+        max_resident_set_size = ''
+    elif 'Elapsed (wall clock) time' in line:
+        elapsed_wall_clock_time = line.strip().split('): ')[-1]
+
+        max_resident_set_size = ''
+    elif 'Maximum resident set size' in line:
+        max_resident_set_size = line.strip().split('): ')[-1]
+
+        # Check if all information are available
+        if input and s and l and p and n and k and B and elapsed_wall_clock_time and max_resident_set_size:
+            print(input, s, l, p, n, k, B, elapsed_wall_clock_time, max_resident_set_size)
+
+        input = ''
+        s = ''
+        l = ''
+        p = ''
+        n = ''
+        k = ''
+        B = ''
+        elapsed_wall_clock_time = ''
+        max_resident_set_size = ''

### 2022-01-19 scripts/log2info.py
update HPRC+ workflow

@@ -1,5 +1,5 @@
 # Usage:
-#    grep 'Command\|wall\|Maximum' logs/*/*.log | python3 scripts/log2info.py | sort -k 1,3 -k4,4n -k 5,6n -k 7,7 | column -t
+#    cat *.log | python3 scripts/log2info.py | sort -k 1,3 -k4,4n -k 5,5n -k 6,6n -k 7,7 | column -t
 
 import sys
 

### 2022-01-20 manuscript/main.tex
clean intro

@@
 
-\title[Multi-reference pangenome graphs]{Multi-reference pangenome graphs}
+\title[Unbiased pangenome graphs]{Unbiased pangenome graphs}
 \author[Garrison \textit{et~al}.]{
@@
 Existing approaches depend on a single reference or a specific ordering of genomes, and do not provide base-level information about the mapping of genomes into the pangenome.
-A scalable method to build pangenome graphs without such limitations would be a key functional step in pangenome construction and manipulation pipelines. \\
+A scalable method to build pangenome graphs without such limitations would be a key step in pangenome construction and manipulation pipelines. \\
 \textbf{Results:}
@@
 %, and may provide a simpler representation of structurally-variable regions.
-Popular approaches model the pangenome alignment as a \textit{pangenome graph}.
-\FIXME{Conceptually, the pangenome graph relates genomes as walks through and underlying DNA sequence language encoded as a sequence graph.}
-In a pangenome graph, variation can be understood in the context of any part of any included genome, which lets us avoid the problem of reference bias, or a limitation of analyses to only sequences that are very similar to a chosen reference genome \citep{Eizenga_2020}.
+Many approaches model the pangenome alignment as a \textit{pangenome graph} \cite{Garrison_2018,Yokoyama2019,Hickey:2020}.
+A pangenome graph encodes DNA sequences as walks through an underlying language encoded in a sequence graph \citep{Hein_1989}.
+In a pangenome graph, variation can be understood in the context of any part of any included genome  \citep{Eizenga_2020}.
+This lets us avoid the problem of reference bias, which can be understood as the a limitation of analyses to genome sequences that are similar to a chosen reference genome.
+%Reference bias distorts an inferred genome to appear more similar to the reference genome than it actually is.
 
-Surprisingly, existing methods to build pangenome graphs from large genome assemblies are reference-based.
-Minigraph progressively builds a pangenome graph by layering an ordered series of input genomes onto a specific reference \citep{Li:2020}.
-Cactus depends on a single guide-tree to structure pairwise alignments between genomes \citep{Armstrong:2020}.
-Both methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, and guide-tree topology.
-Furthermore, Minigraph does not provide base-level information, limiting the analysis to only large structural variation.
-Although a reference-guided approach is pragmatic when we only have a single reference genome, recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies in humans and other species \citep{Nurk_2021}.
+An unbiased pangenome graph would represent the alignment of all included genomes to all others.
+Existing methods approximate this relationship by progressive alignment to a graph initially based on a reference genome \citep{Li:2020}, or through a global structuring of the genome relationships in a neighbor joining phylogenetic tree \citep{Armstrong:2020}.
+These methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, and guide-tree topology.
 
-A class of methods to represent pangenomes involves the \textit{sequence graph} \citep{Hein_1989}, where identical segments of different genomes collapse into a single representative segment (often node) in the graph.
-In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
-A \textit{bidirected} sequence graph represents both strands of DNA.
-On this model, variation graphs add the concept of \textit{path} to embed linear sequences (\textit{e.g.}, genomes or haplotypes) into the graph \citep{Garrison_2018}.
-Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+We consider the problem of building a pangenome graph without these potential sources of bias.
+Such a graph would be an ideal system to represent variation between two or more high-quality genomes.
+Given the rapid development of complete genome assemblies for humans and other vertebrates \citep{Rhie_2021,Nurk_2021}, we need a practical approach that can achieve this for tens to thousands of genomes on commodity hardware.
+Here, we present \textit{seqwish}, an algorithm for the generation of a pangenome graph from of pairwise alignments.
+%It transforms a collection of sequences and their alignments into a series of intermediate representations that support the computation of an equivalent pangenome graph.
+Our solution is simple, but experiments on diverse sequence collections demonstrate that it easily scales to large pangenome building problems.
+%It is also, by focus on the key graph induction step, a simple component.
 
-A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Eizenga_2020}.
-Motivated by the need to generate ``multi-reference'' pangenome graphs, we develop \textit{seqwish}, an algorithm that allows us to build pangenome graphs in the form of variation graphs directly from a collection of sequences and pairwise alignments between them.
-Here, we formally describe the algorithm, evaluate its basic combinatorial bounds and costs, and present applications of the method to pangenome building problems in a variety of species.
+%Our solution is simple, but flexible, and we demonstrate its scalability to pangenome construction problems in diverse genomes.
+
+%As the diffic for model organisms \citep{Rhie_2021,Nurk_2021}
+%, we consider the need to efficiently construct \textit{unbiased} pangenome graphs based on symmetric relationships among sequences.
+
+%Furthermore, Minigraph does not provide base-level information, limiting the analysis to only large structural variation.
+%Although a reference-guided approach is pragmatic when we only have a single reference genome, recent advances in sequencing and assembly are driving the generation of many reference-quality assemblies in humans and other species \citep{Nurk_2021}.
+
+%A class of methods to represent pangenomes involves the \textit{sequence graph} \citep{Hein_1989}, where identical segments of different genomes collapse into a single representative segment (often node) in the graph.
+%In \textit{node-labeled} sequence graphs, nodes indicate DNA sequences, with edges connecting the nodes that are concatenated in the sequences represented in the graph.
+%A \textit{bidirected} sequence graph represents both strands of DNA.
+%On this model, variation graphs add the concept of \textit{path} to embed linear sequences (\textit{e.g.}, genomes or haplotypes) into the graph \citep{Garrison_2018}.
+%Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
+
+%A pangenome graph is a sequence model that encodes the mutual alignment of many genomes \citep{Eizenga_2020}.
+
+%Motivated by the need to generate ``multi-reference'' pangenome graphs, we develop \textit{seqwish}, an algorithm that allows us to build pangenome graphs in the form of variation graphs directly from a collection of sequences and pairwise alignments between them.
+%Here, we formally describe the algorithm, evaluate its basic combinatorial bounds and costs, and present applications of the method to pangenome building problems in a variety of species.
+
 %Our approach is both highly generic and scalable.
@@
 It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
+In our experiments, we have not explored the full problem, including alignment.
+We regard this as outside of the scope of the presented work, which focuses on a key kernel which is a bottleneck in the pangenome graph construction process.
 

### 2022-01-20 manuscript/main.tex
typos

@@
 In a pangenome graph, variation can be understood in the context of any part of any included genome  \citep{Eizenga_2020}.
-This lets us avoid the problem of reference bias, which can be understood as the a limitation of analyses to genome sequences that are similar to a chosen reference genome.
+This lets us avoid the problem of reference bias, which can be understood as the limitation of analyses to genome sequences that are similar to a chosen reference genome.
 %Reference bias distorts an inferred genome to appear more similar to the reference genome than it actually is.
@@
 Given the rapid development of complete genome assemblies for humans and other vertebrates \citep{Rhie_2021,Nurk_2021}, we need a practical approach that can achieve this for tens to thousands of genomes on commodity hardware.
-Here, we present \textit{seqwish}, an algorithm for the generation of a pangenome graph from of pairwise alignments.
+Here, we present \textit{seqwish}, an algorithm for the generation of a pangenome graph from pairwise alignments.
 %It transforms a collection of sequences and their alignments into a series of intermediate representations that support the computation of an equivalent pangenome graph.

### 2022-01-21 manuscript/main.tex
tweaks

--- 2022-01-20 manuscript/main.tex
@@
 We continue until all matches have been used.
-Finally, we establish paths ($P$) by walking them in $\mathcal{V}$ using $Z$, and record edges ($E$) where nodes occur successively in paths.
+Finally, we establish paths ($\mathcal{P}$) by walking them in $\mathcal{V}$ using $Z$, and record edges ($\mathcal{E}$) where nodes occur successively in paths.
 
@@
 Therefore, we need a compression approach that exploits redundancy in the input genomes to reduce the costs of the algorithm.
-In pratical, with large numbers of genomes, alignments dominate the computational costs.
-A simple technique is to generalize matches $m = (i, j)$ to pairs of ranges in $Q$.
-For highly-similar sequences, we expect exact matches to occur in long runs.
+When working with large numbers of genomes, alignments dominate the computational costs.
+A simple technique is to generalize matches $m = (i, j)$, which are between individual characters, to \textit{range}-matches over pairs of ranges of characters in $Q$.
+For highly-similar sequences, our expectation is that exact matches will occur in long runs.
 If the average pairwise diversity of sequences in our input is $1/k$, we expect exact matches to be around $k$ characters long.
-By encoding matches as pairs of ranges of characters, $r = (a, b) : a, b = (i, j) : i, j \in Q$, we can obtain a $\approx$ $k$-fold compression of $A$, which for clarity we label as $\mathcal{A}$.
-If sorted, $\mathcal{A}$ can be treated as an \textit{implicit interval tree} \citep{Li_bedtk_2020}, which allows queries of containment and overlap in $O(|\mathcal{A}| \log |\mathcal{A}|)$ time.
+By encoding matches as pairs of ranges of characters, $r = (a, b) : a, b = (i, j) : i, j \in Q$, we can obtain a $\approx$ $k$-fold compression of $A$, yielding the range-match array $\mathcal{A}$.
+
+If sorted, $\mathcal{A}$ can be treated as an \textit{implicit interval tree} \citep{Li_bedtk_2020}, which allows queries of containment and overlap in $O(\log |\mathcal{A}|)$ time.
 This compression requires trivial changes to our graph induction model.
@@
 \FIXME{to do}
-We can apply our method to any collection of sequences and alignments.
+
+We can apply our method to various collections of sequences and alignments.
 It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
+
+However, this exposition comes with significant limitations.
 In our experiments, we have not explored the full problem, including alignment.
-We regard this as outside of the scope of the presented work, which focuses on a key kernel which is a bottleneck in the pangenome graph construction process.
+These lie outside of the scope of the presented work, wherein whe have focused on a key kernel which is a bottleneck in the pangenome graph construction process.
+Instead, \textit{seqwish} is a generic ``graph inducer'' of potentially many uses.
+But, it is important to note that the graphs which it produces are simply raw representations of an underlying alignment set.
+Due to differences in alignments that arise in regions of lower entropy, in practice, it is necessary to normalize generated graphs to simplify the local alignment structures represented in the graph.
 

### 2022-01-21 scripts/seqwish_with_logging.sh
Clean the directory after seqwish (in case it ends incorrectly)

@@ -12,9 +12,15 @@ hostname
 
 cd /scratch
 
+# Start the process to fill the log file
 (while true; do (date +%s; du -s .; ls -s; echo) >>"$LOG"; sleep "$SECS"; done) &
 SIZE_DEMON_PID=$!
 
+# Run seqwish
 \time -v ~/tools/seqwish/bin/seqwish-ccfefb016fcfc9937817ce61dc06bbcf382be75e -t 48 -s "$ASSEMBLIES" -p "$PAF" -g "$GFA" -k "$k" -B "$B" -P
 
+# Stop filling the log file
 kill $SIZE_DEMON_PID
+
+# Clean the directory (in case seqwish ends incorrectly)
+rm "$GFA".{sqa,sqi,sql,sqn,sqp,sqq,sqs}

### 2022-01-21 scripts/seqwish_with_logging.sh
ignore nonexistent files and arguments, never prompt

@@ -23,4 +23,4 @@ SIZE_DEMON_PID=$!
 kill $SIZE_DEMON_PID
 
 # Clean the directory (in case seqwish ends incorrectly)
-rm "$GFA".{sqa,sqi,sql,sqn,sqp,sqq,sqs}
+rm -f "$GFA".{sqa,sqi,sql,sqn,sqp,sqq,sqs}

### 2022-01-21 scripts/log2info.py
filter failing logs

@@ -13,8 +13,12 @@ B = ''
 elapsed_wall_clock_time = ''
 max_resident_set_size = ''
 
+reject_result = False
+
 for line in sys.stdin:
-    if 'Command being timed' in line:
+    if 'Command terminated by signal' in line:
+        reject_result = True
+    elif 'Command being timed' in line:
         gfa_name = line.split('.gfa')[0].split('/')[-1]
         if len(gfa_name.split('.')) == 7:
             # Input dataset
@@ -33,7 +37,7 @@ for line in sys.stdin:
         max_resident_set_size = line.strip().split('): ')[-1]
 
         # Check if all information are available
-        if input and s and l and p and n and k and B and elapsed_wall_clock_time and max_resident_set_size:
+        if not reject_result and input and s and l and p and n and k and B and elapsed_wall_clock_time and max_resident_set_size:
             print(input, s, l, p, n, k, B, elapsed_wall_clock_time, max_resident_set_size)
 
         input = ''
@@ -45,3 +49,5 @@ for line in sys.stdin:
         B = ''
         elapsed_wall_clock_time = ''
         max_resident_set_size = ''
+
+        reject_result = False

### 2022-01-22 scripts/log2info.py
remove (old) logs; updated the scripts, by adding a new `sizeLog2diskUsage.py`

@@ -1,8 +1,9 @@
 # Usage:
-#    cat *.log | python3 scripts/log2info.py | sort -k 1,3 -k4,4n -k 5,5n -k 6,6n -k 7,7 | column -t
+#    cat *.log | python3 scripts/log2info.py
 
 import sys
 
+gfa_name = ''
 input = ''
 s = ''
 l = ''
@@ -36,10 +37,14 @@ for line in sys.stdin:
     elif 'Maximum resident set size' in line:
         max_resident_set_size = line.strip().split('): ')[-1]
 
-        # Check if all information are available
-        if not reject_result and input and s and l and p and n and k and B and elapsed_wall_clock_time and max_resident_set_size:
-            print(input, s, l, p, n, k, B, elapsed_wall_clock_time, max_resident_set_size)
+        # Convert in Gbytes
+        max_resident_set_size = '{:.4f}'.format(float(max_resident_set_size)/1024/1024)
 
+        # Check if all information are available
+        if not reject_result and gfa_name and input and s and l and p and n and k and B and elapsed_wall_clock_time and max_resident_set_size:
+            print('\t'.join([gfa_name, input, s, l, p, n, k, B, elapsed_wall_clock_time, max_resident_set_size]))
+
+        gfa_name = ''
         input = ''
         s = ''
         l = ''

### 2022-01-22 scripts/seqwish_with_logging.sh

@@ -13,6 +13,11 @@ hostname
 cd /scratch
 
 # Start the process to fill the log file
+### ls -s;  -s/--sizeprint the allocated size of each file, in blocks
+### To check the block_size:
+###echo 1 > sizeTest
+###ls -s --block-size 1 sizeTest
+###rm sizeTest
 (while true; do (date +%s; du -s .; ls -s; echo) >>"$LOG"; sleep "$SECS"; done) &
 SIZE_DEMON_PID=$!
 

### 2022-01-22 scripts/sizeLog2diskUsage.py

@@ -1,0 +1,45 @@
+# Usage:
+#    cat *.log | python3 scripts/sizeLog2diskUsage.py
+
+import sys
+
+BLOCK_SIZE_BYTES = 1024
+
+file_to_max_size_bytes_dict = {}
+
+for line in sys.stdin:
+    if 'total' not in line and '.seqnames.tmp' not in line:
+        if len(line.strip().split(' ')) == 2:
+            current_size_bytes, file = line.strip().split(' ')
+
+            if 'gfa' in file:
+                current_size_bytes = int(current_size_bytes)
+
+                if file not in file_to_max_size_bytes_dict:
+                    file_to_max_size_bytes_dict[file] = current_size_bytes
+                else:
+                    file_to_max_size_bytes_dict[file] = max(file_to_max_size_bytes_dict[file], current_size_bytes)
+
+# Convert the measurements in bytes
+for file in file_to_max_size_bytes_dict:
+    file_to_max_size_bytes_dict[file] *= BLOCK_SIZE_BYTES
+
+# Debugging: print GFA output files in GBs
+#for file, max_size_bytes in file_to_max_size_bytes_dict.items():
+#    if file.endswith('.gfa'):
+#        print(file, max_size_bytes/1024/1024/1024)
+
+# Sum max sizes for each run
+run_to_max_disk_usage_bytes_dict = {}
+
+for file, max_size_bytes in file_to_max_size_bytes_dict.items():
+    prefix = file.split('.gfa')[0]
+
+    if prefix not in run_to_max_disk_usage_bytes_dict:
+        run_to_max_disk_usage_bytes_dict[prefix] = max_size_bytes
+    else:
+        run_to_max_disk_usage_bytes_dict[prefix] += max_size_bytes
+
+for run, max_disk_usage_bytes in run_to_max_disk_usage_bytes_dict.items():
+    max_disk_usage_str = '{:.4f}'.format(max_disk_usage_bytes/1024/1024/1024)
+    print('\t'.join([run, max_disk_usage_str]))

### 2022-01-25 scripts/log2info.py
put all statistics together in a single table

@@ -3,8 +3,9 @@
 
 import sys
 
+input_fasta = ''
+input_paf = ''
 gfa_name = ''
-input = ''
 s = ''
 l = ''
 p = ''
@@ -21,10 +22,11 @@ for line in sys.stdin:
         reject_result = True
     elif 'Command being timed' in line:
         gfa_name = line.split('.gfa')[0].split('/')[-1]
-        if len(gfa_name.split('.')) == 7:
-            # Input dataset
-            input = gfa_name.split('.')[0]
 
+        input_fasta = line.split('-s ')[1].split('.fasta.gz')[0].split('/')[-1]
+        input_paf = line.split('-p ')[1].split('.paf.gz')[0].split('/')[-1]
+
+        if len(gfa_name.split('.')) == 7:
             # Remove the parameter name (single character) from each string
             s, l, p, n, k, B = [x[1:] for x in gfa_name.split('.')[1:]]
 
@@ -41,11 +43,12 @@ for line in sys.stdin:
         max_resident_set_size = '{:.4f}'.format(float(max_resident_set_size)/1024/1024)
 
         # Check if all information are available
-        if not reject_result and gfa_name and input and s and l and p and n and k and B and elapsed_wall_clock_time and max_resident_set_size:
-            print('\t'.join([gfa_name, input, s, l, p, n, k, B, elapsed_wall_clock_time, max_resident_set_size]))
+        if not reject_result and input_fasta and input_paf and gfa_name and input and s and l and p and n and k and B and elapsed_wall_clock_time and max_resident_set_size:
+            print('\t'.join([gfa_name, input_fasta, input_paf, s, l, p, n, k, B, elapsed_wall_clock_time, max_resident_set_size]))
 
+        input_fasta = ''
+        input_paf = ''
         gfa_name = ''
-        input = ''
         s = ''
         l = ''
         p = ''

### 2022-01-26 scripts/log2info.py
runtime in seconds; compress GFA files

@@ -35,6 +35,21 @@ for line in sys.stdin:
     elif 'Elapsed (wall clock) time' in line:
         elapsed_wall_clock_time = line.strip().split('): ')[-1]
 
+        if len(elapsed_wall_clock_time.split(':')) == 3:
+            # hh:mm:ss
+            hh, mm, ss = elapsed_wall_clock_time.split(':')
+
+            elapsed_wall_clock_time = str(int(hh) * 3600 + int(mm) * 60 + int(ss))
+        elif len(elapsed_wall_clock_time.split(':')) == 2:
+            # mm:ss.xx
+            mm, ss = elapsed_wall_clock_time.split(':')
+            ss = ss.split('.')[0]
+
+            elapsed_wall_clock_time = str(int(mm) * 60 + int(ss))
+        else:
+            # Not expected
+            elapsed_wall_clock_time = ''
+
         max_resident_set_size = ''
     elif 'Maximum resident set size' in line:
         max_resident_set_size = line.strip().split('): ')[-1]

### 2022-01-28 scripts/experiment_plot.R
add figure to describe experiments

@@ -1,0 +1,18 @@
+#!/usr/bin/env Rscript
+
+require(tidyverse)
+require(ggrepel)
+
+x <- read.delim('stats/stats.tsv')
+
+ggplot(subset(x, input.fasta=="athaliana16" & min.mapping.len=="l60k" | input.fasta=="hprcplus38" | input.fasta=="hpylori250" | input.fasta=="zmays41" & min.mapping.len=="l60k"),
+       aes(x=graph.length/((Gbps*1e9)/n), y=time.seconds/3600, shape=map.ident, color=as.factor(k), label=min.mapping.len)) +
+geom_point() +
+facet_wrap(~input.fasta, scales="free") +
+scale_x_continuous("graph length / single genome length") +
+scale_y_continuous("runtime (hours)") +
+scale_color_discrete("seqwish -k") +
+scale_shape_discrete("wfmash -p")
+#+ geom_text_repel(size=2)
+
+ggsave("manuscript/fig_experiment_stats.pdf", height=5, width=7)

### 2022-01-28 scripts/experiment_plot.R
add experiment figure

@@ -15,4 +15,4 @@ scale_color_discrete("seqwish -k") +
 scale_shape_discrete("wfmash -p")
 #+ geom_text_repel(size=2)
 
-ggsave("manuscript/fig_experiment_stats.pdf", height=5, width=7)
+ggsave("manuscript/fig_experiment_stats.pdf", height=4, width=5.5)

### 2022-01-28 manuscript/main.tex
performance table
italics for the species
add experiment figure
less italics

@@
 %A \textit{bidirected} sequence graph represents both strands of DNA.
-%On this model, variation graphs add the concept of \textit{path} to embed linear sequences (\textit{e.g.}, genomes or haplotypes) into the graph \citep{Garrison_2018}.
+%On this model, variation graphs add the concept of \textit{path} to embed linear sequences (e.g., genomes or haplotypes) into the graph \citep{Garrison_2018}.
 %Paths provide a stable coordinate system, allowing graph annotations and interoperability between different graphs.
@@
 \begin{proof}
-By definition \ref{def:vg}, two paths have identical subsequences where they walk (or step) through the same series of oriented nodes (\textit{e.g.} $s_1 s_2 s_3$).
+By definition \ref{def:vg}, two paths have identical subsequences where they walk (or step) through the same series of oriented nodes (e.g. $s_1 s_2 s_3$).
 An identical set of path steps is thus equivalent to a sequence match.
@@
 If $A$ is sorted, we can find the matches of a given character in $Q$ using binary search, which allows us to compute $m^+$ for each character.
-We do so non-redundantly by marking each used character in $Q$ in an auxiliary data structure (\textit{e.g.} a bitvector of $|Q|$).
+We do so non-redundantly by marking each used character in $Q$ in an auxiliary data structure (e.g. a bitvector of $|Q|$).
 As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{N}$ and the sequence-to-graph mapping $Z$, which, like $A$, consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{N}$.
@@
 
-Assuming that we want to build a graph, \textit{e.g.}, 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx O(\num{e15})$ identifiers to store the full model.
+Assuming that we want to build a graph, e.g., 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx O(\num{e15})$ identifiers to store the full model.
 Such a design is almost infeasible for inputs larger than a handful of genomes.
@@
 %    A lock-free, parallel union-find algorithm \FIXME{Wait-free parallel algorithms for the union-find problem} allows us to, for each base pair of input sequence, collect all other base pairs in the pangenome that are aligned to it.
-%    By processing this transitive closure operation in chunks (\textit{e.g.} 10Mbp at a time), we limit the maximum working memory (RAM) requirements of the process without affecting the output.
+%    By processing this transitive closure operation in chunks (e.g. 10Mbp at a time), we limit the maximum working memory (RAM) requirements of the process without affecting the output.
 %    After the transitive closure, we write our final graph on disk in a set of arrays, then augment it with the paths of our input sequences,
@@
 \FIXME{to do}
+
+\begin{table*}[!ht]
+    \centering
+    \caption{Performance of the graph induction algorithm}
+    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
+        \hline
+        species & sequences & haplotypes & fasta.Gbps & min.match.bp & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbps & components \\ \hline
+        \textit{A. thaliana} & 922 & 16 & 1.90251 & 49 & 468 & 43.1287 & 7.1218 & 0.234284 & 100 \\ \hline
+        \textit{H. sapiens} & 17478 & 38 & 114.627 & 49 & 39723 & 347.9261 & 464.3392 & 6.2322 & 724 \\ \hline
+        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 475 & 33.7270 & 11.9739 & 0.0302351 & 33 \\ \hline
+        \textit{Z. mays} & 46289 & 41 & 90.2491 & 49 & 31043 & 351.1235 & 402.8716 & 13.8838 & 925 \\ \hline
+    \end{tabular}
+\end{table*}
 
@@
 %\FIXME{to do}
+
+\input{fig_experiments.tex}
 

### 2022-01-28 manuscript/main.tex
on the min match length filter

@@
 The graph sequence vector $\mathcal{N}$ is simply written by appending characters to a file.
-We mark nodes to generate $\mathcal{N}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \cite{Gog_2014} for support of the final emission of the graph $\mathcal{V}$.
+We mark nodes to generate $\mathcal{N}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \citep{Gog_2014} for support of the final emission of the graph $\mathcal{V}$.
+
+\subsection{Short match filter}
+
+Building a graph from an all-to-all alignment does not guarantee that the local structure of the graph is easy to understand.
+The all-to-all alignment is not coordinated, with each mapping aligned in isolation, and in consequence it fails to resolve the indel alignment normalization problem \citep{Mose_2019}.
+This ambiguity can introduce deeply looping structures in the graph which collapse e.g. microsatellites and other short VNTRs into very small numbers of nodes with very complex local topologies.
+Such motifs can cause problems with downstream analysis.
+We find that ambiguity about the arrangement of very short matches tends to drive complex local structures in the graph.
+
+We mitigate this issue with a simple filter, \textit{seqwish -k}, which simply ignores exact matches that are shorter than $k$ characters.
+This filter necessarily increases the size of the induced graph.
+But, it also replaces complex motifs shorter than $k$ with single bubbles.
+In doing so, it also removes short, expensive matches, reducing the overall space requirements for \textit{seqwish}.
+When set very high, this filter can be used to generate a coarse, high-confidence graph built only from very long exact matches which will tend to be unique in the genome.
+Although the application of the $k>$ filter can result in a graph that is relatively ``under-aligned'', we can further refine it through the application of local multiple sequence alignment \citep{Gao_2020}, or graph normalization \citep{Doerr_2021}.
+In a pangenomic context, underalignment caused by $k>$ match filtering can be mitigated by transitive relationships present in the pangenome.
+
+%Also, short matches are expensive to store, and may not add much to our understanding of variation or our particular use of the graph.
 
@@
     \centering
-    \caption{Performance of the graph induction algorithm}
+    \caption{Performance of the graph induction algorithm.
+        For each pangenome we have chosen a single experiment with the given \textit{seqwish -k} filter setting of 49bp.
+        }
     \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}

### 2022-01-28 scripts/experiment_plot.R
update statistics and R script; cleaning

@@ -5,8 +5,8 @@ require(ggrepel)
 
 x <- read.delim('stats/stats.tsv')
 
-ggplot(subset(x, input.fasta=="athaliana16" & min.mapping.len=="l60k" | input.fasta=="hprcplus38" | input.fasta=="hpylori250" | input.fasta=="zmays41" & min.mapping.len=="l60k"),
-       aes(x=graph.length/((Gbps*1e9)/n), y=time.seconds/3600, shape=map.ident, color=as.factor(k), label=min.mapping.len)) +
+ggplot(subset(x, input.fasta=="athaliana16" & min.mapping.len=="l60k" | input.fasta=="hprcplus38" | input.fasta=="hpylori250" & min.mapping.len=="l15k" | input.fasta=="zmays41" & min.mapping.len=="l60k"),
+       aes(x=graph.length.Gbps/n, y=time.seconds/3600, shape=map.ident, color=as.factor(k), label=min.mapping.len)) +
 geom_point() +
 facet_wrap(~input.fasta, scales="free") +
 scale_x_continuous("graph length / single genome length") +

### 2022-01-29 scripts/experiment_plot.R
fix R script and image

@@ -6,7 +6,7 @@ require(ggrepel)
 x <- read.delim('stats/stats.tsv')
 
 ggplot(subset(x, input.fasta=="athaliana16" & min.mapping.len=="l60k" | input.fasta=="hprcplus38" | input.fasta=="hpylori250" & min.mapping.len=="l15k" | input.fasta=="zmays41" & min.mapping.len=="l60k"),
-       aes(x=graph.length.Gbps/n, y=time.seconds/3600, shape=map.ident, color=as.factor(k), label=min.mapping.len)) +
+       aes(x=graph.length.Gbps/(Gbps/n), y=time.seconds/3600, shape=map.ident, color=as.factor(k), label=min.mapping.len)) +
 geom_point() +
 facet_wrap(~input.fasta, scales="free") +
 scale_x_continuous("graph length / single genome length") +

### 2022-02-06 manuscript/main.tex
update table
update table with a more compressed hpylori graph
updated statistics and manuscript with hprc+ results using 95% as identity threshold
ambig

@@
 The all-to-all alignment is not coordinated, with each mapping aligned in isolation, and in consequence it fails to resolve the indel alignment normalization problem \citep{Mose_2019}.
-This ambiguity can introduce deeply looping structures in the graph which collapse e.g. microsatellites and other short VNTRs into very small numbers of nodes with very complex local topologies.
+This ambiguity can introduce deeply looping structures in the graph which collapse polymorphic microsatellites and other short VNTRs into very small numbers of nodes with very complex local topologies.
 Such motifs can cause problems with downstream analysis.

### 2022-02-06 manuscript/fig_experiments.tex
fig experiments

\begin{figure}[b]
%  \centering
   \includegraphics[width=\linewidth]{fig_experiment_stats.pdf}
   \caption{
     Experimental results from the application of \textit{seqwish} to four different pangenomes.
     Multiple minimum identity settings for the mapping (\textit{wfmash -p}) and different minimum match length filter settings result in a collection of graph builds per pangenome input.
     We compare the runtime in hours with the size of the resulting graphs relative to the average size of an input genome in the particular set.
     Lower limits on pairwise identity result in more compact graphs.
     Similarly, filtering short matches increases graph size relative to not (\textit{seqwish -k}=0, red).
     Increasing the \textit{seqwish -k} parameter tends to increase the size of the graph.
    }
    \label{fig:experiments}
\end{figure}

### 2022-02-07 manuscript/main.tex
add cost of bitvector to mark transclosure progress

@@
 If $A$ is sorted, we can find the matches of a given character in $Q$ using binary search, which allows us to compute $m^+$ for each character.
-We do so non-redundantly by marking each used character in $Q$ in an auxiliary data structure (e.g. a bitvector of $|Q|$).
+We do so non-redundantly by marking each used character in $Q$ in an auxiliary data structure $\mathcal{X}$, which could be encoded as a bitvector of length $|Q|$.
 As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{N}$ and the sequence-to-graph mapping $Z$, which, like $A$, consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{N}$.
@@
 The induction must maintain reference to all characters in all input sequences $O(\mathcal{Q})$, all character-to-character matches $O(|A|) \approx O(\mathcal{Q}N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(\mathcal{Q})$, the nodes of the graph $O(|\mathcal{N}|) \approx O(\mathcal{Q}/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(\mathcal{Q})$, and the set of paths $O(|\mathcal{P}|) \approx O(\mathcal{Q})$.
-In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(\mathcal{Q}N^2 + 4\mathcal{Q} + \mathcal{Q}/N)$ space.
+We also maintain the bitvector $\mathcal{X}$ to mark seen characters of $Q$ during graph induction, which requires $O(\mathcal{Q})$ bits, equivalent to $O(\mathcal{Q} / \log_2\mathcal{Q})$ integer identifiers.
+In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(\mathcal{Q}(N^2 + 1/N + 1/log_2{Q} + 4))$ space.
 
-Assuming that we want to build a graph, e.g., 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx O(\num{e15})$ identifiers to store the full model.
+Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
 Such a design is almost infeasible for inputs larger than a handful of genomes.
-For instance, we would need $O(\num{3e12})$ identifiers for just 5 human genomes.
+For instance, we would need $\approx \num{3e12}$ identifiers for just 5 human genomes.
 Although it is feasible to compute such a graph using external memory, the approximately 200-fold increase in space relative to the input renders this clearly impractical.
@@
         \textit{A. thaliana} & 922 & 16 & 1.90251 & 49 & 468 & 43.1287 & 7.1218 & 0.234284 & 100 \\ \hline
-        \textit{H. sapiens} & 17478 & 38 & 114.627 & 49 & 39723 & 347.9261 & 464.3392 & 6.2322 & 724 \\ \hline
-        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 475 & 33.7270 & 11.9739 & 0.0302351 & 33 \\ \hline
+        \textit{H. sapiens} & 17478 & 38 & 114.627 & 49 & 46268 & 347.4983 & 604.4261 & 4.47126 & 474 \\ \hline
+        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 790 & 75.7514 & 20.2070 & 0.01421 & 5 \\ \hline
         \textit{Z. mays} & 46289 & 41 & 90.2491 & 49 & 31043 & 351.1235 & 402.8716 & 13.8838 & 925 \\ \hline

### 2022-02-07 manuscript/main.tex
tweaks

@@
 \textbf{Motivation:}
-Pangenome graphs, often formally encoded as variation graphs, model the mutual alignment of collections of genomes.
-A set of pairwise alignments implies a variation graph, but there are no scalable methods to generate such a graph or lossless string graph from these alignments.
-Existing approaches depend on a single reference or a specific ordering of genomes, and do not provide base-level information about the mapping of genomes into the pangenome.
-A scalable method to build pangenome graphs without such limitations would be a key step in pangenome construction and manipulation pipelines. \\
+Pangenome variation graphs model the mutual alignment of collections of DNA sequences.
+A set of pairwise alignments implies a variation graph, but there are no scalable methods to generate such a graph from these alignments.
+Existing related approaches depend on a single reference, a specific ordering of genomes, and do not provide base-level information about the mapping of genomes into the pangenome.
+A scalable, self-contained method to build pangenome graphs without such limitations would be a key step in pangenome construction and manipulation pipelines. \\
 \textbf{Results:}
@@
 Existing methods approximate this relationship by progressive alignment to a graph initially based on a reference genome \citep{Li:2020}, or through a global structuring of the genome relationships in a neighbor joining phylogenetic tree \citep{Armstrong:2020}.
-These methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, and guide-tree topology.
+These methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, or guide-tree topology.
 
@@
 %Assume that $i$ and $j$ are coordinates in $Q$.
-A match $m = (i, j)$ asserts the evolutionary equivalence of two characters in sequences in $Q$.
+A match $m = (i, j)$ asserts the aligned equivalence of two characters in sequences in $Q$.
 %$\forall_{(i, j) : j = |g| - i} g[i] = \bar{g}[j]$.

### 2022-02-10 manuscript/main.tex
add algorithm pseudocode

@@
 \newtheorem{lemma}[theorem]{Lemma}
+
+\usepackage{algorithm2e}
+\SetKwRepeat{Do}{do}{while}%
 
@@
 We present an implementation that operates in external memory, using disk-backed data structures and a lock-free parallel algorithm to drive the core graph induction step.
-We demonstrate that our method scales to very large graph induction problems, and apply our work to build graphs from dozens of genomes from the same or different species in clades throughout the tree of life.
-\\
+We demonstrate that our method scales to very large graph induction problems, and apply our work to build graphs from dozens of genomes from species in clades throughout the tree of life. \\
 \textbf{Availability:}
@@
 
+\input{seqwish_pseudocode.tex}
+
 \subsection{Node compaction}
@@
 
-\FIXME{For reasons of computational efficiency, node identifiers are numeric.}
+%\FIXME{For reasons of computational efficiency, node identifiers are numeric.}
 

### 2022-02-10 manuscript/fig_experiments.tex
l/m swap in the pseudocode and tweaks

@@
 %  \centering
-   \includegraphics[width=\linewidth]{fig_experiment_stats.pdf}
+   \includegraphics[width=\linewidth, trim=0.4cm 0cm 0.4cm 1cm]{fig_experiment_stats}
    \caption{
      Experimental results from the application of \textit{seqwish} to four different pangenomes.
-     Multiple minimum identity settings for the mapping (\textit{wfmash -p}) and different minimum match length filter settings result in a collection of graph builds per pangenome input.
+     Multiple minimum identity settings for the mapping (\textit{wfmash -p}) and different minimum match length filter settings (\textit{seqwish -k}) result in a collection of graph builds per pangenome input.
      We compare the runtime in hours with the size of the resulting graphs relative to the average size of an input genome in the particular set.
@@
      Similarly, filtering short matches increases graph size relative to not (\textit{seqwish -k}=0, red).
-     Increasing the \textit{seqwish -k} parameter tends to increase the size of the graph.
+     The other way around, increasing the \textit{seqwish -k} parameter tends to increase the size of the graph.
     }

### 2022-02-10 manuscript/main.tex
l/m swap in the pseudocode and tweaks

@@
     \caption{Performance of the graph induction algorithm.
-        For each pangenome we have chosen a single experiment with the given \textit{seqwish -k} filter setting of 49bp.
+        For each pangenome we report a single experiment with the given \textit{seqwish -k} filter setting of 49bp.
         }
@@
 
-Code and links to data resources used to build this manuscript and its figures, can be found in the paper's public repository: \url{https://github.com/pangenome/seqwish-paper}.
+Code and links to data resources used to build this manuscript and its figures can be found in the paper's public repository: \url{https://github.com/pangenome/seqwish-paper}.
 

### 2022-02-11 manuscript/main.tex
might be ready

@@
 A set of pairwise alignments implies a variation graph, but there are no scalable methods to generate such a graph from these alignments.
-Existing related approaches depend on a single reference, a specific ordering of genomes, and do not provide base-level information about the mapping of genomes into the pangenome.
+Existing related approaches depend on a single reference, a specific ordering of genomes, a de Bruijn model, and may not describe the base-level mapping of genomes into the pangenome.
 A scalable, self-contained method to build pangenome graphs without such limitations would be a key step in pangenome construction and manipulation pipelines. \\
@@
 An unbiased pangenome graph would represent the alignment of all included genomes to all others.
-Existing methods approximate this relationship by progressive alignment to a graph initially based on a reference genome \citep{Li:2020}, or through a global structuring of the genome relationships in a neighbor joining phylogenetic tree \citep{Armstrong:2020}.
-These methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, or guide-tree topology.
+Existing methods approximate this relationship by progressive alignment to a graph initially based on a reference genome \citep{Li:2020}, through a global structuring of the genome relationships in a neighbor joining phylogenetic tree \citep{Armstrong:2020}, or via creation of a \textit{de Bruijn} graph based on a fixed $k$-mer length \citep{Minkin_2016,Yu_2021}.
+These methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, guide-tree topology, or $k$-mer length.
 
@@
 %Any position in $\mathcal{G}$ so marked must be the start or end of a node.
-We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ such that $|n_i| \ge 1 \forall_{i \in 1 \ldots |\mathcal{N}'|}$.
+We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ where compacted node boundaries are marked in a auxilary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{N}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
+$\mathcal{B}$ allows us to compute compacted node ids using efficient rank operation \citep{Gog_2014}.
+%such that $\forall_{i \in 1 \ldots |\mathcal{N}'|} |n_i| \ge 1$.
 
@@
 \subsection{Induction partitioning}
+\label{sec:partitioning}
 
@@
 \subsection{Short match filter}
+\label{sec:filter}
 
@@
 \label{sec:results}
-\FIXME{to do}
+
+We evaluate \textit{seqwish} through application to four pangenomes collected from \textit{A. thaliana}, \textit{H. sapiens}, \textit{H. pylori}, and \textit{Z. mays}.
+This limited survey is intended to demonstrate basic scaling properties of the method, and its practicality when applied to real pangenomes.
+We also consider the effect of the minimum match length filter described in section \ref{sec:filter}.
+Experiments were conducted on compute nodes with 386GB of RAM and AMD EPYC 7402P processors with 48 vCPUs.
+
+%\subsection{Alignment and graph induction pipeline}
+
+To construct the graph we first generate alignments with \textit{wfmash} \citep{wfmash}, which is designed specifically for high performance all-to-all alignment of fully-assembled genomes.
+\textit{wfmash} combines an algorithm for generating whole-genome homology maps (\textit{MashMap2}) \citep{Jain_2018} with an extension of the wavefront algorithm (WFA) \citep{Marco_Sola_2020} capable of obtaining base-level alignments for whole chromosomes.
+\textit{MashMap2} allows the user to define a homology length and pairwise divergence, expressed as a percent identity, over which to generate homology maps.
+This is useful when constructing pagnenome graphs, because, in contrast to methods that are based on $k$-mer chaining \citep{Li_2018}, it allows us to query the homology space of input genomes using two easily-interpretable parameters.
+The version of \textit{wfmash} used in these experiments allows us to align sequences with up to 10\% divergence between them, providing highly-sensitive input for our experiments.
+
+%\subsection{Experimental insights}
+
+In table \ref{tab:experiments} we provide input and construced graph parameters for a single parameter setting of \textit{wfmash} and \textit{seqwish}, obtaining graph parameters with the ODGI toolkit \citep{Guarracino_2021}.
+Figure \ref{fig:experiments} provides displays runtime versus graph size relative to the average input assembly length across the range of parameters chosen for each pangenome.
+These provide a consistent set of insights.
+Reducing the sensitivity of alignments by increasing the identity threshold results in a larger graph.
+Filtering short matches results in a larger graph, and for higher divergence collections like \textit{H. pylori} tends to obliterate much of the homology information in the pangenome graph.
+Reducing the size of $\mathcal{A}$ using these changes also greatly reduces runtime.
+In all cases, we find that the initial alignment step takes longer than graph induction.
+
+%\subsection{Limitations}
+
+Although we use disk-backed data structures to represent the graph, the maximum memory requirements of \textit{seqwish} are governed by the largest transitive match closure in the pangenome graph.
+We find that our particular partitioning scheme---we compact the graph in chunks as described in \ref{sec:partitioning}, using 50 Mbp chunks in these experiments---does not allow us to complete the graph induction for $p=95$ and $k=0$ for the \textit{H.
+sapiens} set, nor for \textit{Z.
+mays} with $k < 49$, where we run out of working memory.
+In practice, setting the chunk size lower tends to resolve this problem, but will also increase runtime.
+To simplify comparisons between the different parameter settings, we have not re-run these settings with a different partition size.
 
@@
         }
+    \label{tab:experiments}
     \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
@@
 \label{sec:discussion}
-\FIXME{to do}
 
-We can apply our method to various collections of sequences and alignments.
+We have presented a straightforward algorithm to generate a pangenome graph from a collection of genomes and alignments between them.
+By exploiting a simple model of this algorithm, we provide computational bounds that give insight into the complexity of the problem.
+We then make this approach practical by applying the concept of \textit{match compression}, which reduces the expected computational complexity by a factor proportional to the diversity of input sequences.
+Our experimental results domenstrate that we can apply our method to various collections of sequences and alignments.
 It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
+%Although we have focused on the problem of building pangenomes,
+\textit{seqwish} is a generic sequence graph inducer of potentially many uses.
+We envision that it can serve as a component in diverse sequence analysis and assembly pipelines, and hope that our thorough description of its core algorithm and functionality will enable its reuse by other researchers.
 
-However, this exposition comes with significant limitations.
-In our experiments, we have not explored the full problem, including alignment.
-These lie outside of the scope of the presented work, wherein whe have focused on a key kernel which is a bottleneck in the pangenome graph construction process.
-Instead, \textit{seqwish} is a generic ``graph inducer'' of potentially many uses.
-But, it is important to note that the graphs which it produces are simply raw representations of an underlying alignment set.
-Due to differences in alignments that arise in regions of lower entropy, in practice, it is necessary to normalize generated graphs to simplify the local alignment structures represented in the graph.
+It is also a potentially novel approach.
+Despite the existence of many methods for pangenome building, we are not aware of any comparable method which can losslessly convert an all-to-all alignment to a variation graph.
+This direct relationship allows users to adjust the shape of the resulting graph by modifying alignment parameters, allowing the design of custom graph construction processes based on domain-specific knowledge and potentially manual curation of assembly alignments.
+In contrast to existing methods, which depend on particular structuring of their input \citep{Li:2020,Armstrong:2020}, \textit{seqwish} is unbiased in that it directly and uniformly represents sequence relationships given on input in the resulting graph.
+Although we did not compare with \textit{de Bruijn} graph methods, which are also unbiased in their uniform treatment of input genomes \citep{Minkin_2016,Yu_2021}, we believe such methods are fundamentally different in that they collapse all exact matches of a given length $k$.
+This prevents high-level structuring of patterns of homology and orthology in the resulting graph, and they furthermore cannot be guided by a specific alignment set.
+
+Our presentation is necessarily limited, so as to focus on and describe the unique problem of variation graph induction.
+Thus, in this manuscript and our experiments, we have not explored the full problem of \textit{pangenome graph building}, which include both the initial alignment step and downstream processing of the resulting graph.
+These topics lie outside of the scope of the presented work, wherein whe have focused on a key kernel which is a bottleneck in the pangenome graph construction process.
+But, they are important for readers to consider.
+Although \textit{seqwish} perfectly represents its input alignments, the problem of generating and filtering an alignment set remains critical, as it determines the structure of the built graph.
+And this lossless property does not guarantee that the resulting graph is easy to work with or navigate---in practice find that downsteam processing is required to normalize the graph for many applications.
+We will cover these topics in future work.
+
+%However, users of \textit{seqwish} should note that although the graph 
+%But, it is important to note that the graphs which it produces are simply raw representations of an underlying alignment set.
+%Due to differences in alignments that arise in regions of lower entropy, in practice, it is necessary to normalize generated graphs to simplify the local alignment structures represented in the graph.
 
@@
 
-%We thank Richard Durbin, Benedict Paten, and Simon Puglisi for support and helpful discussions in the development of this algorithm.
-We are grateful to members of the HPRC production teams for their development of resources used in our exposition.
+We are grateful to members of the HGSVC and HPRC production teams for their development of resources used in our exposition.
+We thank the authors of the pangenome resources made available on GenBank which have made our experiments possible.
 
@@
 
-We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG).
+We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG) and efforts by Dr. Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy (AG).
 \linebreak

### 2022-02-11 manuscript/main.tex
typo

@@
 A set of pairwise alignments implies a variation graph, but there are no scalable methods to generate such a graph from these alignments.
-Existing related approaches depend on a single reference, a specific ordering of genomes, a de Bruijn model, and may not describe the base-level mapping of genomes into the pangenome.
+Existing related approaches depend on a single reference, a specific ordering of genomes, or a \textit{de Bruijn} model, and may not describe the base-level mapping of genomes into the pangenome.
 A scalable, self-contained method to build pangenome graphs without such limitations would be a key step in pangenome construction and manipulation pipelines. \\

### 2022-02-11 manuscript/main.tex
tweaks everywhere; be a bit more verbose

--- 2022-02-11 manuscript/main.tex
@@
 A set of pairwise alignments implies a variation graph, but there are no scalable methods to generate such a graph from these alignments.
-Existing related approaches depend on a single reference, a specific ordering of genomes, a de Bruijn model, and may not describe the base-level mapping of genomes into the pangenome.
+Existing related approaches depend on a single reference, a specific ordering of genomes, or a \textit{de Bruijn} model, and may not describe the base-level alignment of genomes into the pangenome.
 A scalable, self-contained method to build pangenome graphs without such limitations would be a key step in pangenome construction and manipulation pipelines. \\
@@
 
-Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
+Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bps, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
 Such a design is almost infeasible for inputs larger than a handful of genomes.
@@
 We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ where compacted node boundaries are marked in a auxilary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{N}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
-$\mathcal{B}$ allows us to compute compacted node ids using efficient rank operation \citep{Gog_2014}.
+$\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.
 %such that $\forall_{i \in 1 \ldots |\mathcal{N}'|} |n_i| \ge 1$.
@@
 
-To construct the graph we first generate alignments with \textit{wfmash} \citep{wfmash}, which is designed specifically for high performance all-to-all alignment of fully-assembled genomes.
+To construct the graph we first generate alignments with \textit{wfmash} \citep{wfmash}, a DNA sequence aligner designed specifically for high performance all-to-all alignment of fully-assembled genomes.
 \textit{wfmash} combines an algorithm for generating whole-genome homology maps (\textit{MashMap2}) \citep{Jain_2018} with an extension of the wavefront algorithm (WFA) \citep{Marco_Sola_2020} capable of obtaining base-level alignments for whole chromosomes.
 \textit{MashMap2} allows the user to define a homology length and pairwise divergence, expressed as a percent identity, over which to generate homology maps.
-This is useful when constructing pagnenome graphs, because, in contrast to methods that are based on $k$-mer chaining \citep{Li_2018}, it allows us to query the homology space of input genomes using two easily-interpretable parameters.
+This is useful when constructing pangenome graphs, because, in contrast to methods that are based on $k$-mer chaining \citep{Li_2018}, it allows us to query the homology space of input genomes using two easily-interpretable parameters.
 The version of \textit{wfmash} used in these experiments allows us to align sequences with up to 10\% divergence between them, providing highly-sensitive input for our experiments.
@@
 
-In table \ref{tab:experiments} we provide input and construced graph parameters for a single parameter setting of \textit{wfmash} and \textit{seqwish}, obtaining graph parameters with the ODGI toolkit \citep{Guarracino_2021}.
-Figure \ref{fig:experiments} provides displays runtime versus graph size relative to the average input assembly length across the range of parameters chosen for each pangenome.
+In table \ref{tab:experiments} we provide input and constructed graph parameters for a single parameter setting of \textit{wfmash} and \textit{seqwish}, obtaining graph statistics with the ODGI toolkit \citep{Guarracino_2021}.
+Figure \ref{fig:experiments} displays runtime versus graph size relative to the average input genome length across the range of parameters chosen for each pangenome.
 These provide a consistent set of insights.
-Reducing the sensitivity of alignments by increasing the identity threshold results in a larger graph.
-Filtering short matches results in a larger graph, and for higher divergence collections like \textit{H. pylori} tends to obliterate much of the homology information in the pangenome graph.
-Reducing the size of $\mathcal{A}$ using these changes also greatly reduces runtime.
+Reducing the sensitivity of alignments by increasing the identity threshold results in larger graphs.
+Filtering short matches results in larger graphs too, and for higher divergence collections of genomes, like \textit{H.
+pylori}, tends to obliterate much of the homology information in the pangenome graph.
+This is visible from the fact that the "graph length / average genome length" ratio grows strongly as $k$ increases.
+Such a reduction of the size of the set of matches considered for graph inductions also greatly reduces runtime.
 In all cases, we find that the initial alignment step takes longer than graph induction.
@@
 Although we use disk-backed data structures to represent the graph, the maximum memory requirements of \textit{seqwish} are governed by the largest transitive match closure in the pangenome graph.
-We find that our particular partitioning scheme---we compact the graph in chunks as described in \ref{sec:partitioning}, using 50 Mbp chunks in these experiments---does not allow us to complete the graph induction for $p=95$ and $k=0$ for the \textit{H.
+We find that our particular partitioning scheme (we compact the graph in chunks as described in~\ref{sec:partitioning}, using 50 Mbps chunks in all the experiments except for \textit{H.
+pylori} where we used chunks 10 Mbps long) does not allow us to complete the graph induction for $p=95$ and $k=0$ for the \textit{H.
 sapiens} set, nor for \textit{Z.
-mays} with $k < 49$, where we run out of working memory.
+mays} with $k \leq 29$, where we run out of working memory.
 In practice, setting the chunk size lower tends to resolve this problem, but will also increase runtime.
-To simplify comparisons between the different parameter settings, we have not re-run these settings with a different partition size.
+To simplify comparisons between the different parameter settings, we have not re-run these settings with a different partition size nor on computer nodes with a larger, then different, amount of RAM.
 
@@
     \caption{Performance of the graph induction algorithm.
-        For each pangenome we report a single experiment with the given \textit{seqwish -k} filter setting of 49bp.
+        For each pangenome we report a single experiment with \textit{seqwish -k} filter set to 49bps.
+        From left to right, the columns indicate the species, the number of sequences (that is, the number of contigs), number of haplotypes (that is, number of individuals), the sum of the length of all sequences in Gbps, the length of the short match filter applied in bps, the time in seconds and the amount of memory and disk space in Gbytes required for the graph induction, the length of the resulting graph in Gbps and the number its connected components.
         }
@@
         \hline
-        species & sequences & haplotypes & fasta.Gbps & min.match.bp & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbps & components \\ \hline
+        species & sequences & haplotypes & fasta.Gbps & min.match.bps & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbps & components \\ \hline
         \textit{A. thaliana} & 922 & 16 & 1.90251 & 49 & 468 & 43.1287 & 7.1218 & 0.234284 & 100 \\ \hline
@@
 We then make this approach practical by applying the concept of \textit{match compression}, which reduces the expected computational complexity by a factor proportional to the diversity of input sequences.
-Our experimental results domenstrate that we can apply our method to various collections of sequences and alignments.
+Our experimental results demenstrate that we can apply our method to various collections of sequences and alignments.
 It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
@@
 This direct relationship allows users to adjust the shape of the resulting graph by modifying alignment parameters, allowing the design of custom graph construction processes based on domain-specific knowledge and potentially manual curation of assembly alignments.
-In contrast to existing methods, which depend on particular structuring of their input \citep{Li:2020,Armstrong:2020}, \textit{seqwish} is unbiased in that it directly and uniformly represents sequence relationships given on input in the resulting graph.
-Although we did not compare with \textit{de Bruijn} graph methods, which are also unbiased in their uniform treatment of input genomes \citep{Minkin_2016,Yu_2021}, we believe such methods are fundamentally different in that they collapse all exact matches of a given length $k$.
+In contrast to existing methods, which depend on particular structuring of their input \citep{Li:2020,Armstrong:2020}, \textit{seqwish} is unbiased in that as it directly and uniformly represents sequence relationships given on input in the resulting graph.
+Although we did not compare with \textit{de Bruijn} graph methods, which are also unbiased in their uniform treatment of input genomes \citep{Minkin_2016,Yu_2021}, we believe such methods are fundamentally different in that because they collapse all exact matches of a given length $k$.
 This prevents high-level structuring of patterns of homology and orthology in the resulting graph, and they furthermore cannot be guided by a specific alignment set.
 
-Our presentation is necessarily limited, so as to focus on and describe the unique problem of variation graph induction.
+Our presentation is necessarily limited, in order to focus on and describe the unique problem of variation graph induction.
 Thus, in this manuscript and our experiments, we have not explored the full problem of \textit{pangenome graph building}, which include both the initial alignment step and downstream processing of the resulting graph.
-These topics lie outside of the scope of the presented work, wherein whe have focused on a key kernel which is a bottleneck in the pangenome graph construction process.
+These topics lie outside of the scope of the presented work, wherein we have focused on a key kernel which is a bottleneck in the pangenome graph construction process.
 But, they are important for readers to consider.
 Although \textit{seqwish} perfectly represents its input alignments, the problem of generating and filtering an alignment set remains critical, as it determines the structure of the built graph.
-And this lossless property does not guarantee that the resulting graph is easy to work with or navigate---in practice find that downsteam processing is required to normalize the graph for many applications.
+And this lossless property does not guarantee that the resulting graph is easy to work with or navigate; in practice, downstream processing is usually required to normalize the graph for many applications.
 We will cover these topics in future work.

### 2022-02-11 scripts/experiment_plot.R

@@ -9,7 +9,7 @@ ggplot(subset(x, input.fasta=="athaliana16" & min.mapping.len=="l60k" | input.fa
        aes(x=graph.length.Gbps/(Gbps/n), y=time.seconds/3600, shape=map.ident, color=as.factor(k), label=min.mapping.len)) +
 geom_point() +
 facet_wrap(~input.fasta, scales="free") +
-scale_x_continuous("graph length / single genome length") +
+scale_x_continuous("graph length / average genome length") +
 scale_y_continuous("runtime (hours)") +
 scale_color_discrete("seqwish -k") +
 scale_shape_discrete("wfmash -p")

### 2022-02-11 manuscript/main.tex
fix links and add ORCIDs
update hpylori induction and simplify the text

@@
 
-\usepackage[draft]{hyperref}
 \usepackage{makecell}
-\usepackage{comment}
 
@@
 
-\usepackage{algorithm2e}
 \usepackage[usenames,dvipsnames]{xcolor}
@@
 
+\usepackage{orcidlink}
+\hypersetup{hidelinks}
+
 \def\labelitemi{--}
@@
 \author[Garrison \textit{et~al}.]{
-Erik~Garrison\,$^{\text{\sfb 1}*}$,
-Andrea~Guarracino\,$^{\text{\sfb 2}}$
+Erik~Garrison\,$^{\orcidlink{0000-0003-3821-631X}\text{\sfb 1}*}$,
+Andrea~Guarracino\,$^{\orcidlink{0000-0001-9744-131X}\text{\sfb 2}}$
 }
@@
 Although we use disk-backed data structures to represent the graph, the maximum memory requirements of \textit{seqwish} are governed by the largest transitive match closure in the pangenome graph.
-We find that our particular partitioning scheme (we compact the graph in chunks as described in~\ref{sec:partitioning}, using 50 Mbps chunks in all the experiments except for \textit{H.
-pylori} where we used chunks 10 Mbps long) does not allow us to complete the graph induction for $p=95$ and $k=0$ for the \textit{H.
+We find that our particular partitioning scheme (we compact the graph in chunks as described in~\ref{sec:partitioning}, using 50 Mbps chunks in all the experiments) does not allow us to complete the graph induction for $p=95$ and $k=0$ for the \textit{H.
 sapiens} set, nor for \textit{Z.
@@
         \textit{H. sapiens} & 17478 & 38 & 114.627 & 49 & 46268 & 347.4983 & 604.4261 & 4.47126 & 474 \\ \hline
-        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 790 & 75.7514 & 20.2070 & 0.01421 & 5 \\ \hline
+        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 777 & 74.9484 & 20.2070 & 0.01421 & 5 \\ \hline
         \textit{Z. mays} & 46289 & 41 & 90.2491 & 49 & 31043 & 351.1235 & 402.8716 & 13.8838 & 925 \\ \hline

### 2022-02-14 manuscript/main.tex
tweaks

--- 2022-02-11 manuscript/main.tex
@@
 \textit{MashMap2} allows the user to define a homology length and pairwise divergence, expressed as a percent identity, over which to generate homology maps.
-This is useful when constructing pangenome graphs, because, in contrast to methods that are based on $k$-mer chaining \citep{Li_2018}, it allows us to query the homology space of input genomes using two easily-interpretable parameters.
+This is useful when constructing pangenome graphs, because, in contrast to methods that are based on $k$-mer chaining \citep{harris2007lastz,Li_2018}, it allows us to query the homology space of input genomes using two easily-interpretable parameters.
 The version of \textit{wfmash} used in these experiments allows us to align sequences with up to 10\% divergence between them, providing highly-sensitive input for our experiments.

### 2022-02-14 manuscript/main.tex
stuff

@@
 A set of pairwise alignments implies a variation graph, but there are no scalable methods to generate such a graph from these alignments.
-Existing related approaches depend on a single reference, a specific ordering of genomes, or a \textit{de Bruijn} model, and may not describe the base-level alignment of genomes into the pangenome.
+Existing related approaches depend on a single reference, a specific ordering of genomes, or a \textit{de Bruijn} model based on a fixed $k$-mer length.
+%, and may not describe the base-level alignment of genomes into the pangenome.
 A scalable, self-contained method to build pangenome graphs without such limitations would be a key step in pangenome construction and manipulation pipelines. \\
@@
 To build up the variation graph, we query this tree-based representation of the alignments to reduce transitive matches into single DNA segments in a sequence graph.
-By recording the mapping from input sequence to output graph, we can trace the original paths through this graph, yielding a pangenome graph.
+By recording the mapping from input sequence to output graph, we can trace the original paths through this graph, yielding a pangenome variation graph.
 We present an implementation that operates in external memory, using disk-backed data structures and a lock-free parallel algorithm to drive the core graph induction step.
-We demonstrate that our method scales to very large graph induction problems, and apply our work to build graphs from dozens of genomes from species in clades throughout the tree of life. \\
+We demonstrate that our method scales to very large graph induction problems by applying it to build pangenome graphs for several species. \\
 \textbf{Availability:}
-seqwish is published as free software under the MIT open source license.
+\textit{seqwish} is published as free software under the MIT open source license.
 Source code and documentation are available at \url{https://github.com/ekg/seqwish}.
-seqwish can be installed via Bioconda \url{https://bioconda.github.io/recipes/seqwish/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/seqwish.scm}. \\
+\textit{seqwish} can be installed via Bioconda \url{https://bioconda.github.io/recipes/seqwish/README.html} or GNU Guix \url{https://github.com/ekg/guix-genomics/blob/master/seqwish.scm}. \\
 \textbf{Contact:} \href{egarris5@uthsc.edu}{egarris5@uthsc.edu} \\
@@
 %, and may provide a simpler representation of structurally-variable regions.
-Many approaches model the pangenome alignment as a \textit{pangenome graph} \cite{Garrison_2018,Yokoyama2019,Hickey:2020}.
+Many approaches model the pangenome alignment as a \textit{pangenome graph} \citep{Garrison_2018,Yokoyama2019,Hickey:2020}.
 A pangenome graph encodes DNA sequences as walks through an underlying language encoded in a sequence graph \citep{Hein_1989}.
@@
 
-Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bps, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
+Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
 Such a design is almost infeasible for inputs larger than a handful of genomes.
@@
 Although we use disk-backed data structures to represent the graph, the maximum memory requirements of \textit{seqwish} are governed by the largest transitive match closure in the pangenome graph.
-We find that our particular partitioning scheme (we compact the graph in chunks as described in~\ref{sec:partitioning}, using 50 Mbps chunks in all the experiments) does not allow us to complete the graph induction for $p=95$ and $k=0$ for the \textit{H.
+We find that our particular partitioning scheme (we compact the graph in chunks as described in~\ref{sec:partitioning}, using 50 Mbp chunks in all the experiments) does not allow us to complete the graph induction for $p=95$ and $k=0$ for the \textit{H.
 sapiens} set, nor for \textit{Z.
@@
     \caption{Performance of the graph induction algorithm.
-        For each pangenome we report a single experiment with \textit{seqwish -k} filter set to 49bps.
-        From left to right, the columns indicate the species, the number of sequences (that is, the number of contigs), number of haplotypes (that is, number of individuals), the sum of the length of all sequences in Gbps, the length of the short match filter applied in bps, the time in seconds and the amount of memory and disk space in Gbytes required for the graph induction, the length of the resulting graph in Gbps and the number its connected components.
+        For each pangenome we report a single experiment with \textit{seqwish -k} filter set to 49bp.
+        From left to right, the columns indicate the species, the number of sequences (that is, the number of contigs), number of haplotypes (that is, number of individuals), the sum of the length of all sequences in Gbp, the length of the short match filter applied in bp, the time in seconds and the amount of memory and disk space in Gbytes required for the graph induction, the length of the resulting graph in Gbp and the number its connected components.
         }
@@
         \hline
-        species & sequences & haplotypes & fasta.Gbps & min.match.bps & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbps & components \\ \hline
+        species & sequences & haplotypes & fasta.Gbp & min.match.bp & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbp & components \\ \hline
         \textit{A. thaliana} & 922 & 16 & 1.90251 & 49 & 468 & 43.1287 & 7.1218 & 0.234284 & 100 \\ \hline

### 2022-02-14 manuscript/main.tex
add reference to union-find algorithm

@@
 We do so by computing the graph induction for a collection of initial characters in $Q$.
-We induce the related graph for each partition, appending the induction results to the appropriate data structures.
+%We induce the related graph for each partition, appending the induction results to the appropriate data structures.
+In each partition, we apply a lock-free parallel union-find algorithm to derive the match closures \citep{anderson1991wait}, appending results to appropriate data structures.
 This partitioning can introduce boundary effects which change the contents of $\mathcal{Z}$ and $\bar{\mathcal{Z}}$ by splitting ranges at the boundaries of our partitions.
 However, while this will affect the compressed node definition $\mathcal{N}'$, it does not affect $\mathcal{N}$, and it can be corrected via a post-processing step to sort and compact the id space.
-Partitioning reduces the working memory limits to be in the order of the size of the largest transitive match closure found in the input.
+%Partitioning reduces the working memory limits to be in the order of the size of the largest transitive match closure found in the input.
 

### 2022-02-14 manuscript/main.tex
stuff

@@
 By recording the mapping from input sequence to output graph, we can trace the original paths through this graph, yielding a pangenome variation graph.
-We present an implementation that operates in external memory, using disk-backed data structures and a lock-free parallel algorithm to drive the core graph induction step.
+We present an implementation that operates in external memory, using disk-backed data structures and lock-free parallel methods to drive the core graph induction step.
 We demonstrate that our method scales to very large graph induction problems by applying it to build pangenome graphs for several species. \\

### 2022-02-14 manuscript/main.tex
mention algorithm 1
spell check

@@
 We then examine the bounds of a na\"{i}ve implementation of this algorithm.
-Finally, we propose compression and partitioning techniques to reduce the space and working memory complexity of the induction process by a large constant factor modulated by the degree of sequence divengence in the input pangenome.
+Finally, we propose compression and partitioning techniques to reduce the space and working memory complexity of the induction process by a large constant factor modulated by the degree of sequence divergence in the input pangenome.
 This yields a practical algorithm for variation graph induction that can scale to the largest available pangenomes.
@@
 Additionally, we store the inverse of $\mathcal{Z}$, which maps ranges from $\mathcal{N} \to Q$, as $\bar{\mathcal{Z}}$.
-We use $\mathcal{Z}$ to compact unbranching regions of $\mathcal{V}$ into single nodes, and $\bar{\mathcal{Z}}$ to accelerate our calculation of links in the graph.
+We use $\mathcal{Z}$ to compact non-branching regions of $\mathcal{V}$ into single nodes, and $\bar{\mathcal{Z}}$ to accelerate our calculation of links in the graph.
 
@@
 %Any position in $\mathcal{G}$ so marked must be the start or end of a node.
-We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ where compacted node boundaries are marked in a auxilary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{N}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
+We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{N}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
 $\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.
 %such that $\forall_{i \in 1 \ldots |\mathcal{N}'|} |n_i| \ge 1$.
+%In Algorithm \ref{alg:seqwish} we provide a detailed description of the graph induction algorithm.
 
@@
 
-We have presented a complete model for variation graph induction from sequences and their pairwise alignments.
+We have presented a complete model for variation graph induction from sequences and their pairwise alignments (Algorithm \ref{alg:seqwish}).
 Here, we describe our specific implementation of this algorithm: \textit{seqwish}.
@@
 We then make this approach practical by applying the concept of \textit{match compression}, which reduces the expected computational complexity by a factor proportional to the diversity of input sequences.
-Our experimental results demenstrate that we can apply our method to various collections of sequences and alignments.
+Our experimental results demonstrate that we can apply our method to various collections of sequences and alignments.
 It easily scales to some of the largest species pangenome construction problems possible using publicly-available, high-quality genome assemblies.
@@
 
-\section*{Acknowledgements}
+\section*{Acknowledgments}
 

### 2022-04-01 manuscript/main.tex
fix affiliations

@@
 \address{
-$^{\text{\sf 1}}$University of Tennessee Health Science Center, Memphis, TN, USA \\
-$^{\text{\sf 2}}$Genomics Research Centre, Human Technopole, Milan, Italy
+$^{\text{\sf 1}}$Department of Genetics, Genomics and Informatics, University of Tennessee Health Science Center, Memphis, 38163, Tennessee, USA \\
+$^{\text{\sf 2}}$Genomics Research Centre, Human Technopole, Viale Rita Levi‑Montalcini 1, Milan, 20157, Italy
 }

### 2022-05-15 manuscript/main.tex
nsf ppos funding
update ODGI citation

@@
 
-In table \ref{tab:experiments} we provide input and constructed graph parameters for a single parameter setting of \textit{wfmash} and \textit{seqwish}, obtaining graph statistics with the ODGI toolkit \citep{Guarracino_2021}.
+In table \ref{tab:experiments} we provide input and constructed graph parameters for a single parameter setting of \textit{wfmash} and \textit{seqwish}, obtaining graph statistics with the ODGI toolkit \citep{Guarracino2022}.
 Figure \ref{fig:experiments} displays runtime versus graph size relative to the average input genome length across the range of parameters chosen for each pangenome.

### 2022-08-08 manuscript/main.tex
Replace `\mathcal{P}` with `|P|`; specify that we refer to the size of the array; add missing backslash

@@
 We first sort the sequence-to-graph mapping array $Z$ in $O(\mathcal{Q} \log \mathcal{Q})$ operations, and then compute $\mathcal{P}$ in $O(\mathcal{Q})$ queries which each cost $O(\log \mathcal{Q})$.
-To obtain unique edges and generate $\mathcal{E}$, we must build and sort an array of $O(2\mathcal{P})$, and then iterate through it for $O(2\mathcal{P} \log 2\mathcal{P} + 2\mathcal{P})$ operations.
-In sum, we would expect to require $O(|A|\log|A| + \mathcal{N} \log |A| + 2(\mathcal{Q} \log \mathcal{Q}) + 2(\mathcal{P} \log 2\mathcal{P} + \mathcal{P}))$.
-Using our approximate relationships to $\mathcal{Q}$ given previously and simplifying, we arrive at $O[\mathcal{Q}(2N^2 \log N / \log \mathcal{Q} + (1/N + 4)\log \mathcal{Q} + 2 + log(4))]$.
+To obtain unique edges and generate $\mathcal{E}$, we must build and sort an array of size $O(2|P|)$, and then iterate through it for $O(2|P| \log 2|P| + 2|P|)$ operations.
+In sum, we would expect to require $O(|A|\log|A| + \mathcal{N} \log |A| + 2(\mathcal{Q} \log \mathcal{Q}) + 2(|P| \log 2|P| + |P|))$.
+Using our approximate relationships to $\mathcal{Q}$ given previously and simplifying, we arrive at $O[\mathcal{Q}(2N^2 \log N / \log \mathcal{Q} + (1/N + 4)\log \mathcal{Q} + 2 + \log(4))]$.
 
@@
 
-We gratefully acknowledge support from NIH/NIDA U01DA047638 (EG) and efforts by Dr. Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy (AG).
+We gratefully acknowledge support from NIH/NIDA U01DA047638 and NSF PPoSS Award \#2118709 (EG) and efforts by Dr.
+Nicole Soranzo to establish a pangenome research unit at the Human Technopole in Milan, Italy (AG).
 \linebreak

### 2022-08-08 manuscript/main.tex
replace `is` with `may be`
canonical graph notation

@@
 Variation graphs are a common formalism to encode pangenome graphs \citep{Garrison_2019_thesis}.
-In the variation graph $\mathcal{V} = (\mathcal{N}, \mathcal{E}, \mathcal{P})$, nodes $\mathcal{N} = n_1\ldots n_{|\mathcal{N}|}$ contain sequences of DNA.
-Each node $n_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{n_i}$.
+In the variation graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{P})$, nodes (or vertices) $\mathcal{V} = v_1\ldots v_{|\mathcal{V}|}$ contain sequences of DNA.
+Each node $v_i$ has a unique identifier $i$ and an implicit reverse complement $\bar{v_i}$.
 A node strand $s$ corresponds to one node orientation.
@@
 
-We build a graph $\mathcal{V}$ inductively.
+We build a graph $\mathcal{G}$ inductively.
 We take the first match in $A$, $m_1$, and execute a union-find operation to obtain $m_1^+$.
-We add the character of the match $c(m_1^+)$ as a node $n_1$ in $\mathcal{V}$, and record the mapping from $m_1^+ \to n_1$.
-To induce the graph, we take the next unused match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $\mathcal{V}$.
-To allow the annotation of paths, we record the set of characters in $Q$ that match to a given node in $\mathcal{V}$ in mapping $Z = Q \to \mathcal{N} = m_1 \ldots m_{|\mathcal{N}|}$.
+We add the character of the match $c(m_1^+)$ as a node $v_1$ in $\mathcal{G}$, and record the mapping from $m_1^+ \to v_1$.
+To induce the graph, we take the next unused match in $A : m_i \notin \forall_{j < i} m_j^+$, obtain $m_i^+$, and add $c(m_i^+)$ to $\mathcal{G}$.
+To allow the annotation of paths, we record the set of characters in $Q$ that match to a given node in $\mathcal{G}$ in mapping $Z = Q \to \mathcal{V} = m_1 \ldots m_{|\mathcal{V}|}$.
 We continue until all matches have been used.
-Finally, we establish paths ($\mathcal{P}$) by walking them in $\mathcal{V}$ using $Z$, and record edges ($\mathcal{E}$) where nodes occur successively in paths.
+Finally, we establish paths ($\mathcal{P}$) by walking them in $\mathcal{G}$ using $Z$, and record edges ($\mathcal{E}$) where nodes occur successively in paths.
 
@@
 
-The set of alignments represented by a variation graph is strictly larger than the set of alignments used to induce it.
+The set of alignments represented by a variation graph may be strictly larger than the set of alignments used to induce it.
 The graph must contain at least the set of alignments given in input.
 It may also contain new implied pairwise relationships that arise due to transitive match relationships, as shown in Figure \ref{fig:induction} for closures 1 and 6.
-However, by definition of $\mathcal{V}$, it cannot contain \textit{less} match information than represented in the set of matches ($A$).
+However, by definition of $\mathcal{G}$, it cannot contain \textit{less} match information than represented in the set of matches ($A$).
 
@@
 We do so non-redundantly by marking each used character in $Q$ in an auxiliary data structure $\mathcal{X}$, which could be encoded as a bitvector of length $|Q|$.
-As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{N}$ and the sequence-to-graph mapping $Z$, which, like $A$, consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{N}$.
+As we compute the transitive closures, we emit both the nodes (single characters) of the graph $\mathcal{V}$ and the sequence-to-graph mapping $Z$, which, like $A$, consists of match pairs, but rather than mapping $Q \to Q$, maps $Q \to \mathcal{V}$.
 As with $A$, we can sort $Z$ to obtain random access via binary search.
@@
 Assume an all-to-all alignment of $N$ sequences in $A$ as input, and that all sequences are approximately identical, so that the induced variation graph has $\mathcal{Q}/N$ nodes.
-The induction must maintain reference to all characters in all input sequences $O(\mathcal{Q})$, all character-to-character matches $O(|A|) \approx O(\mathcal{Q}N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(\mathcal{Q})$, the nodes of the graph $O(|\mathcal{N}|) \approx O(\mathcal{Q}/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(\mathcal{Q})$, and the set of paths $O(|\mathcal{P}|) \approx O(\mathcal{Q})$.
+The induction must maintain reference to all characters in all input sequences $O(\mathcal{Q})$, all character-to-character matches $O(|A|) \approx O(\mathcal{Q}N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(\mathcal{Q})$, the nodes of the graph $O(|\mathcal{V}|) \approx O(\mathcal{Q}/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(\mathcal{Q})$, and the set of paths $O(|\mathcal{P}|) \approx O(\mathcal{Q})$.
 We also maintain the bitvector $\mathcal{X}$ to mark seen characters of $Q$ during graph induction, which requires $O(\mathcal{Q})$ bits, equivalent to $O(\mathcal{Q} / \log_2\mathcal{Q})$ integer identifiers.
@@
 
-Considering the time complexity of induction, we anticipate $O(|A|\log|A|)$ time to sort the match set and $O(\mathcal{N} \log |A|)$ to query it and compute our $\mathcal{N}$ transitive closures.
-Computing the variation graph paths $\mathcal{P}$ involves converting sequences in $S$ to walks through $\mathcal{N}$.
+Considering the time complexity of induction, we anticipate $O(|A|\log|A|)$ time to sort the match set and $O(\mathcal{V} \log |A|)$ to query it and compute our $\mathcal{V}$ transitive closures.
+Computing the variation graph paths $\mathcal{P}$ involves converting sequences in $S$ to walks through $\mathcal{V}$.
 We first sort the sequence-to-graph mapping array $Z$ in $O(\mathcal{Q} \log \mathcal{Q})$ operations, and then compute $\mathcal{P}$ in $O(\mathcal{Q})$ queries which each cost $O(\log \mathcal{Q})$.
 To obtain unique edges and generate $\mathcal{E}$, we must build and sort an array of size $O(2|P|)$, and then iterate through it for $O(2|P| \log 2|P| + 2|P|)$ operations.
-In sum, we would expect to require $O(|A|\log|A| + \mathcal{N} \log |A| + 2(\mathcal{Q} \log \mathcal{Q}) + 2(|P| \log 2|P| + |P|))$.
+In sum, we would expect to require $O(|A|\log|A| + \mathcal{V} \log |A| + 2(\mathcal{Q} \log \mathcal{Q}) + 2(|P| \log 2|P| + |P|))$.
 Using our approximate relationships to $\mathcal{Q}$ given previously and simplifying, we arrive at $O[\mathcal{Q}(2N^2 \log N / \log \mathcal{Q} + (1/N + 4)\log \mathcal{Q} + 2 + \log(4))]$.
@@
 The same encoding can be used to replace the sequence-to-graph mapping $Z$, yielding $\mathcal{Z}$.
-Rather than pairs of characters in $Q$ and $\mathcal{N}$, we record runs of matches between them as range matches.
-Although in expectation the length of these matches should be strictly less than $k$, due to the interruption of the graph by variation between genomes, this still allows us to reduce the size of $Z$ using runs of matches between $Q$ and $\mathcal{N}$.
-Additionally, we store the inverse of $\mathcal{Z}$, which maps ranges from $\mathcal{N} \to Q$, as $\bar{\mathcal{Z}}$.
-We use $\mathcal{Z}$ to compact non-branching regions of $\mathcal{V}$ into single nodes, and $\bar{\mathcal{Z}}$ to accelerate our calculation of links in the graph.
+Rather than pairs of characters in $Q$ and $\mathcal{V}$, we record runs of matches between them as range matches.
+Although in expectation the length of these matches should be strictly less than $k$, due to the interruption of the graph by variation between genomes, this still allows us to reduce the size of $Z$ using runs of matches between $Q$ and $\mathcal{V}$.
+Additionally, we store the inverse of $\mathcal{Z}$, which maps ranges from $\mathcal{V} \to Q$, as $\bar{\mathcal{Z}}$.
+We use $\mathcal{Z}$ to compact non-branching regions of $\mathcal{G}$ into single nodes, and $\bar{\mathcal{Z}}$ to accelerate our calculation of links in the graph.
 
@@
 Rather than recording an identifier for each character in a sequence graph, it is useful to compact characters that form trivial linear components in the graph into single nodes.
-Broadly, the size of nodes will be bounded by the average distance between variants, which, for pangenomes built from \textasciitilde100 individuals of the same species, often provides a great reduction in the total number of nodes (and thus identifiers) required for $\mathcal{V}$ and its components.
+Broadly, the size of nodes will be bounded by the average distance between variants, which, for pangenomes built from \textasciitilde100 individuals of the same species, often provides a great reduction in the total number of nodes (and thus identifiers) required for $\mathcal{G}$ and its components.
 
-To compact $\mathcal{V}$, we traverse $Q$, finding each entry in $\mathcal{Z}$ in turn, recording its start and end in $\mathcal{N}$, which can be understood as a character vector or string containing all the sequence in the nodes of $\mathcal{V}$.
-%in a ``graph vector'' $\mathcal{G} = n_1,n_2 \ldots n_{|\mathcal{N}|}$ containing the ordered characters of the graph.
+To compact $\mathcal{G}$, we traverse $Q$, finding each entry in $\mathcal{Z}$ in turn, recording its start and end in $\mathcal{V}$, which can be understood as a character vector or string containing all the sequence in the nodes of $\mathcal{G}$.
+%in a ``graph vector'' $\mathcal{G} = v_1,v_2 \ldots v_{|\mathcal{V}|}$ containing the ordered characters of the graph.
 %Any position in $\mathcal{G}$ so marked must be the start or end of a node.
-We subsequently use these markings to subdivide $\mathcal{N}$ into a compacted version $\mathcal{N}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{N}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
+We subsequently use these markings to subdivide $\mathcal{V}$ into a compacted version $\mathcal{V}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{V}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
 $\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.
-%such that $\forall_{i \in 1 \ldots |\mathcal{N}'|} |n_i| \ge 1$.
+%such that $\forall_{i \in 1 \ldots |\mathcal{V}'|} |v_i| \ge 1$.
 %In Algorithm \ref{alg:seqwish} we provide a detailed description of the graph induction algorithm.
@@
 This partitioning can introduce boundary effects which change the contents of $\mathcal{Z}$ and $\bar{\mathcal{Z}}$ by splitting ranges at the boundaries of our partitions.
-However, while this will affect the compressed node definition $\mathcal{N}'$, it does not affect $\mathcal{N}$, and it can be corrected via a post-processing step to sort and compact the id space.
+However, while this will affect the compressed node definition $\mathcal{V}'$, it does not affect $\mathcal{V}$, and it can be corrected via a post-processing step to sort and compact the id space.
 %Partitioning reduces the working memory limits to be in the order of the size of the largest transitive match closure found in the input.
@@
 
-For output in GFA, we iterate over nodes in $\mathcal{N}'$, writing each as a node record.
+For output in GFA, we iterate over nodes in $\mathcal{V}'$, writing each as a node record.
 Edges are similarly produced from the disk-backed multiset representing $\mathcal{E}$.
 The most computationally expensive part of graph emission is the rendering of the input sequences $S$ as paths $\mathcal{P}$ through the graph.
-For each input sequence in the \textit{seqidx}, we walk through the offsets in $S$ contained in the sequence and look up their mapping into $\mathcal{N}'$ using $\mathcal{Z}$.
+For each input sequence in the \textit{seqidx}, we walk through the offsets in $S$ contained in the sequence and look up their mapping into $\mathcal{V}'$ using $\mathcal{Z}$.
 Range compression allows us to complete one lookup per range.
@@
 This allows us work with $\mathcal{A}$, $\mathcal{Z}$, and $\bar{\mathcal{Z}}$ in external memory.
-By storing pairs of numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$ in terms of offsets in $\mathcal{N}$.
-The graph sequence vector $\mathcal{N}$ is simply written by appending characters to a file.
-We mark nodes to generate $\mathcal{N}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \citep{Gog_2014} for support of the final emission of the graph $\mathcal{V}$.
+By storing pairs of numerical identifiers in the backing array, we are able to generate a disk-backed multiset model which we use to compute the unique set of edges $\mathcal{E}$ in terms of offsets in $\mathcal{V}$.
+The graph sequence vector $\mathcal{V}$ is simply written by appending characters to a file.
+We mark nodes to generate $\mathcal{V}'$ using a bitvector kept in main memory, over which we subsequently generate a rank/select dictionary \citep{Gog_2014} for support of the final emission of the graph $\mathcal{G}$.
 

### 2022-08-08 manuscript/fig_from_alignment_to_variation_graph.tex
l/m swap in the pseudocode and tweaks
Figure1: replace `text[1,2,3]` with `seq[1,2,3]`

@@
 %  \centering
-   \includegraphics[width=\linewidth]{fig_induction.pdf}
+   \includegraphics[width=\linewidth]{fig_induction}
    \caption{
@@
    The unique set of node pairings in the paths provide the edges of the output graph.
-   Closures 1 and 6 imply pairwise relationships between sequences text1 and text3, and text1 and text2, respectively, that are both absent in the input alignments.
+   Closures 1 and 6 imply pairwise relationships between sequences seq1 and seq3, and seq1 and seq2, respectively, that are both absent in the input alignments.
    %   Black nodes are numbered by the rank of the transitive closure operation
    %   At bottom, a sequence graph obtained by contracting 
-   %   At top, three sequences (text1, text2, text3) 
+   %   At top, three sequences (seq1, seq2, seq3)
     }

### 2022-08-08 manuscript/main.tex
replace `\mathcal{Q}` with ``|S|`

@@
 However, a na\"{i}ve algorithm based on this model would require a very large amount of space.
-Take $\mathcal{Q} = |S|$.
-Although our identifier space $\mathcal{Q}$ must include all of $Q$, in practice, we only store $S$, as $\bar{S}$ can be trivially computed.
-Assume an all-to-all alignment of $N$ sequences in $A$ as input, and that all sequences are approximately identical, so that the induced variation graph has $\mathcal{Q}/N$ nodes.
-The induction must maintain reference to all characters in all input sequences $O(\mathcal{Q})$, all character-to-character matches $O(|A|) \approx O(\mathcal{Q}N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(\mathcal{Q})$, the nodes of the graph $O(|\mathcal{V}|) \approx O(\mathcal{Q}/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(\mathcal{Q})$, and the set of paths $O(|\mathcal{P}|) \approx O(\mathcal{Q})$.
-We also maintain the bitvector $\mathcal{X}$ to mark seen characters of $Q$ during graph induction, which requires $O(\mathcal{Q})$ bits, equivalent to $O(\mathcal{Q} / \log_2\mathcal{Q})$ integer identifiers.
-In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(\mathcal{Q}(N^2 + 1/N + 1/log_2{Q} + 4))$ space.
+Although our identifier space $|S|$ must include all of $Q$, in practice, we only store $S$, as $\bar{S}$ can be trivially computed.
+Assume an all-to-all alignment of $N$ sequences in $A$ as input, and that all sequences are approximately identical, so that the induced variation graph has $|S|/N$ nodes.
+The induction must maintain reference to all characters in all input sequences $O(|S|)$, all character-to-character matches $O(|A|) \approx O(|S| N^2)$, the mapping of $Q$ into the graph $O(|Z|) \approx O(|S|)$, the nodes of the graph $O(|\mathcal{V}|) \approx O(|S|/N)$, the size of the edge set $O(|\mathcal{E}|) \approx O(|S|)$, and the set of paths $O(|\mathcal{P}|) \approx O(|S|)$.
+We also maintain the bitvector $\mathcal{X}$ to mark seen characters of $Q$ during graph induction, which requires $O(|S|)$ bits, equivalent to $O(|S| / \log_2 |S|)$ integer identifiers.
+In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(|S|(N^2 + 1/N + 1/log_2{Q} + 4))$ space.
 
-Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $\mathcal{Q} \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
+Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $|S| \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
 Such a design is almost infeasible for inputs larger than a handful of genomes.
@@
 Computing the variation graph paths $\mathcal{P}$ involves converting sequences in $S$ to walks through $\mathcal{V}$.
-We first sort the sequence-to-graph mapping array $Z$ in $O(\mathcal{Q} \log \mathcal{Q})$ operations, and then compute $\mathcal{P}$ in $O(\mathcal{Q})$ queries which each cost $O(\log \mathcal{Q})$.
+We first sort the sequence-to-graph mapping array $Z$ in $O(|S| \log |S|)$ operations, and then compute $\mathcal{P}$ in $O(|S|)$ queries which each cost $O(\log |S|)$.
 To obtain unique edges and generate $\mathcal{E}$, we must build and sort an array of size $O(2|P|)$, and then iterate through it for $O(2|P| \log 2|P| + 2|P|)$ operations.
-In sum, we would expect to require $O(|A|\log|A| + \mathcal{V} \log |A| + 2(\mathcal{Q} \log \mathcal{Q}) + 2(|P| \log 2|P| + |P|))$.
-Using our approximate relationships to $\mathcal{Q}$ given previously and simplifying, we arrive at $O[\mathcal{Q}(2N^2 \log N / \log \mathcal{Q} + (1/N + 4)\log \mathcal{Q} + 2 + \log(4))]$.
+In sum, we would expect to require $O(|A|\log|A| + \mathcal{V} \log |A| + 2(|S| \log |S|) + 2(|P| \log 2|P| + |P|))$.
+Using our approximate relationships to $|S|$ given previously and simplifying, we arrive at $O[|S|(2N^2 \log N / \log |S| + (1/N + 4)\log |S| + 2 + \log(4))]$.
 
-Due to our dependence on sorting, and the logarithmic-time cost of queries, growth in $\mathcal{Q}$ drives $\Omega(\mathcal{Q} \log \mathcal{Q})$ growth in overall complexity.
+Due to our dependence on sorting, and the logarithmic-time cost of queries, growth in $|S|$ drives $\Omega(|S| \log |S|)$ growth in overall complexity.
 As $N$ grows, both time and space complexity are dominated by the number of alignments, which in the case of our example is $O(N^2)$.
@@
 
-%$O[(\mathcal{Q}/N + 5\mathcal{Q}) \log \mathcal{Q} + \mathcal{Q}(2 + log(4))]$.
-%$O(\mathcal{Q} \log \mathcal{Q} + \mathcal{Q}/N \log \mathcal{Q} + 2(\mathcal{Q} \log \mathcal{Q}) + 2(\mathcal{Q} \log 2\mathcal{Q} + \mathcal{Q}))$.
+%$O[(|S|/N + 5|S|) \log |S| + |S|(2 + log(4))]$.
+%$O(|S| \log |S| + |S|/N \log |S| + 2(|S| \log |S|) + 2(|S| \log 2|S| + |S|))$.
 

### 2022-09-08 manuscript/main.tex
add simplified asymptotic notation
stuff

@@
 We also maintain the bitvector $\mathcal{X}$ to mark seen characters of $Q$ during graph induction, which requires $O(|S|)$ bits, equivalent to $O(|S| / \log_2 |S|)$ integer identifiers.
-In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(|S|(N^2 + 1/N + 1/log_2{Q} + 4))$ space.
+In total, na\"{i}ve theorem \ref{thm:seqwish}-based induction would require approximately $O(|S|(N^2 + 1/N + 1/log_2{Q} + 4))$ space, or simply $O(|S| N^2)$.
 
@@
 %Any position in $\mathcal{G}$ so marked must be the start or end of a node.
-We subsequently use these markings to subdivide $\mathcal{V}$ into a compacted version $\mathcal{V}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{V}|$ where the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
+We subsequently use these markings to subdivide $\mathcal{V}$ into a compacted version $\mathcal{V}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{V}|$ such that the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
 $\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.

### 2022-09-13 manuscript/main.tex
order bias workflow

@@
 An unbiased pangenome graph would represent the alignment of all included genomes to all others.
-Existing methods approximate this relationship by progressive alignment to a graph initially based on a reference genome \citep{Li:2020}, through a global structuring of the genome relationships in a neighbor joining phylogenetic tree \citep{Armstrong:2020}, or via creation of a \textit{de Bruijn} graph based on a fixed $k$-mer length \citep{Minkin_2016,Yu_2021}.
+Existing methods approximate this relationship by progressive alignment to a graph initially based on a reference genome \citep{Li:2020}, through a global structuring of the genome relationships in a neighbor joining phylogenetic tree \citep{Armstrong:2020}, or via creation of a \textit{de Bruijn} graph based on a fixed $k$-mer length \citep{Minkin_2016,Sheikhizadeh_2016,Yu_2021}.
 These methods limit computational costs by reducing the number of pairwise comparisons, but in turn their results depend on input genome order, selected reference, guide-tree topology, or $k$-mer length.

### 2022-09-14 manuscript/main.tex
yay

@@
 
+\begin{table*}[!ht]
+    \centering
+    \caption{Performance of the graph induction algorithm.
+        For each pangenome we report a single experiment with \textit{seqwish -k} filter set to 49bp.
+        From left to right, the columns indicate the species, the number of sequences (that is, the number of contigs), number of haplotypes (that is, number of individuals), the sum of the length of all sequences in Gbp, the length of the short match filter applied in bp, the time in seconds and the amount of memory and disk space in Gbytes required for the graph induction, the length of the resulting graph in Gbp and the number its connected components.
+        }
+    \label{tab:experiments}
+    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
+        \hline
+        species & sequences & haplotypes & fasta.Gbp & min.match.bp & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbp & components \\ \hline
+        \textit{A. thaliana} & 922 & 16 & 1.90251 & 49 & 468 & 43.1287 & 7.1218 & 0.234284 & 100 \\ \hline
+        \textit{H. sapiens} & 17478 & 38 & 114.627 & 49 & 46268 & 347.4983 & 604.4261 & 4.47126 & 474 \\ \hline
+        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 777 & 74.9484 & 20.2070 & 0.01421 & 5 \\ \hline
+        \textit{Z. mays} & 46289 & 41 & 90.2491 & 49 & 31043 & 351.1235 & 402.8716 & 13.8838 & 925 \\ \hline
+    \end{tabular}
+\end{table*}
+
 \section{Results}
@@
 
-\begin{table*}[!ht]
+\input{fig_experiments.tex}
+
+\subsection{Comparison to similar graph building methods}
+\label{sec:comparison}
+
+We have so far described our method for graph induction, provided detail on its implementation, and demonstrated its behavior over a set of pangenome graph building problems.
+These provide intuition about \textit{seqwish} and its behavior.
+However, the reader may wonder how the method compares to other similar methods for graph construction, or what variation might be caused by changes in the alignment method used.
+
+To focus on this questions, we examine the case of a single chromosome (chrV) of \textit{S. cerevisiae} for which we have 7 assemblies \citep{Yue_2017}.
+We compute 100 random permutations \citep{Williams2009-ou} of these assemblies and provide them to several graph construction methods, including \textit{minigraph} \citep{Li:2020}, \textit{TwoPaCo} \citep{Minkin_2016}, and \textit{seqwish} based on \textit{minimap2} \citep{Li_2018} or \textit{wfmash} \citep{wfmash} alignments.
+
+Of existing methods, \textit{TwoPaCo} is the only that provides a variation graph in the same form as \textit{seqwish}.
+However, it is based on a De Bruijn Graph that must have a relatively low $k$-mer length, leading to complex topologies that in our experience cannot be resolved easily by increasing $k$.
+We do not expect \textit{TwoPaCo} to be biased with respect to input genome order.
+In contrast, \textit{minigraph} develops a kind of partial order alignment over the minimizers (sparse $k$-mer set) of input genomes, and it progressively builds the graph by adding new variation from each genome in order.
+We do expect it be order and reference (first input) biased, although the degree of bias requires experimentation to establish.
+Finally, \textit{seqwish} itself is by definition unbiased with respect to input order, but if the alignments are biased due to changes in order, we should see effects on the resulting graph.
+
+We use ODGI to compute the size of the resulting graphs \citep{Guarracino2022}, and use this as a proxy for graph identity.
+When the graph length changes, it provides a clear indication that the graph itself has changed.
+It also gives an indication of the scale of the differences, which may indicate how large an effect of a different construction approach might be.
+Conversely, if the graph length is the same for two large graphs, we assume that in most cases they really do have the same topology and sequence space, even if they may have different node identifiers and ordering in the serialization of the graph.
+
+\input{fig_chrV_yeast.tex}
+
+\begin{table}[!h]
     \centering
-    \caption{Performance of the graph induction algorithm.
-        For each pangenome we report a single experiment with \textit{seqwish -k} filter set to 49bp.
-        From left to right, the columns indicate the species, the number of sequences (that is, the number of contigs), number of haplotypes (that is, number of individuals), the sum of the length of all sequences in Gbp, the length of the short match filter applied in bp, the time in seconds and the amount of memory and disk space in Gbytes required for the graph induction, the length of the resulting graph in Gbp and the number its connected components.
+    \caption{Variance in graph properties across genome input order permutations.
+    We report the mean ($\mu$) and variance ($\sigma^2$) in standard deviations ($\sigma$) for graph length (bp), node count (node), and edge count (edge).
+    We show results from \textit{minigraph}, \textit{TwoPacCo} with $k=19$ (\textit{twopaco.k19}), \textit{seqwish} using \textit{minimap2} alignments (\textit{seqwish.mm2}), and \textit{seqwish} using \textit{wfmash} alignments (\textit{seqwish.wfm}).
         }
     \label{tab:experiments}
-    \begin{tabular}{|l|l|l|l|l|l|l|l|l|l|}
+    \begin{tabular}{|l|l|l|l|l|l|l|}
         \hline
-        species & sequences & haplotypes & fasta.Gbp & min.match.bp & time.seconds & memory.Gbytes & disk.Gbytes & graph.Gbp & components \\ \hline
-        \textit{A. thaliana} & 922 & 16 & 1.90251 & 49 & 468 & 43.1287 & 7.1218 & 0.234284 & 100 \\ \hline
-        \textit{H. sapiens} & 17478 & 38 & 114.627 & 49 & 46268 & 347.4983 & 604.4261 & 4.47126 & 474 \\ \hline
-        \textit{H. pylori} & 292 & 250 & 0.407782 & 49 & 777 & 74.9484 & 20.2070 & 0.01421 & 5 \\ \hline
-        \textit{Z. mays} & 46289 & 41 & 90.2491 & 49 & 31043 & 351.1235 & 402.8716 & 13.8838 & 925 \\ \hline
+        method & $\mu$.bp & $\sigma$.bp & $\mu$.node & $\sigma$.node & $\mu$.edge & $\sigma$.edge \\ \hline
+        \textit{minigraph} &  648967 & 8112 & 190 & 11.6 & 267 & 15.9 \\ \hline
+        \textit{seqwish.mm2}  & 609039  &  4.08 & 30692 &  2.89 &  41723  &  6.32 \\ \hline
+        \textit{seqwish.wfm}  & 683470 & 0 & 28756 & 0 & 39080 & 0 \\ \hline
+        \textit{twopaco.k19} & 1270115 & 0 & 26383 & 0 & 35284 & 0 \\ \hline
     \end{tabular}
-\end{table*}
+\end{table}
 
@@
 %\FIXME{to do}
-
-\input{fig_experiments.tex}
 

### 2022-09-14 manuscript/main.tex
onward

@@
 The version of \textit{wfmash} used in these experiments allows us to align sequences with up to 10\% divergence between them, providing highly-sensitive input for our experiments.
+Our use of \textit{wfmash} is pragmatic, and any mapping method capable of generating PAF with base-level alignments in CIGAR strings is capable of being used as input to \textit{seqwish}.
 
@@
 
+Our results with these four species-specific pangenomes give a basic demonstration of how pangenome graphs can be adjusted through key parameters of alignment and match filtering.
+These results are demonstrative of graph induction.
+In typical applications, we expect \textit{seqwish} to be applied as part of a larger process to build and refine a pangenome graph \citep{Garrison2022_pggb,Liao_2022}.
+
 \input{fig_experiments.tex}
 
-\subsection{Comparison to similar graph building methods}
+\subsection{Comparison to related graph building methods}
 \label{sec:comparison}
@@
 
-Of existing methods, \textit{TwoPaCo} is the only that provides a variation graph in the same form as \textit{seqwish}.
+Of existing methods, \textit{TwoPaCo} provides a graph in a form that is similar to \textit{seqwish}'s.
 However, it is based on a De Bruijn Graph that must have a relatively low $k$-mer length, leading to complex topologies that in our experience cannot be resolved easily by increasing $k$.
-We do not expect \textit{TwoPaCo} to be biased with respect to input genome order.
+Furthermore, the overlaps between nodes are incompatible with the variation graph model, and must be reduced or ``bluntified'' with additional processing steps to be usable as input to variation graph tools \citep{Eizenga2021-yg}.
+Because it is based on a De Bruijn Graph, which is symmetrically constructed from a set of $k$-mers and not the sequences themselves, we do not expect \textit{TwoPaCo} to be biased with respect to input genome order.
+
 In contrast, \textit{minigraph} develops a kind of partial order alignment over the minimizers (sparse $k$-mer set) of input genomes, and it progressively builds the graph by adding new variation from each genome in order.
 We do expect it be order and reference (first input) biased, although the degree of bias requires experimentation to establish.
-Finally, \textit{seqwish} itself is by definition unbiased with respect to input order, but if the alignments are biased due to changes in order, we should see effects on the resulting graph.
 
-We use ODGI to compute the size of the resulting graphs \citep{Guarracino2022}, and use this as a proxy for graph identity.
+Finally, \textit{seqwish} itself is by definition unbiased with respect to input order, but if the alignments are affected by changes in order, the resulting graph will also change.
+
+We use ODGI to compute the size of the resulting graphs \citep{Guarracino2022}.
 When the graph length changes, it provides a clear indication that the graph itself has changed.
@@
 Conversely, if the graph length is the same for two large graphs, we assume that in most cases they really do have the same topology and sequence space, even if they may have different node identifiers and ordering in the serialization of the graph.
+
+As we observe in Figure \ref{fig:yeast}, \textit{minigraph} indeed generates a distribution of graph lengths with clusters that apparently correspond to similar regions of the permutation space.
+However, this is not the case for the \textit{seqwish} builds based on \textit{minimap2} and \textit{wfmash}, which appear to have almost exactly the same length across all permutations.
+We further consider variance in the length, node count, and edge count of all graphs.
+This shows that, as expected, \textit{minigraph}'s output depends markedly on input genome order.
+We also find that \textit{minimap2} configuration we used is not order-independent, and can produce slightly different outputs when the reference and query sequences are ordered differently, demonstrating that \textit{seqwish} is only ``unbiased'' insofar as the input mappings are.
+For \textit{TwoPaCo} and \textit{wfmash}+\textit{seqwish}, we observe no variation in length.
+The length of \textit{TwoPaCo} graphs is significantly larger due to the repetition in node sequences, which overlap by $k-1$ base pairs.
+%, when considering the variance of the length, node count, and edge count of the graphs
 

### 2022-09-14 manuscript/main.tex
almost done

@@
 
-As we observe in Figure \ref{fig:yeast}, \textit{minigraph} indeed generates a distribution of graph lengths with clusters that apparently correspond to similar regions of the permutation space.
+As we observe in Figure \ref{fig:yeast}, \textit{minigraph} indeed generates a distribution of graph lengths with clusters corresponding to permutations beginning with the same genome.
+This suggests that most of the variation in \textit{minigraph}'s graph derives from the first genome which is picked.
+The graphs are thus biased towards the chosen base reference.
 However, this is not the case for the \textit{seqwish} builds based on \textit{minimap2} and \textit{wfmash}, which appear to have almost exactly the same length across all permutations.
+
 We further consider variance in the length, node count, and edge count of all graphs.
@@
 The length of \textit{TwoPaCo} graphs is significantly larger due to the repetition in node sequences, which overlap by $k-1$ base pairs.
-%, when considering the variance of the length, node count, and edge count of the graphs
+
+Although this particular study focuses on only a single small eukaryotic chromosome, it reveals that our basic understanding of reference bias in thes graph construction methods is correct.
+By design, progressive pangenome graph construction methods will be affected by input genome order.
+But, other symmetric methods avoid order bias.
+%This leads to difficulty in building and designing pangenomes that is avoided by unbiased methods.
 

### 2022-09-14 manuscript/main.tex
yupp

@@
 
-In contrast, \textit{minigraph} develops a kind of partial order alignment over the minimizers (sparse $k$-mer set) of input genomes, and it progressively builds the graph by adding new variation from each genome in order.
+In contrast, \textit{minigraph} develops a kind of partial order alignment \citep{Lee_2002} over the minimizers (sparse $k$-mer set) of input genomes, and it progressively builds the graph by adding new variation from each genome in order.
 We do expect it be order and reference (first input) biased, although the degree of bias requires experimentation to establish.
 
-Finally, \textit{seqwish} itself is by definition unbiased with respect to input order, but if the alignments are affected by changes in order, the resulting graph will also change.
+Finally, \textit{seqwish} itself is by definition unbiased with respect to input order.
+But, if the alignments are affected by changes in order, the resulting graph will also change.
 
-We use ODGI to compute the size of the resulting graphs \citep{Guarracino2022}.
-When the graph length changes, it provides a clear indication that the graph itself has changed.
+We use ODGI to compute the size, node count, and edge count of the resulting graphs \citep{Guarracino2022}.
+Changes in these statistics indicate changes in graphs.
 It also gives an indication of the scale of the differences, which may indicate how large an effect of a different construction approach might be.
-Conversely, if the graph length is the same for two large graphs, we assume that in most cases they really do have the same topology and sequence space, even if they may have different node identifiers and ordering in the serialization of the graph.
+If graph parameters are the same for two large graphs, we consider them equivalent, which is not formally correct but provides a sufficient approximation for this analysis.
 
@@
 
-We further consider variance in the length, node count, and edge count of all graphs.
+In Table \ref{tab:yeast}, we further consider variance in the length, node count, and edge count of all graphs.
 This shows that, as expected, \textit{minigraph}'s output depends markedly on input genome order.
-We also find that \textit{minimap2} configuration we used is not order-independent, and can produce slightly different outputs when the reference and query sequences are ordered differently, demonstrating that \textit{seqwish} is only ``unbiased'' insofar as the input mappings are.
+We also find that \textit{minimap2} configuration we used is not order-independent, and can produce slightly different outputs when the reference and query sequences are ordered differently, demonstrating that \textit{seqwish} is only ``unbiased'' insofar as its input mappings are.
 For \textit{TwoPaCo} and \textit{wfmash}+\textit{seqwish}, we observe no variation in length.
@@
     \caption{Variance in graph properties across genome input order permutations.
-    We report the mean ($\mu$) and variance ($\sigma^2$) in standard deviations ($\sigma$) for graph length (bp), node count (node), and edge count (edge).
+    We report the mean ($\mu$) and variance in standard deviations ($\sigma$) for graph length (bp), node count (node), and edge count (edge).
     We show results from \textit{minigraph}, \textit{TwoPacCo} with $k=19$ (\textit{twopaco.k19}), \textit{seqwish} using \textit{minimap2} alignments (\textit{seqwish.mm2}), and \textit{seqwish} using \textit{wfmash} alignments (\textit{seqwish.wfm}).
         }
-    \label{tab:experiments}
+    \label{tab:yeast}
     \begin{tabular}{|l|l|l|l|l|l|l|}
@@
 In contrast to existing methods, which depend on particular structuring of their input \citep{Li:2020,Armstrong:2020}, \textit{seqwish} is unbiased in that as it directly and uniformly represents sequence relationships given on input in the resulting graph.
-Although we did not compare with \textit{de Bruijn} graph methods, which are also unbiased in their uniform treatment of input genomes \citep{Minkin_2016,Yu_2021}, we believe such methods are fundamentally different in that because they collapse all exact matches of a given length $k$.
+Our comparison with related pangenome graph-building methods demonstrates that these effects can be significant.
+\textit{minigraph}'s output varies substantially, with $\sigma \approx 1.25\%$ of graph length for a 7-chromosome input with $\approx 99.3\%$ pairwise identity.
+%Although we did not compare with \textit{de Bruijn} graph methods, which are also unbiased in their uniform treatment of input genomes \citep{Minkin_2016,Yu_2021}, we believe such methods are fundamentally different in that because they collapse all exact matches of a given length $k$.
 This prevents high-level structuring of patterns of homology and orthology in the resulting graph, and they furthermore cannot be guided by a specific alignment set.

### 2022-09-14 manuscript/main.tex
final tweaks

@@
 \textit{minigraph}'s output varies substantially, with $\sigma \approx 1.25\%$ of graph length for a 7-chromosome input with $\approx 99.3\%$ pairwise identity.
-%Although we did not compare with \textit{de Bruijn} graph methods, which are also unbiased in their uniform treatment of input genomes \citep{Minkin_2016,Yu_2021}, we believe such methods are fundamentally different in that because they collapse all exact matches of a given length $k$.
-This prevents high-level structuring of patterns of homology and orthology in the resulting graph, and they furthermore cannot be guided by a specific alignment set.
+De Bruijn Graph methods, which are also unbiased with their treatment of input genomes \citep{Minkin_2016,Sheikhizadeh_2016,Yu_2021}.
+%However, our approach is different than these, and allows the direct
 
 Our presentation is necessarily limited, in order to focus on and describe the unique problem of variation graph induction.
-Thus, in this manuscript and our experiments, we have not explored the full problem of \textit{pangenome graph building}, which include both the initial alignment step and downstream processing of the resulting graph.
+Thus, in this manuscript and our experiments, we have not explored the full problem of \textit{pangenome graph building}, which include both the initial alignment step and downstream processing of the resulting graph \citep{Liao_2022}.
 These topics lie outside of the scope of the presented work, wherein we have focused on a key kernel which is a bottleneck in the pangenome graph construction process.
@@
 And this lossless property does not guarantee that the resulting graph is easy to work with or navigate; in practice, downstream processing is usually required to normalize the graph for many applications.
-We will cover these topics in future work.
+We expect to cover these topics in future work.
 

### 2022-09-14 manuscript/main.tex
supercazzola

@@
 
-We have so far described our method for graph induction, provided detail on its implementation, and demonstrated its behavior over a set of pangenome graph building problems.
+We have so far described our method for graph induction, provided detail on its implementation, and demonstrated its performance over a set of pangenome graph building problems.
 These provide intuition about \textit{seqwish} and its behavior.
@@
 
-To focus on this questions, we examine the case of a single chromosome (chrV) of \textit{S. cerevisiae} for which we have 7 assemblies \citep{Yue_2017}.
+To focus on these questions, we examine the case of a single chromosome (chrV) of \textit{S. cerevisiae} for which we have 7 assemblies \citep{Yue_2017}.
 We compute 100 random permutations \citep{Williams2009-ou} of these assemblies and provide them to several graph construction methods, including \textit{minigraph} \citep{Li:2020}, \textit{TwoPaCo} \citep{Minkin_2016}, and \textit{seqwish} based on \textit{minimap2} \citep{Li_2018} or \textit{wfmash} \citep{wfmash} alignments.
@@
 Of existing methods, \textit{TwoPaCo} provides a graph in a form that is similar to \textit{seqwish}'s.
-However, it is based on a De Bruijn Graph that must have a relatively low $k$-mer length, leading to complex topologies that in our experience cannot be resolved easily by increasing $k$.
+However, it is based on a de Bruijn graph that must have a relatively low $k$-mer length, leading to complex topologies that in our experience cannot be resolved easily by increasing $k$.
 Furthermore, the overlaps between nodes are incompatible with the variation graph model, and must be reduced or ``bluntified'' with additional processing steps to be usable as input to variation graph tools \citep{Eizenga2021-yg}.
-Because it is based on a De Bruijn Graph, which is symmetrically constructed from a set of $k$-mers and not the sequences themselves, we do not expect \textit{TwoPaCo} to be biased with respect to input genome order.
+Because it is based on a de Bruijn graph, which is symmetrically constructed from a set of $k$-mers and not the sequences themselves, we do not expect \textit{TwoPaCo} to be biased with respect to input genome order.
 
@@
 \textit{minigraph}'s output varies substantially, with $\sigma \approx 1.25\%$ of graph length for a 7-chromosome input with $\approx 99.3\%$ pairwise identity.
-De Bruijn Graph methods, which are also unbiased with their treatment of input genomes \citep{Minkin_2016,Sheikhizadeh_2016,Yu_2021}.
+De Bruijn graph methods, which are also unbiased with their treatment of input genomes \citep{Minkin_2016,Sheikhizadeh_2016,Yu_2021}.
 %However, our approach is different than these, and allows the direct
@@
 And this lossless property does not guarantee that the resulting graph is easy to work with or navigate; in practice, downstream processing is usually required to normalize the graph for many applications.
-We expect to cover these topics in future work.
+We expect to cover these topics in future work, some of which are ongoing \citep{Liao_2022,Garrison2022_pggb}.
+
+Despite its apparently high costs, the symmetric pangenome construction modality that \textit{seqwish} presents provides key advantages over progressive approaches.
+Although a complete all-vs-all alignment of a set of genomes is costly, it is trivial to run in parallel.
+This feature is unavailable to progressive construction methods, which necessarily involve a serial introduction of new information into the graph.
+Although the total costs of a symmetric alignment and pangenome graph construction modality are high, they are in some sense more manageable.
+The modular, independent nature of the hardest part of the computation---the derivation of the alignments---can be scaled out to large compute clusters as well as cached for incremental construction and update of large pangenomes.
+
+We furthermore expect that local giant components will tend to arise in the alignment graph without requiring a complete pairwise alignment, suggesting that lessons from graph theory may help to guide a justified sparsification, or downsampling, of the input \citep{Janson1993-ac}.
+This suggests that simple random sampling of alignments may evoke minimal changes in graph structure, provided that there are many homologous copies of each locus in the pangenome.
+It will also be possible to apply \textit{seqwish} to measure at what level of mapping sparsification we observe changes in the induced graph.
+Although this may be infeasible for very large problems, it will allow us to develop an empirical understanding of how to reduce the complexity of the initial alignment step without affecting the resulting pangenome graph.
 

### 2022-09-14 manuscript/main.tex
ok

@@
 
-Although this particular study focuses on only a single small eukaryotic chromosome, it reveals that our basic understanding of reference bias in thes graph construction methods is correct.
+Although this particular study focuses on only a single small eukaryotic chromosome, it reveals that our basic understanding of reference bias in these graph construction methods is correct.
 By design, progressive pangenome graph construction methods will be affected by input genome order.

### 2022-10-08 manuscript/main.tex
tweak the redundant sentence

@@
 The graph must contain at least the set of alignments given in input.
-It may also contain new implied pairwise relationships that arise due to transitive match relationships, as shown in Figure \ref{fig:induction} for closures 1 and 6.
-However, by definition of $\mathcal{G}$, it cannot contain \textit{less} match information than represented in the set of matches ($A$).
+In other words, by definition of $\mathcal{G}$, it cannot contain \textit{less} match information than represented in the set of matches ($A$).
+However, the graph may also contain new implied pairwise relationships that arise due to transitive match relationships, as shown in Figure \ref{fig:induction} for closures 1 and 6.
 

### 2022-10-08 manuscript/main.tex
do not replace the 'approximately' word with '\approx'

@@
 
-Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $|S| \approx \num{e11}$, we might expect to use $\approx \num{e15}$ identifiers to store the full model.
+Assuming that we want to build a graph of 100 haploid human genomes of $\num{3e9}$ bp, where $N=100$ and $|S| \approx \num{e11}$, we might expect to use approximately $\num{e15}$ identifiers to store the full model.
 Such a design is almost infeasible for inputs larger than a handful of genomes.
-For instance, we would need $\approx \num{3e12}$ identifiers for just 5 human genomes.
+For instance, we would need approximately $\num{3e12}$ identifiers for just 5 human genomes.
 Although it is feasible to compute such a graph using external memory, the approximately 200-fold increase in space relative to the input renders this clearly impractical.
@@
 If the average pairwise diversity of sequences in our input is $1/k$, we expect exact matches to be around $k$ characters long.
-By encoding matches as pairs of ranges of characters, $r = (a, b) : a, b = (i, j) : i, j \in Q$, we can obtain a $\approx$ $k$-fold compression of $A$, yielding the range-match array $\mathcal{A}$.
+By encoding matches as pairs of ranges of characters, $r = (a, b) : a, b = (i, j) : i, j \in Q$, we can obtain an approximately $k$-fold compression of $A$, yielding the range-match array $\mathcal{A}$.
 
@@
 
-%We can save $\approx k$ match pairs per recorded range
+%We can save approximately $k$ match pairs per recorded range
 %a pair of ranges $r = [(a_0, a_1), (b_0, b_1)]$ in $Q$ requires twice the space, but we expect it to
@@
 Our comparison with related pangenome graph-building methods demonstrates that these effects can be significant.
-\textit{minigraph}'s output varies substantially, with $\sigma \approx 1.25\%$ of graph length for a 7-chromosome input with $\approx 99.3\%$ pairwise identity.
+\textit{minigraph}'s output varies substantially, with $\sigma \approx 1.25\%$ of graph length for a 7-chromosome input with approximately $99.3\%$ pairwise identity.
 De Bruijn graph methods, which are also unbiased with their treatment of input genomes \citep{Minkin_2016,Sheikhizadeh_2016,Yu_2021}.

### 2022-10-08 manuscript/main.tex
English tweaks

@@
 %Any position in $\mathcal{G}$ so marked must be the start or end of a node.
-We subsequently use these markings to subdivide $\mathcal{V}$ into a compacted version $\mathcal{V}'$ where compacted node boundaries are marked in a auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{V}|$ such that the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
+We subsequently use these markings to subdivide $\mathcal{V}$ into a compacted version $\mathcal{V}'$ where compacted node boundaries are marked in an auxiliary bitvector $\mathcal{B} : |\mathcal{B}| = |\mathcal{V}|$ such that the first character in each compacted node is marked by a $1$ and other characters are marked $0$.
 $\mathcal{B}$ allows us to compute compacted node ids using efficient rank operations \citep{Gog_2014}.
@@
 In contrast, \textit{minigraph} develops a kind of partial order alignment \citep{Lee_2002} over the minimizers (sparse $k$-mer set) of input genomes, and it progressively builds the graph by adding new variation from each genome in order.
-We do expect it be order and reference (first input) biased, although the degree of bias requires experimentation to establish.
+We do expect it to be order and reference (first input) biased, although the degree of bias requires experimentation to establish.
 

### 2022-10-08 manuscript/main.tex
tweak minimap+seqwish and be more explicit about being a seqwish configuration

@@
 This shows that, as expected, \textit{minigraph}'s output depends markedly on input genome order.
-We also find that \textit{minimap2} configuration we used is not order-independent, and can produce slightly different outputs when the reference and query sequences are ordered differently, demonstrating that \textit{seqwish} is only ``unbiased'' insofar as its input mappings are.
+We also find that the \textit{minimap2}+\textit{seqwish} configuration is not order-independent because \textit{minimap2} can produce slightly different outputs when the reference and query sequences are ordered differently.
+This demonstrates that \textit{seqwish} is only ``unbiased'' insofar as its input mappings are.
 For \textit{TwoPaCo} and \textit{wfmash}+\textit{seqwish}, we observe no variation in length.

### 2022-10-08 manuscript/main.tex
fix incomplete sentence

@@
 \textit{minigraph}'s output varies substantially, with $\sigma \approx 1.25\%$ of graph length for a 7-chromosome input with approximately $99.3\%$ pairwise identity.
-De Bruijn graph methods, which are also unbiased with their treatment of input genomes \citep{Minkin_2016,Sheikhizadeh_2016,Yu_2021}.
+Although we did not compare with de Bruijn graph methods, which are also unbiased with their treatment of input genomes \citep{Minkin_2016,Sheikhizadeh_2016,Yu_2021}, we believe such methods are fundamentally different in that because they collapse all exact matches of a given length k.
+This prevents high-level structuring of patterns of homology and orthology in the resulting graph, and they furthermore cannot be guided by a specific alignment set.
 %However, our approach is different than these, and allows the direct

### 2022-10-08 manuscript/main.tex
fix the references

@@
 \label{sec:introduction}
-A pangenome models the full genomic information of a species or clade \citep{Medini_2005,Sherman_2020}.
+A pangenome models the full genomic information of a species or clade \citep{Medini_2005,Sherman_2020}.Garrison_2018_thesis
 In contrast to reference-based approaches that relate sequences to a particular reference genome, methods that use pangenome reference systems attempt to model the mutual relationship between all represented genomes \citep{cpang2018}.
@@
 \label{def:vg}
-Variation graphs are a common formalism to encode pangenome graphs \citep{Garrison_2019_thesis}.
+Variation graphs are a common formalism to encode pangenome graphs \citep{Garrison_2018_thesis}.
 In the variation graph $\mathcal{G} = (\mathcal{V}, \mathcal{E}, \mathcal{P})$, nodes (or vertices) $\mathcal{V} = v_1\ldots v_{|\mathcal{V}|}$ contain sequences of DNA.

### 2022-11-04 manuscript/fig_chrV_yeast.tex
get it going

\begin{figure}[b]
%  \centering
   \includegraphics[width=\linewidth]{yeast_chrV_length_vs_100orders}
   \caption{
     Base-pair lengths of graphs built from 100 permutations of yeast chrV.
     Each point corresponds to the length of a graph built from a specific permutation and the given method.
     We color each point by the first genome in the order (the ``reference'' in \textit{minigraph}).
     A boxplot per first genome group provides the mean and the first and third quartile intervals for each reference grouping.
    }
    \label{fig:yeast}
\end{figure}

### 2022-11-14 manuscript/main.tex
get it going
remove links from the references

--- 2022-10-08 manuscript/main.tex
@@
 \label{sec:introduction}
-A pangenome models the full genomic information of a species or clade \citep{Medini_2005,Sherman_2020}.
+A pangenome models the full genomic information of a species or clade \citep{Medini_2005,Sherman_2020,Garrison_2019_thesis}.
 In contrast to reference-based approaches that relate sequences to a particular reference genome, methods that use pangenome reference systems attempt to model the mutual relationship between all represented genomes \citep{cpang2018}.
@@
 Our implementation reads standard data formats, FASTA or FASTQ for the input sequences, and PAF \citep{Li_2018} for pairwise alignments.
-It writes the graph in standard Graphical Fragment Assembly (GFA) format \citep{GFA}.
+It writes the graph in standard Graphical Fragment Assembly (GFA) format (\url{https://github.com/GFA-spec/GFA-spec}).
 
@@
 In our implementation, we rely on several basic external memory kernels.
-To reduce working memory requirements to an absolute minimum, we use a disk-backed version of the implicit interval tree that memory-maps the sorted array of intervals \citep{mmmulti}.
+To reduce working memory requirements to an absolute minimum, we use a disk-backed version of the implicit interval tree that memory-maps the sorted array of intervals (\url{https://github.com/ekg/mmmulti}).
 Indexing the implicit interval tree requires a sorting step which dominates the runtime of our algorithm.
@@
 When set very high, this filter can be used to generate a coarse, high-confidence graph built only from very long exact matches which will tend to be unique in the genome.
-Although the application of the $k>$ filter can result in a graph that is relatively ``under-aligned'', we can further refine it through the application of local multiple sequence alignment \citep{Gao_2020}, or graph normalization \citep{Doerr_2021}.
+Although the application of the $k>$ filter can result in a graph that is relatively ``under-aligned'', we can further refine it through the application of local multiple sequence alignment \citep{Gao_2020}, or graph normalization (\url{https://github.com/marschall-lab/GFAffix}).
 In a pangenomic context, underalignment caused by $k>$ match filtering can be mitigated by transitive relationships present in the pangenome.
@@
 %    After the transitive closure, we write our final graph on disk in a set of arrays, then augment it with the paths of our input sequences,
-%    finally emitting the resulting graph in the standard Graphical Fragment Assembly (GFA) format~\citep{GFA}.
+%    finally emitting the resulting graph in the standard Graphical Fragment Assembly (GFA) format (\url{https://github.com/GFA-spec/GFA-spec}).
 
@@
 
-To construct the graph we first generate alignments with \textit{wfmash} \citep{wfmash}, a DNA sequence aligner designed specifically for high performance all-to-all alignment of fully-assembled genomes.
+To construct the graph we first generate alignments with \textit{wfmash} (\url{https://github.com/waveygang/wfmash}), a DNA sequence aligner designed specifically for high performance all-to-all alignment of fully-assembled genomes.
 \textit{wfmash} combines an algorithm for generating whole-genome homology maps (\textit{MashMap2}) \citep{Jain_2018} with an extension of the wavefront algorithm (WFA) \citep{Marco_Sola_2020} capable of obtaining base-level alignments for whole chromosomes.
@@
 To focus on these questions, we examine the case of a single chromosome (chrV) of \textit{S. cerevisiae} for which we have 7 assemblies \citep{Yue_2017}.
-We compute 100 random permutations \citep{Williams2009-ou} of these assemblies and provide them to several graph construction methods, including \textit{minigraph} \citep{Li:2020}, \textit{TwoPaCo} \citep{Minkin_2016}, and \textit{seqwish} based on \textit{minimap2} \citep{Li_2018} or \textit{wfmash} \citep{wfmash} alignments.
+We compute 100 random permutations \citep{Williams2009-ou} of these assemblies and provide them to several graph construction methods, including \textit{minigraph} \citep{Li:2020}, \textit{TwoPaCo} \citep{Minkin_2016}, and \textit{seqwish} based on \textit{minimap2} \citep{Li_2018} or \textit{wfmash} (\url{https://github.com/waveygang/wfmash}) alignments.
 
